\documentclass[11pt]{article}

% Geometry
%
\usepackage[papersize={12.85cm,19.84cm},%
            includehead,%
            includefoot,%
            bindingoffset=1cm%, showframe, showcrop
            ]{geometry}

%\setlength{\evensidemargin}{-50pt}
%\setlength{\oddsidemargin}{-50pt}
\setlength{\topmargin}{-50pt}
\setlength{\textheight}{500pt}
%\setlength{\textwidth}{530pt}

\usepackage[T1]{fontenc}
\usepackage[main=english,hungarian]{babel}
\AddToHook{begindocument}[test]{\catcode\string``=12 }
\DeclareHookRule{begindocument}{test}{before}{babel}
\usepackage{ebgaramond}
%\usepackage{emptypage}
\usepackage{microtype}

%\usepackage{verse}
%\poemlines{0} % No verse numbering
\usepackage[final]{pdfpages}
\usepackage{hyphenat}
%\usepackage{pgfornament}

\frenchspacing  % Follow French conventions after a period
\hfuzz 2pt      % Do not report overhang less than 2pt

% Title and author
%
\title{Companion English-Hungarian Glossary\\
to \emph{The Tiger and the Roe Deer}}
\author{Christian Rinderknecht\\}
\date{\oldstylenums{2026}}

\begin{document}

\maketitle

\noindent \textbf{\large A} \\
\textbf{abban} \emph{az} `that' + iness. \emph{-ban} `in' \\
\textbf{ablakok} \emph{ablak} `window' + pl. \emph{-ok} \\
\textbf{adott} vb. \emph{ad} `to give' + 3rd pers. sg. ind. past. indef. \\
\textbf{ahhoz} \emph{az} `that' + allat. \emph{-hoz} `towards' \\
\textbf{ahogy} \emph{az} `that' + \emph{hogy} $\rightarrow$ `as, as
soon as' \\
\textbf{ahol} art. \emph{az} + interr. adv. \emph{hol} `where?' \\
\textbf{ajkad} arch. \emph{ajk} `lip', syn. \emph{ajak} + 2nd
pers. sg. single\hyp{}poss. \emph{-d} \\
\textbf{ajkaimon} arch. \emph{ajk} `lip', syn. \emph{ajak} +  1st
pers. sg. poss. mult.\hyp{}poss. \emph{-aim} + super. \emph{-on} `on' \\
\textbf{akár} lit. `just like' \\
\textbf{akárhogyan} lit. \emph{akár} `just like' + \emph{hogyan} `how'
$\rightarrow$ `anyway, anyhow, whatever', alt. \emph{akárhogy} \\
\textbf{akarja} vb. `to want' + 3rd
pers. sg. ind. pres. indef. \emph{-ja} \\
\textbf{akarod} vb. \emph{akar} `to want' + 2nd
pers. sg. ind. pres. def. \emph{-od} \\
\textbf{aki} conj. `who' \\
\textbf{akik} pron. \emph{aki} `who' + pl. \emph{-k} \\
\textbf{akiket} pron. \emph{aki} `who' + pl. \emph{-k} +
acc. \emph{-et} \\
\textbf{akinek} pron. \emph{aki} `who' + attr. poss. \emph{-nek} \\
\textbf{akiknek} pron. \emph{aki} `who' + pl. \emph{-k} +
attr. poss. \emph{-nek} \\
\textbf{akkor} \emph{az} `that' + \emph{-kor} `time, at the time of'
$\rightarrow$ `then, at that time' \\
\textbf{alatt} post. \emph{al-} `lower part, below' +
locat. \emph{-att} \\
\textbf{alattam} post. \emph{al-} `lower part, below' +
locat. \emph{-att} + poss. \emph{-a} + 1st pers. sg. \emph{-m}
$\rightarrow$ `under me' \\
\textbf{alhatnék} vb. \emph{alszik}
`to sleep' $\rightarrow$ `to fall asleep' + pot. 1st
pers. sg. cond. indef. \emph{-hatnék} \\
\textbf{alhatom} vb. inf. \emph{alszik} `to sleep' + 1st pers. sg. pot. \\
\textbf{alig} adv. `barely' \\
\textbf{alkonyat} \emph{alkony} `dusk, twilight, nightfall' +
noun\hyp{}forming \emph{-at} \\
\textbf{almák} \emph{alma} `apple' + pl. \emph{-k} \\
\textbf{alól} \emph{al} `lower part' + obs. ablat. \emph{-l} `from' \\
\textbf{alusznak} vb. \emph{alszik} `to sleep' $\rightarrow$ `to fall
asleep' + 3rd pers. pl. ind. pres. indef. \emph{-nak} \\
\textbf{ami} rel. pron. `which, that' \\
\textbf{amíg} adv. `as long as'; `until' if followed by negated
verb \\
\textbf{amint} \emph{az} + \emph{mint} `like': `as soon as',
lit. `as' \\
\textbf{amióta} art. \emph{az} + pron. \emph{mi} `which, that'
+ \emph{óta} `since (a point in time)' \\
\textbf{amire} \emph{ami} `that, which' + subl. \emph{-re} `onto' \\
\textbf{amott} adv. `there, over there, far away' \\
\textbf{ampát} \emph{ampa} `bucket' + acc. \emph{-t} \\
\textbf{annak} art. \emph{az} `that' + attr. poss. \emph{-nak}
  $\rightarrow$ \emph{annak} \\
\textbf{annyi} pron. `so much, so many' \\
\textbf{annyira} pron. \emph{annyi} + subl. \emph{-ra}
`on[to]' $\rightarrow$ adv. `so much' \\
\textbf{anyám} \emph{anya} `mother' + 1st
pers. sg. poss. single\hyp{}poss. \emph{-m} \\
\textbf{anyja} \emph{anya} `mother' + 3rd
pers. sg. poss. single\hyp{}poss. \emph{-ja} \\
\textbf{anyókára} \emph{anyó} `elderly woman' + dim. \emph{-ka} (term
of endearment; rare. term of address to the wife by her husband in old
age) + subl. \emph{-ra} `on[to]' \\
\textbf{apja} \emph{apa} `father' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-ja} \\
\textbf{arany} `gold, golden' \\
\textbf{aranykeretben} \emph{arany} `gold' + \emph{keret} `frame' +
iness. \emph{-ben} `in' \\
\textbf{asztalomat} \emph{asztal} `table' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-om} + acc. \emph{-at} \\
\textbf{asszony} `lady' \\
\textbf{autó} `car' \\
\textbf{azért} art. \emph{az} `that' + causal \emph{-ért}
$\rightarrow$ adv. `for that reason, that is why' \\
\textbf{azon} art. \emph{az} `that' + super. \emph{-on}
$\rightarrow$ `on (that manner)' \\
\textbf{azóta} art. \emph{az} `that' + \emph{óta} `since'
$\rightarrow$ `since then' \\
\textbf{azt} \emph{az} `that' + acc. \emph{-t} \\
%%
\ \\
\noindent \textbf{\large Á} \\
\textbf{ág} `branch' + acc. \emph{-at} \\
\textbf{ágra} \emph{ág} `branch' + subl. \emph{-ra}
`onto' \\
\textbf{ágról} \emph{ág} `branch' + delat. \emph{-ról} `off, from' \\
\textbf{ágy} `bed' \\
\textbf{ágyban} `bed' + iness. \emph{-ban} `in' \\
\textbf{ágyúk} \emph{ágyú} `cannon' + pl. \emph{-k} \\
\textbf{áhit} vb. `to wish (for smth)' \\
\textbf{áldalak} vb. \emph{áld} `to bless, to praise' + 1st
pers. sg. ind. pres. 2nd obj. \emph{-alak} \\
\textbf{áll} vb. `to stand' \\
\textbf{állt} vb. `to stand' + 3rd pers. sg. ind. past indef. \\
\textbf{állatnak} vb. \emph{áll} `to stand' +
noun\hyp{}forming \emph{-at} $\rightarrow$ \emph{állat} `animal' +
dat. \emph{-nak} \\
\textbf{álljatok} vb. \emph{áll} `to stand' + 2nd pers. pl. informal
subj. \\
\textbf{állok} vb. \emph{áll} `to stand' + 1st
pers. sg. ind. pres. indef. \emph{-ok} \\
\textbf{álma} \emph{álom} `dream' + 3rd
pers. sg. single\hyp{}poss. `his' \\
\textbf{álmaiban} \emph{álom} `dream' + 2nd
pers. sg. inform. poss. \emph{-a} + poss. pl. \emph{-i} +
iness. \emph{-ban} `in' $\rightarrow$ \emph{álmaiban} \\
\textbf{álmaim} \emph{álom} `dream' + 1st pers. sg. poss. \emph{-a} +
poss. pl. \emph{-i} + 1st pers. sg. pers. \emph{-m} \\
\textbf{álmaimban} \emph{álom} `dream' + 1st pers. sg. poss. \emph{-a}
+ poss. pl. \emph{-i} + 1st pers. sg. pers. \emph{-m} +
  iness. \emph{-ban} `in' $\rightarrow$ \emph{álmaimban} \\
\textbf{álmot} \emph{álom} `dream' + acc. \emph{-ot} \\
\textbf{álmukban} \emph{álom} `dream'+ 3rd
pers. pl. single\hyp{}poss. poss. \emph{-uk} + iness. \emph{-ban} `in' \\
\textbf{ám} poetic, interj. contrad. `but' \\
\textbf{ár} `price, value', `flood, tide, flow (as opposed to ebb)' \\
\textbf{áradás} vb. \emph{árad} `to flood' + noun\hyp{}forming \emph{-ás} \\
\textbf{áramló} vb. \emph{áramlik} `to flow, stream' + pres. part. \emph{-ó} \\
\textbf{árnyak} \emph{árny} `shadow' + nom. pl. \emph{-ak} \\
\textbf{árnyat} \emph{árny} `shadow' + acc. \emph{-t} \\
\textbf{árnyán} \emph{árny} `shadow' + adess. `in' \\
\textbf{árnyát} \emph{árny} `shadow' + 3rd
  pers. sg. single\hyp{}poss. poss. \emph{-a} + acc. \emph{-t} $\rightarrow$ \emph{-át} \\
\textbf{árnyukon} \emph{árny} `shadow' + 3rd
pers. pl. single\hyp{}poss. poss. \emph{-uk} + super. \emph{-on}
`on' \\
\textbf{árultam} \emph{áru} `good, merchandise' +
vb.\hyp{}forming \emph{-l} $\rightarrow$ `to sell, peddle, deal in' +
1st pers. sg. ind. past \emph{-tam} \\
\textbf{át} adv. or prep. rare `across, over, through' \\
\textbf{átjár} vb. pref. \emph{át-} `through, across' + vb. \emph{jár}
`to walk, to pass by', `to dance', `to visit, go about (repetitive
actions)' etc. + 3rd pers. sg. ind. pres. indef. \\
%%
\ \\
\noindent \textbf{\large B} \\
\textbf{babrált} vb. \emph{babrál} `to fiddle' \\
\textbf{baltája} \emph{balta} `axe' + 3rd
pers. sg. sing. single\hyp{}poss. \emph{-ja} \\
\textbf{bambát} \emph{bamba} `dim\hyp{}witted, fool' +
acc. \emph{-at} \\
\textbf{barátom} \emph{barát} `friend' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-om} \\
\textbf{barom} arch. `head of cattle'  \\
\textbf{bámultam} vb. \emph{bámul} `to admire, to gaze upon' + 1st
pers. sg. ind. past indef. \\
\textbf{bánat} vb. \emph{bán} `to be bothered' +
noun\hyp{}forming \emph{-at} $\rightarrow$ `sorrow, grief' \\
\textbf{bánni} vb. \emph{megbán} `to regret':
perfect. vb. pref. \emph{meg-} + vb. \emph{bán} `to mind (negatively),
to be bothered' \\
\textbf{bánt} vb. \emph{bán} `to be bothered' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{bántasz} vb. \emph{bán} `to be bothered' + caus. \emph{-t}
$\rightarrow$ `to hurt' + 2nd pers. sg. ind. pres. indef. \\
\textbf{bántották} vb. \emph{bán} `to mind (negatively),
to be bothered' + caus. \emph{-t} $\rightarrow$ \emph{bánt} `to hurt',
inf. \emph{bántani} + 3rd pers. pl. ind. past \emph{-ták} \\
\textbf{bántottalak} vb. \emph{bán} `to mind (negatively),
to be bothered' + caus. \emph{-t} $\rightarrow$ \emph{bánt} `to hurt',
inf. \emph{bántani} + 1st pers. sg. ind. pres. 2nd obj. \\
\textbf{bármikor} \emph{bár} `any' + adv. \emph{mikor} `when'
  $\rightarrow$ `any time' \\
\textbf{bársonyon} \emph{bársony} `velvet' + super. \emph{-on}
`on (that manner)' \\
\textbf{bátyám} \emph{bátya} dated `uncle' + 1st
pers. sg. poss. single\hyp{}poss. \emph{-am} \\
\textbf{be} adv. `in[wardly]' (anton. \emph{ki}) \\
\textbf{behallatszik} vb. pref. \emph{be-} `inwardly' +
vb. \emph{hall} `to hear' + caus. \emph{-at} +
vb.-forming \emph{-szik} $\rightarrow$ `to cause to sound in' \\
\textbf{behallik} syn. \emph{behallatszik} \\
\textbf{bekopog} vb. pref. \emph{be-} `inwardly' + vb. \emph{kopog}
`to knock' \\
\textbf{belehalt} vb. pref. \emph{bele-} `into' + vb. \emph{hal} `to
die' $\rightarrow$ `to die from (smth, smb)' + 3rd pers. sg. ind. past
indef. \\
\textbf{belénk} pron. \emph{belé} (folksy, syn. \emph{bele}) `into him' + 1st
pers. pl. $\rightarrow$ `with us, into us' \\
\textbf{belőled} pron. \emph{belőle} `out of him' + 2nd
pers. sg. \emph{-d} \\
\textbf{belül} `inside' \\
\textbf{benn} adv. `inside'. Syn. \emph{bent} \\
\textbf{bennem} pron. \emph{benne} `inside smth, smb' + 1st
pers. sg. pers. \emph{-m} \\
\textbf{bent} adv. `inside' \\
\textbf{benyúl} vb. pref. \emph{be-} `inwardly' + vb. `to reach in
smth to take smth' \\
\textbf{bevallhatom} vb. pref. \emph{be-} `inwardly' + vb. \emph{vall} `to confess, admit, declare, show,
speak of' $\rightarrow$ `to admit' + 1st
pers. sg. pot. ind. pres. def. \\
\textbf{békés} \emph{béke} `peace' + adj.\hyp{}forming \emph{-s}
$\rightarrow$ `peaceful' \\
\textbf{békességet} \emph{béke} `peace' + abst. noun\hyp{}forming
  \emph{-ség} + acc. \emph{-et} \\
\textbf{bérház} \emph{bér} `rental' + \emph{ház} `house' $\rightarrow$
`tenement' \\
\textbf{bicegve} vb. \emph{biceg} `to limp, to wobble, to halt' +
adv. part. \emph{-ve} \\
\textbf{bilincseket} \emph{bilincs} `handcuff, shackle' +
pl. \emph{-ek} + acc. \emph{-et} \\
\textbf{biz} adv. `indeed' \\
\textbf{bizonyosabbat} adj. \emph{bizonyos} `certain' +
comp. \emph{-abb} `more' + acc. \emph{-t} \\
\textbf{bíztam} vb. \emph{bíz} `to entrust smth (to smb)' \\
\textbf{bocsánatot} \emph{bocsánat} `pardon' + acc. \emph{-ot} \\
\textbf{bogai} \emph{bog} `knot' + 3rd
pers. sg. poss. mult.\hyp{}poss. \emph{-ai} \\
\textbf{bogarak} \emph{bogár} `beetle' + pl. \emph{-ak}
  $\rightarrow$ \emph{bogarak} \\
\textbf{bogáncsokat} \emph{bogáncs} `thistle' + pl. \emph{-ok} +
acc. \emph{-at} \\
\textbf{bokra} \emph{bokor} `shrub, bush' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-a} \\
\textbf{bokron} \emph{bokor} `shrub, bush' + super. \emph{-on} `on' \\
\textbf{boldogságot} adj. \emph{boldog} `happy' + noun\hyp{}forming
  \emph{-ság} (state of being) + acc. \emph{-ot} \\
\textbf{bolond} adj. `silly, foolish' \\
\textbf{bolyong} vb. `to wander' + 3rd pers. sg. ind. pres. indef. \\
\textbf{borítja} vb. \emph{borít} `to cover' + 3rd
pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{bólintott} vb. \emph{bólint} `to nod' + 3rd
pers. sg. ind. past indef. \\
\textbf{bólintva}  vb. \emph{bólint} `to nod' + adv. part. \emph{-va} \\
\textbf{buckákra} \emph{bucka} `dune' + subl. sg. \emph{-ra} `onto, to' \\
\textbf{bujtam} vb. \emph{bújik} `to hide' + 1st pers. sg. ind. past
indef. \\
\textbf{burkolózom} vb. \emph{burkol} `to wrap, cover' +
freq.\hyp{}reflex. \emph{-ózik} $\rightarrow$ inf. \emph{burkolózik}
`wrap oneself' + 1st pers. sg. ind. pres. indef. \emph{-ózom} \\
\textbf{búrok} \emph{búr} `Boer' + pl. \emph{-ok} \\
\textbf{búsuljatok} \emph{bú} `sorrow' + adj.\hyp{}forming \emph{-s}
+ vb.\hyp{}forming \emph{-ul} $\rightarrow$ vb. \emph{búsul} `to be
sad' (obs. `to be angry') + 2nd pers. pl. ind. pres. \emph{-jatok} \\
\textbf{búval} lit. \emph{bú} `sorrow, grief' + instr. \emph{-val}
`with' \\
\textbf{búzaföldekkel} \emph{búza} `wheat' + \emph{föld} `land, field'
+ pl. \emph{-ek} + instr. \emph{-kel} \\
\textbf{büszke} adj. `proud' \\
%%
\ \\
\noindent \textbf{\large C} \\
\textbf{cella} `cell' \\
\textbf{cementfalak} \emph{cement} `cement'+ \emph{fal} `wall' +
pl. \emph{-ak} \\
\textbf{csak} adv. `only, merely, just' \\
\textbf{csap} vb. `to strike, beat' + 3rd pers. sg. ind. pres. indef. \\
\textbf{csalogány} poet. `nightingale', syn. \emph{fülemüle} \\
\textbf{csend} `silence'. alt. \emph{csönd} \\
\textbf{csendhez} `silence' + allat. \emph{-hez} `to' \\
\textbf{csiklándani} lit. \emph{csiklánd}, syn. \emph{csikland} +
vb. ending `to tickle (playful courtship)' \\
\textbf{csikorgó} vb. \emph{csikorog} `to creak, to squeak' + pres. part. \emph{-ó} \\
\textbf{csikó} `foal' \\
\textbf{csillag} `star' \\
\textbf{csillagfellegek} \emph{csillag} `star' + lit. \emph{felleg}
`cloud' + pl. \emph{-ek} \\
\textbf{csillagok} \emph{csillag} `star' + pl. \emph{-ok} \\
\textbf{csillan} vb. `to sparkle, glitter' + 3rd pers. sg. ind. pres. \\
\textbf{csillognak} vb. \emph{csillog} `to glitter' + 3rd
pers. pl. ind. pres. indef. \\
\textbf{csilló} vb. \emph{csillog} `to glitter' +
 adj.\hyp{}forming \emph{-o} $\rightarrow$ \emph{csilló};
 usually \emph{csillogó} \\
 \textbf{csinál} vb. `to do, to make' + 3rd
 pers. sg. ind. pres. indef. \\
\textbf{csobogó} vb. \emph{csobog} `to splash, to babble (water
 flowing)' + pres. part. \emph{-ó} \\
\textbf{csodáltam} \emph{csoda} `miracle' + vb\hyp{}forming \emph{-l}
$\rightarrow$ `to admire' + 1st pers. sg. ind. past \\
\textbf{csontos} \emph{csont} `bone' + adj.\hyp{}forming \emph{-os} \\
\textbf{csókolnom} vb. \emph{csókol} `to kiss' + 1st
pers. sg. ind. pres. def. \\
\textbf{csönd} `silence', alt. \emph{csend} \\
\textbf{csöndes} adj. sg. `silent', alt. \emph{csendes} \\
\textbf{csendesen} adj. \emph{csendes} `silent, quiet' +
super. \emph{-en} `on' $\rightarrow$ adv. `silently, quietly' \\
\textbf{csörömpölés} vb. \emph{csörömpöl} `to clank' +
  vb. noun\hyp{}forming \emph{-és} $\rightarrow$ `clanking' \\
\textbf{csősz} `field-keeper, guard (fig.)' \\
\textbf{csupa} `all, mere, pure' \\
\textbf{csupán} \emph{csupa} `all, mere, pure' +
deadj. adv.\hyp{}forming suffix \emph{-n} $\rightarrow$ `merely, only' \\
%%
\ \\
\noindent \textbf{\large D} \\
\textbf{dalol} vb. `to sing' + 3rd pers. sg. ind. pres. indef. \\
\textbf{dalt} \emph{dal} `song' + acc. \emph{-t} \\
\textbf{darab} `piece (self\hyp{}contained, not as a fraction of
  smth, like in the theatre)' \\
\textbf{darabokra} adv. \emph{darab} `piece, chunk (self\hyp{}contained)' +
  pl. \emph{-ok} + subl. \emph{-ra} `on[to]' \\
\textbf{de} adv. `how!' \textbar\textbar{} conj. `but, surely'
\textbf{deszkát} \emph{deszk} `board' + acc. \emph{-át} \\
\textbf{determinált} Latin \emph{determinare} + vb.\hyp{}forming \emph{-ál}
`to determine' + past part. \emph{-t} \\
\textbf{dél} `noon, south (sg.)' \\
\textbf{dicső} lit. `heroic, glorious' \\
\textbf{dinnyét} \emph{dinnye} `melon' + acc. \emph{-t} \\
\textbf{divatos} \emph{divat} `fashion' +
adj.\hyp{}forming \emph{-os} \\
\textbf{dobálják} vb. \emph{dob} `to throw' + repetitive \emph{-ál} + 3rd
pers. pl. ind. pres. def. \emph{-ják} $\rightarrow$ `to throw
repeatedly' \\
\textbf{dolog} `thing' \\
\textbf{dombon} \emph{domb} `hill, mound' + super. \emph{-on} `on' \\
\textbf{dorombol} vb. `to purr (cat)', lit. `to rumble' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{döngenek} vb. \emph{döng} `to bang, batter' + arch. 3rd
pers. pl. ind. pres. indef. \\
\textbf{döntsd} vb. \emph{dönt} `to decide about or on
smth', \textbar\textbar{} `to fell
(a tree), to topple' + 2nd pers. sg. informal subj. indef. \\
\textbf{döntöd} perfect. vb. pref. \emph{el-} + vb. \emph{dönt} `to
decide' + 2nd pers. sg. ind. pres. def. \\
\textbf{dörgölődzik} vb. `to press against', 3rd
pers. sg. ind. pres. indef. \\
\textbf{dörmög} vb. `to rumble (bear), to grumble (person)' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{drága} adj. `expensive' \\
\textbf{düdedék} lit. `ruin, rubble, wasteland' \\
\textbf{dúdolom} vb. \emph{dúdol} `to hum' + 1st
pers. sg. ind. pres. def. \emph{-om} \\
%%
\ \\
\noindent \textbf{\large E} \\
\textbf{e} poetic abbrv. \emph{eme, ez} `this' \\
\textbf{ebbe} \emph{ez} `this' + illative \emph{-be}
`into (the inside of)' $\rightarrow$ `in this' \\
\textbf{ebed} noun. \emph{eb} `dog' + 2nd
pers. sg. single\hyp{}poss. poss. $\rightarrow$ `your dog' \\
\textbf{egek} \emph{ég} `sky' + pl. \emph{-ek} \\
\textbf{egekre} \emph{ég} `sky' + pl. \emph{-ek} + subl. \emph{-re} `onto' \\
\textbf{egeret} \emph{egér} `mouse' + acc. \emph{-et}
$\rightarrow$ \emph{egeret} \\
\textbf{eget} \emph{ég} `sky' + acc. \emph{-et} $\rightarrow$
  \emph{eget} \\
\textbf{egész} adj. `complete, whole, healthy' \textbar\textbar{}
adv. `wholly' \\
\textbf{egészet} \emph{egész} `whole, healthy, complete' +
acc. \emph{-et} \\
\textbf{egy} `one' \\
\textbf{egyben-másban} pron. \emph{egymás} + iness. \emph{-ben/-ban}
`in' $\rightarrow$ `in some ways' \\
\textbf{egyedül} \emph{egy} `one' + noun\hyp{}forming \emph{-ed} +
adv.\hyp{}forming \emph{-ül} $\rightarrow$ `alone' \\
\textbf{egyengettem} vb. \emph{egyenget} `to straighten' + 1st
pers. sg. ind. past \\
\textbf{egyet} \emph{egy} + acc. \emph{-t} \\
\textbf{egymagam} pron. `alone' + 1st pers. sg. \\
\textbf{egymás} pron. `each other, each one another' \\
\textbf{egymásról} pron. \emph{egymá} `each other, each one another' +
delat. \emph{-ról} `off, from' \\
\textbf{egymáson} pron. \emph{egymás} `each other, each one
another' + super. \emph{-on} `on' \\
\textbf{egyik} \emph{egy} `one' + 3rd pers. pl. mult. poss. \emph{-ik}
`of' \\
\textbf{egyszerre} adv. \emph{egyszer} `at once' +
  abl. \emph{-re} `onto' \\
\textbf{egyszerű} \emph{egy} `one' + \emph{-szerű} `-like'
$\rightarrow$ `plain, simple, unadorned, humble (< 1945)' \\
\textbf{együtt} \emph{egy} `one' + locat. \emph{-ütt} $\rightarrow$
`in the company of, together, with' \\
\textbf{ejh} interj. `eh' \\
\textbf{el} adv., conj. `afar, far away' \textbar\textbar{} \emph{el
nem} + past. part. \textbar\textbar{} perf. vb. pref. \\
\textbf{eladott} vb. \emph{elad} `to sell (complete transaction)' + past part. \emph{-ott} \\
\textbf{elakadt} vb. \emph{elakad} `to falter, to get stuck, to stop'
+ 3rd pers. sg. ind. past indef. \\
\textbf{elalszik} perfect. vb. pref. \emph{el-} + vb. \emph{alszik}
`to sleep' $\rightarrow$ `to fall asleep' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{elalszol} vb. pref. \emph{el-} `away' + vb. \emph{alszik} `to
sleep' $\rightarrow$ `to fall asleep' + 2nd
pers. sg. ind. pres. indef. \\
\textbf{elbeszél} \emph{el-} adv., conj. `afar, far away'
+ \emph{beszél} `to speak, to talk' $\rightarrow$ lit. `to narrate' \\
\textbf{eleget} \emph{elég} `enough' +
acc. \emph{-et} \textbar\textbar{} adv. `quite, rather, fairly' \\
\textbf{eleitől} \emph{elei} `beginning, start' + ablat. \emph{-től} `without,
from' \\
\textbf{elejt} `to drop' + 3rd pers. sg. ind. pres. indef. \\
\textbf{elejtettem} vb. \emph{elejt} `to drop' + 1st
pers. sg. ind. past indef. \\
\textbf{elfárad} perfect. vb. pref. \emph{el-} + vb. \emph{fárad} `to
tire, to be tired' + 3rd pers. sg. ind. pres. indef. \\
\textbf{elfoszlott} perfect. vb. pref. \emph{el-} + vb. \emph{fozslik}
`to vanish, unravel' + 3rd pers. sg. ind. past indef. \\
\textbf{elhull} perfect. vb. pref. \emph{el-} + vb. \emph{hull} `to fall' \\
\textbf{elidőz} vb. perf. pref. `el-' + \emph{idő} `time' +
vb.-forming \emph{-z} $\rightarrow$ `to tarry, linger' \\
\textbf{ellened} post. \emph{ellen} `against' + 2nd
pers. sg. single\hyp{}poss. poss. \emph{-ed} `your' \\
\textbf{elmerengnem} lit. formal, perfect. vb. pref. \emph{el-} + vb. \emph{mereng}
`to ponder, to mull over smth' + inf. 1st pers. sg. \\
\textbf{elmerítem} vb. perf. pref. `el-' + \emph{merít} `to
submerge, to dip, to plunge, to sink' \textbar\textbar{} `to draw a
liquid, to scoop, to drain', fig. `to take from a source
(material)' + 1st pers. sg. ind. pres. def. \\
\textbf{elmondom} vb. \emph{elmond} `to tell' + 1st
pers. sg. ind. pres. def. \emph{-om} \\
\textbf{elmorzsol} perfect. vb. pref. \emph{el-} +
vb. \emph{morzsol} `to crumble' \\
\textbf{elmosolyodnak} vb. pref. \emph{el-}
+ \emph{mosoly} `smile' $\rightarrow$ `the process of smiling' +
vb.-forming \emph{-odik} `to become smth' (here `smile') + 3rd
pers. ind. pres. indef. $\rightarrow$ `they start to smile' \\
\textbf{elmulástól} \emph{elmulás} `passage away, death' + \emph{-tól}
`from' \\
\textbf{elnézem} vb. pref. \emph{el-} `continuity over a long time' +
vb. \emph{nez} `to look at' \\
\textbf{eloldja} vb. pref. \emph{el-} `away' + vb. \emph{old} `to
  unbind' + 3rd pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{elő} vb. pref. `front, foreground (action)' \\
\textbf{előtt} \emph{elő} `the front part of smth' + locat. \emph{-tt}
  `in front of, ahead of, before, prior to' \\
\textbf{elpirulsz} perfect. vb. pref. \emph{el-} + vb. \emph{pirul}
`to blush' + 2nd pers. sg. ind. pres. indef. \emph{-sz} \\
\textbf{elromoljanak} perfect. vb. pref. \emph{el-} +
vb. \emph{romlik} `to break down, to deteriorate' + 3rd
pers. pl. subj. indef. \emph{-janak} \\
\textbf{eltöm} vb. perf. pref. `el-' + vb. \emph{töm} `to fill, to
stuff' + 3rd pers. sg. ind. pres. indef. \\
\textbf{elüt} vb. pref. \emph{el-} `away' + vb. \emph{üt} `to hit' +
3rd pers. sg. ind. pres. indef. \\
\textbf{elvegyültem} perfect. vb. pref. \emph{el-} +
vb. \emph{vegyül} `to mix, blend' + 1st
pers. sg. ind. past. \emph{-tem} \\
\textbf{ember} `person' \\
\textbf{embernek} \emph{ember} `person' + (dat. \emph{-nek} `to,
for, as' \textbar\textbar{} attr. poss. \emph{-nek}) \\
\textbf{emlék} `memory' \\
\textbf{emlékezem} \emph{emlék} `memory' $\rightarrow$
vb. \emph{emlékezik} `to remember, to recall' + 1st
pers. sg. ind. pres. indef. \\
\textbf{emlékezés} \emph{emlék} `memory' $\rightarrow$
vb. \emph{emlékezik} `to remember, to recall' +
deverb. noun\hyp{}forming \emph{-és} $\rightarrow$ `recollection,
remembrace, reminiscence' \\
\textbf{engem} pron. \emph{én} `I' + acc. $\rightarrow$ `me' \\
\textbf{enni} inf. of \emph{eszik} `to eat' \\
\textbf{enyhe} \emph{enyh} `light, mild, tender, faint' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-e} `of' \\
\textbf{enyém} \emph{én} `I' + 1st
pers. sg. single\hyp{}poss. \emph{-em} $\rightarrow$ `mine' \\
\textbf{erdő} `forest' \\
\textbf{erdőn} \emph{erdőn} `forest' + iness. \emph{-n} `in' \\
\textbf{erejébe} \emph{erő} `force, power, strength' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-je} + illative \emph{-be}
`into (the inside of)' \\
\textbf{erre} pron. \emph{ez} `this, it' + subl. sg. \emph{-re} `onto,
to' \\
\textbf{esőt} vb. \emph{es}, inf. \emph{esik} `to fall' + pres. part. \emph{-ő} + acc. \emph{-t} \\
\textbf{est} arch. `(in the) evening' \\
\textbf{este} arch.  `(in the) evening'
(syn. \emph{est}) \textbar\textbar{} noun `evening' \\
\textbf{esteledik} \emph{este} `evening' \\
\textbf{eszem} vb. \emph{esz} `to eat', rare without verbal pref.,
inf. \emph{enni} + 1st pers. sg. ind. pres. def. \emph{-em}
  $\rightarrow$ \emph{eszem} \\
\textbf{etess} vb. \emph{eszik} `to eat' +
  causal \emph{-etet} \emph{-tet} $\rightarrow$ \emph{etet} `to feed'
  + 2nd pers. sg. subj. indef. \\
\textbf{ezer} pron. `anything, nothing (negative context)' \\
\textbf{ezért} conj. `therefore' \\
\textbf{ezt} \emph{ez} `this, it' + acc. \emph{-t} \\
%%
\ \\
\noindent \textbf{\large É} \\
\textbf{ébren} adv. `awake' \\
\textbf{édes} adj. `sweet' \\
\textbf{édesebb} comp. `sweeter' \\
\textbf{ég} `sky' \textbar\textbar{} vb. `to burn, to be lit' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{égen} \emph{ég} `sky' + super. \emph{-en} `on' \\
\textbf{égnek} vb. `to burn, to be lit' + 3rd pers. pl. ind. pres. indef. \\
\textbf{égre} \emph{ég} `sky' + subl. sg. \emph{-re} `onto' \\
\textbf{égről} \emph{ég} `sky' + delat. \emph{-ől} `from, about, off' \\
\textbf{éj} `night' \\
\textbf{éjben} poetic \emph{éj} `night' + iness. \emph{-ben} `in' \\
\textbf{éjjel} poetic \emph{éj} `night' + instr. \emph{-vel}
  `with' $\rightarrow$ \emph{éjjel} \\
\textbf{éjjelt} \emph{éjjel} `(with the) night' + acc. \emph{-t} \\
\textbf{éjszaka} \emph{éj} `night' + \emph{szaka} `period of'
    $\rightarrow$ `at night, nighttime, night' \\
\textbf{él} vb. \emph{él} `to live' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{él-e} vb. \emph{él} `to live' + 3rd
pers. sg. ind. pres. indef. + polar interr. \emph{-e} \\
\textbf{élet} `life' \\
\textbf{életet} \emph{élet} `life' + acc. \emph{-et} \\
\textbf{élénken} vb. \emph{él} `to live, exist' + rare
adj.\hyp{}forming \emph{-énk} + super. \emph{-en} `on' $\rightarrow$
`lively' \\
\textbf{élhetek} vb. \emph{él} `to live' + pot. 1st
pers. sg. ind. pres. indef. \\
\textbf{élni} vb. \emph{él} `to live' + inf. \\
\textbf{élted} vb. \emph{él} `to live' + 2nd
pers. sg. inform. ind. past. indef. \\
\textbf{éltem} vb. \emph{él} `to live' + 1st pers. sg. ind. past \\
\textbf{én} pron. `I' \\
\textbf{énekellek} vb. \emph{énekel} `to sing' + 1st
pers. sg. ind. pres. indef. 2nd obj. \\
\textbf{éppolyan} pref. \emph{épp-} `exactly' + arch. pron. \emph{oly}
`such, so' + deadj. adv.\hyp{}forming \emph{-an} $\rightarrow$ adv. `the
same kind, just as, just like' \\
\textbf{érdemelnék} vb. \emph{érdemel} `to merit, deserve' + 1st
pers. sg. cond. pres. indef. \emph{-nék} \\
\textbf{ért} vb. \emph{ért} `to reach', arch. `to touch', fig. `to
understand' + 3rd pers. sg. past indef. \\
\textbf{érte} pron. `for him, her, it'. \\
\textbf{értem} \emph{ért} vb. \emph{ért} `to reach', arch. `to touch',
fig. `to understand' + 1st pers. sg. ind. pres. def. \\
\textbf{érez} vb. `to feel, smell, taste' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{éreztem} vb. \emph{érez} `to feel', 1st pers. sg. past def. \\
\textbf{érzed} vb. \emph{érez} `to feel' + 2nd
pers. sg. ind. pres. def. \\
\textbf{érzelmek} \emph{érzelm} `emotion, feeling' + pl. \emph{-ek} \\
\textbf{érzem} vb. \emph{érez} `to feel' + 1st
pers. sg. ind. pres. def. \\
\textbf{érzéketlen} vb. \emph{érez} `to feel' +
privative \emph{-etlen} \\
\textbf{éhezem} vb. \emph{éhezik} `to starve, to yearn (fig.)' + 1st
pers. sg. ind. pres. def. \\
\textbf{éveket} \emph{év} `year' + pl. \emph{-ek} + acc. \emph{-et} \\
%%
\ \\
\noindent \textbf{\large F} \\
\textbf{fa} `tree' \\
\textbf{fadobálók} \emph{fa} `tree' + vb. \emph{dob} `to throw' +
repetitive \emph{-ál} + pl. \emph{-ók} \\
\textbf{fagy} `frost' \\
\textbf{fagyos} adj. `chilly, wintry, icy' \\
\textbf{fagyosabb} adj. \emph{fagyos} `chilly, wintry, icy' +
comp. \emph{-abb} `more' \\
\textbf{fakad} vb. `to arise, to originate', `to blossom' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{falam} \emph{fal} `wall' + 1st pers. sg. poss. single\hyp{}poss. \emph{-am} \\
\textbf{favágókat} \emph{fa} `tree' + \emph{vágó} `cutter' +
pl. \emph{-k} + acc. \emph{-at} \\
\textbf{fájdalom} vb. \emph{fáj} `to ache, to hurt' +
noun\hyp{}forming \emph{-dalom} $\rightarrow$ `pain, grief, sorrow' \\
\textbf{fájlalom} vb. \emph{fáj} `to hurt' + vb.-forming \emph{-lal}
$\rightarrow$ lit., arch. `to regret' + 1st
pers. sg. ind. pres. def. \\
\textbf{fákra} \emph{fa} `tree' + pl. \emph{-k} + subl. \emph{-ra}
`onto' $\rightarrow$ \emph{fákra} \\
\textbf{fán} \emph{fa} `tree' + inness. \emph{-n} `in' \\
\textbf{fáradt} vb. \emph{fárad} `to tire' + past. part. \emph{-t} \\
\textbf{fáradtan} vb. \emph{fárad} `to tire' + past. part. \emph{-t} +
adv.\hyp{}forming \emph{-an} $\rightarrow$ `tiredly, wearily' \\
\textbf{fát} \emph{fa} `tree' + acc \emph{-t} \\
\textbf{fázom} vb. \emph{fázik} `to be/feel cold' + 1st pers. sg. ind.
pres. indef. \\
\textbf{fecsegő} vb. \emph{fecseg} `to chatter' + pres. part. \emph{-ő} \\
\textbf{fegyvert} \emph{fegyever} `weapon' + acc. \emph{-t} \\
\textbf{fehérhabú} \emph{fehér} `white' + \emph{hab} `foam'
+ adj.\hyp{}forming for nouns \emph{-ú} \\
\textbf{fejedben} \emph{fej} `head' + 2nd
pers. sg. single\hyp{}poss. poss. \emph{-ed} `your' +
iness. \emph{-ben} `in' \\
\textbf{fejsze} `axe' \\
\textbf{fejünk} \emph{fej} `head' + 1st
pers. pl. single\hyp{}poss. poss. \\
\textbf{fekszenek} vb. inf. \emph{fekszik} `to lie (horizontally)' +
3rd pers. pl. ind. pres. indef. \emph{-enek} \\
\textbf{fekszünk} vb. inf. \emph{fekszik} `to lie (horizontally)' +
1st pers. pl. ind. pres. indef. \emph{-ünk} \\
\textbf{feküdtem} vb. inf. \emph{fekszik} `to lie (horizontally)'  +
1st pers. sg. ind. past \\
\textbf{feledni} lit. vb. \emph{feled} `to forget',
alt. \emph{felejt} \\
\textbf{felejtek} vb. \emph{felej} `to forget' + 1st pers. sg. pres. indef. \\
\textbf{felelhetek} \emph{fél} `human being' + vb.-forming \emph{-el}
$\rightarrow$ vb. \emph{felel} `to answer, reply' + poten. 1st
pers. sg. ind. pres. indef. \\
\textbf{felellek} vb. \emph{felel} `to answer' + 1st
pers. sg. ind. pres. indef. 2nd obj. \\
\textbf{felém} post. \emph{felé} `to[wards]' + 1st
pers. sg. poss \emph{-m} $\rightarrow$ pron. `towards me' \\
\textbf{felhő} `cloud, danger, trouble' \\
\textbf{felhők} \emph{felhő} `cloud, danger, trouble' +
pl. \emph{-k} \\
\textbf{feljöjjön} vb. pref. \emph{fel-} `upwards' (see
also \emph{föl-}) + vb. \emph{jön} `to come' + 3rd
pers. sg. subj. indef. \\
\textbf{felleg} lit. `cloud' \\
\textbf{fellegekkel} lit. \emph{felleg} `cloud' + pl. \emph{-ek} +
instr. \emph{-kel} \\
\textbf{felnőttet} \emph{felnőt} `grown-up, adult' + acc. \emph{-et} \\
\textbf{fenyőmről} \emph{fenyő} `pine (tree)' + 1st pers. sg. single\hyp{}poss. \emph{-m} +
delat. \emph{-ről} `off, from' \\
\textbf{föllebeghess} vb. pref. \emph{fel-} `upwards' (see
also \emph{föl-}) + vb. \emph{lebeg} `to float (in the air)' + 2nd
pers. sg. inform. poten. subj. \\
\textbf{fönn} adv. alt. \emph{fenn}, \emph{fent} `above' \\
\textbf{félelem} vb. \emph{fél} `to be afraid' +
noun\hyp{}forming \emph{-elem} $\rightarrow$ `fear' \\
\textbf{félnék} vb. \emph{fél} `to fear, to be afraid of' + 1st
pers. sg. cond. indef. \\
\textbf{félni} vb. inf. \emph{fél} `to fear, to be afraid of' \\
\textbf{féltelek} vb. \emph{félt} `to fear for, care deeply for' +  1st pers. sg. ind. pres 2nd obj. \emph{-elek} \\
\textbf{féltem} vb. \emph{fél} `to be afraid' + acc. \emph{-t}
$\rightarrow$ vb. \emph{félt} `to fear for smb' + 1st
pers. sg. ind. past def. \\
\textbf{fény} `light,  glitter', fig. `happiness, joy' \\
\textbf{fénye} \emph{fény} `light' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-e} \\
\textbf{fényes} \emph{fény} `light, glitter' + adj.\hyp{}forming \emph{-es}
$\rightarrow$ `bright' \\
\textbf{fényesül} \emph{fény} `light, glitter' + adj.\hyp{}forming \emph{-es}
$\rightarrow$ `bright' + adv.\hyp{}forming \emph{-ül} \\
\textbf{fényességgel} \emph{fény} `light' + adj.\hyp{}forming \emph{-es}
    `with, having' $\rightarrow$ `bright' + abst. noun\hyp{}forming
  \emph{-ség} $\rightarrow$ `brightness' +
  instr. \emph{-vel} `with' (\emph{-gel} after consonant
  \emph{g}) $\rightarrow$ `with brightness' \\
\textbf{fényben} \emph{fény} `light, glitter' + iness. \emph{-ben}
`in' \\
\textbf{fénylenek} \emph{fény} `light' +
freq. vb.\hyp{}forming \emph{-lik}: `to glitter' + 3rd
pers. pl. ind. pres. indef. \emph{-ek} \\
\textbf{fértem} vb. \emph{fér} `to fit, have enough room' + 1st
pers. sg. past indef. \emph{-tem} \\
\textbf{fiú} `boy' \\
\textbf{fizettem} vb. \emph{fizet} `to pay' (for smth \emph{-ért},
smb \emph{-nak/-nek}) + 1st pers. sg. ind. past \\
\textbf{fodort} \emph{fodor} `ripple', `ruffle', `curl (hair)' +
acc. \emph{-t} \\
\textbf{fog} vb. `to hold, capture, catch' + 3rd
pers. sg. ind. pres. indef. \textbar\textbar{} aux. vb. `to will' \\
\textbf{fogad} vb. `to receive' + 3rd pers. sg. ind. pres. indef. \\
\textbf{fogaskerekére} \emph{fogas} `tooth' + \emph{kerék} `wheel':
`cogwheel' + 3rd pers. sg. single\hyp{}poss. poss. \emph{-e} + subl. \emph{-re}
`towards' \\
\textbf{fogatlan} \emph{fog} `tooth' +
privative adj.\hyp{}forming \emph{-atlan} `-less' \\
\textbf{foglak} vb. aux. \emph{fog} `to will' + 1st
pers. sg. ind. pres. 2nd obj. \emph{-lak} `you' \\
\textbf{foglalkozz} vb. \emph{foglalkozik} `to study, to take care
of' + 2nd pers. sg. inform. subj. indef. \\
\textbf{foglyul} vb. \emph{fog} `to hold' + noun\hyp{}forming \emph{-oly} $\rightarrow$ \emph{fogly} `prisoner' + essive
  \emph{-ul} `as, with the intention of' \\
\textbf{fogni} vb. inf. \emph{fog} `to will' \\
\textbf{fogod} aux. vb. \emph{fog} `to will' + 2nd
pers. sg. ind. pres. def. (e.g. \emph{Meg fog bánni.}) \\
\textbf{fogva} post. `from, since' (with ablat. \emph{-tól, -től}),
with implied continuity \\
\textbf{fogsz} aux. vb. `to will' + 2nd
pers. sg. inform. ind. pres. def. \\
\textbf{fojtogatja} vb. \emph{fojtogat} `to
choke, to strangle, to drown, to suffocate' + 3rd
pers. sg. ind. pres. def. \emph{-ja} (see \emph{fojt}) \\
\textbf{fojtsanak} emph. vb. pref. \emph{meg-} + vb. \emph{fojt} `to
choke, to strangle, to drown' + 3rd pers. pl. subj. indef. \\
\textbf{folytatni} inf. of \emph{folytat} `to continue' \\
\textbf{folyó} vb. inf. \emph{folyik} `to flow, go on' +
pres. part. \emph{-ó} $\rightarrow$ `river' \textbar\textbar{}
adj. `flowing, leaky, current' \\
\textbf{fordul} vb. `to spin, to turn around (loop)' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{forgács} `chip (of split wood)' \\
\textbf{forgót} `windmill (toy, US pinwheel)' \\
\textbf{föl} adv. syn. \emph{fel} `upwards' \\
\textbf{földek} \emph{föld} `earth, ground' + pl. \emph{-ek} \\
\textbf{földön} \emph{föld} `earth, ground' +
super. \emph{-ön} `on' \\
\textbf{földre} \emph{föld} `earth, ground' + subl. \emph{-re}
`onto' \\
\textbf{fölkereshetnéd} \emph{föl-} `upwards' + \emph{keres} `to
seek': `to visit' + pot. `may, might' \emph{-het} + 2nd
pers. sg. ind. pres. cond. \emph{-ned}, alt. \emph{felkeres} \\
\textbf{fölállok} \emph{föl-} `upwards' + vb. \emph{áll} `to stand' + 1st
  pers. sg. ind. pres. indef. \emph{-ok}, alt. \emph{felállok} \\
\textbf{földeken} \emph{föld} `earth, field' + pl. \emph{-ek} +
super. \emph{-en} `on' \\
\textbf{földtől} \emph{föld} `earth' + ablat. \emph{-től} `without,
from' \\
\textbf{fölfeslik} vb. pref. \emph{föl-} `upwards'+ arch., poetic,
vb. \emph{feslik} `to come unstitched', 3rd
pers. sg. ind. pres. indef. \\
\textbf{fölnéztem} vb. pref. \emph{föl-} `upwards' + vb.
  \emph{néz} `to look' + 1st pers. sg. past. def. \emph{-tem} \\
\textbf{fölött} \emph{föl-} `up[wards]' +  locat. \emph{-ött}:
`above', alt. \emph{elett} \\
\textbf{fölöttem} \emph{fölött} `above' + 1st pers. sg. \emph{-m}
$\rightarrow$ `above me' \\
1st pers. sg. ind. pres. indef. \emph{-ok}, alt. \emph{felállok} \\
\textbf{fölszabadult} vb. pref. \emph{föl-} completion (see
also \emph{fel-}) + \emph{szabad} `free' + vb.-forming \emph{-ul}
    $\rightarrow$ `to set free' + past part. \emph{-t} \\
\textbf{fönn} adv. `above', alt. `fent, fenn' \\
\textbf{fönt} adv. `above, up, awake, upstairs', alt. \emph{fent} \\
\textbf{futnak} vb. \emph{fut} `to run' + 3rd
pers. pl. ind. pres. indef. \emph{-nak} \\
\textbf{fülelt} \emph{fül} `ear' + vb.-forming \emph{-el}: `to listen
(intently)' + 3rd pers. ind. past indef. \emph{-t} \\
\textbf{fülke} \emph{fül} `semi\hyp{}circular obj., shape' +
dim. \emph{-ke} $\rightarrow$ `compartment' \\
\textbf{füvek} \emph{fű} `grass' + pl. \emph{-ek} \\
%%
\ \\
\noindent \textbf{\large G} \\
\textbf{gally} `twig' \\
\textbf{gallya} `twig' + 3rd pers. sg. poss. single\hyp{}poss. \emph{-a} \\
\textbf{gond} arch. sg. `care' \\
\textbf{gondolkozom} vb. \emph{gondol} `to think, plan, guess, worry'
+ 1st pers. sg. ind. pres. indef. \\
\textbf{Göncölök} \emph{Göncöl} + pl. \emph{-ök} $\rightarrow$ `the Great Bear'
  (constellation) \\
\textbf{göndör} adj. `curly' \\
\textbf{görbe} adj. `curved, crooked, bent' \\
\textbf{görcse} \emph{görcs} `knot (in a tree)' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-e} \\
\textbf{görnyedve} vb. \emph{görnyed} `to bend (under a burden)' +
adv.\hyp{}forming \emph{-ve} \\
\textbf{görög} vb. `to roll, scroll' + 3rd
pers. sg. ind. pres. indef. \textbar\textbar{} adj. `Greek' \\
\textbf{gőzei} \emph{gőz} `steam, vapour' + 3rd
pers. sg. poss. \emph{-e} + pl. poss. \emph{-i} \\
\textbf{gyenge} adj. `weak', alt. \emph{gyönge} \\
\textbf{gyere} vb. \emph{jön} `to come' + 2nd
pers. sg. ind. pres. indef. \\
\textbf{gyermek} `child', alt. \emph{gyerek} \\
\textbf{gyermekké} \emph{gyermek} `child' + translative \emph{-ké}
`to become like (a child)' \\
\textbf{gyilkona} vb. \emph{gyilkol} `to kill' + 3rd
pers. sg. cond. pres. indef. \\
\textbf{gyom} `weed' \\
\textbf{gyönge} alt. adj. \emph{gyenge} `weak, feeble (physically)' \\
\textbf{gyöpén} \emph{gyep} `lawn, grass' + super. \emph{-en} `on' \\
\textbf{gyűljetek} vb. inf. \emph{gyűlik} `to assemble, to gather, to muster' + 2nd
pers. pl. informal imper. \emph{-jetek} \\
%%
\ \\
\noindent \textbf{\large H} \\
\textbf{ha} conj. `if', `when, once' \\
\textbf{hadd} lex. of imper. \emph{hagyd} `let her (+ subj. vb.)' \\
\textbf{hajamra} \emph{haj} `hair (head, not facial)' + 1st
pers. sg. single\hyp{}poss. \emph{-am} + subl. \emph{-ra} `on[to]' \\
\textbf{hajnal} `dawn' \\
\textbf{hajnalban} `dawn' + iness. \emph{-ban} `in' \\
\textbf{hajam} \emph{haj} `hair (not beard)' + 1st
pers. sg. single\hyp{}poss. \emph{-am} \\
\textbf{hajósinas} \emph{hajó} `ship' + noun suffix forming an
occupation \emph{-s} $\rightarrow$ `sailor' + \emph{inas} `apprentice'
$\rightarrow$ `ship apprentice' \\
\textbf{hajót} \emph{hajó} `ship' + acc. \emph{-t} \\
\textbf{hajóz} vb. `to navigate, sail' \\
\textbf{hagyna} vb. \emph{hagy} `to allow, leave' + 3rd
pers. sg. cond. pres. indef. \emph{-na} \\
\textbf{halál} vb. \emph{hal} `to die' + arch. noun\hyp{}forming \emph{-ál} $\rightarrow$ `death' \\
\textbf{halálra} vb. \emph{hal} `to die' + arch. noun\hyp{}forming
  \emph{-ál} $\rightarrow$ `death' + subl. `on[to], for' \emph{-ra}
  $\rightarrow$ `to death' \\
\textbf{halász} `fisherman'\\
\textbf{halk} adj. `muffled (of sound)', `quiet, inaudible (of
  movement)' \\
\textbf{hallak} vb. \emph{hall} `to hear' + 1st
pers. sg. ind. pres. 2nd obj. \\
\textbf{hallgatag} vb. \emph{hallgat} `to remain silent, to listen' +
adj.\hyp{}forming \emph{-ag} $\rightarrow$ `silent' \\
\textbf{hallgatok} vb. \emph{hallgat} `to remain silent, to listen' +
1st pers. sg. ind. pres. indef. \emph{-ok} \\
\textbf{hallottam} vb. \emph{hall} `to hear' + 1st pers. sg. ind. past
def. \emph{-ottam} \\
\textbf{halni} vb. inf. \emph{hal} `to die' \\
\textbf{halok} vb. \emph{hal} `to die' +  1st
pers. sg. ind. pres. indef. \\
\textbf{halom} `pile, heap' \\
\textbf{halomba} \emph{halom} `pile, heap' + illative \emph{-ba} `into (the inside of)' \\
\textbf{halott} noun./adj. `dead' \\
\textbf{hanem} \emph{ha} conj. `if', `when, once' + \emph{nem} `no':
conj. `but (following a negative sentence, making it positive)' \\
\textbf{hangosan} adv. `loudly' \\
\textbf{haragszol} \emph{harag} `anger' + vb.-forming \emph{-szol}
$\rightarrow$ vb. inf. \emph{haragszik} `to be angry` + 2nd
pers. inform. sg. ind. pres. indef. \\
\textbf{harangszó} \emph{harang} `(church) bell'+ \emph{szó} `word,
voice' $\rightarrow$ `bell chime' \\
\textbf{harapdált} vb. \emph{harap} `to bite' + freq. \emph{-dál}
$\rightarrow$ `to nibble, gnaw, champ' + 3rd pers. sg. ind. past. indef. \\
\textbf{harcban} \emph{harc} `battle, fight' + iness. \emph{-ban} `in' \\
\textbf{harmattá} \emph{harmat} `dew' + transl. \emph{-vá} `(change,
turn) into' \\
\textbf{harmatos} adj. `dewy' \\
\textbf{harminc} `thirty' \\
\textbf{hasáb} `firewood' (a split log) \\
\textbf{hasábfák} \emph{hasábfa} `log' + pl. \emph{-k} \\
\textbf{hasított} vb. \emph{hasít} `to cleave, split' +
past. part. \emph{-ott} \\
\textbf{hasonlitott} vb. \emph{hasonlít} `to resemble, to compare to'
+ 3rd pers. sg. ind. past. indef. \emph{-ott} \\
\textbf{hatalma} \emph{hatalom} `power' + 3rd
  pers. sg. single\hyp{}poss. poss. \emph{-a} \\
\textbf{határon} \emph{határ} `limit, boundary', `border, outskirt,
frontier, verge' + super. \emph{-on} `on' \\
\textbf{hazaballagunk} \emph{haza} `home' (\emph{haz} arch. `house' (modern
\emph{ház}) + vb. \emph{ballag} `to walk slowly, to stroll' + 1st
pers. pl. ind. pres. indef. \\
\textbf{hazafelé} \emph{haza} `home' (\emph{haz} arch. `house' (modern
\emph{ház}) + short. arch. lative (non\hyp{}extant case in modern
Hungarian) \emph{-á} $\rightarrow$ `(to) home') + \emph{felé}
`towards' $\rightarrow$ adv. `homewards' \\
\textbf{hazám} \emph{ház} + 1st
pers. sg. single\hyp{}poss. \emph{-am} \\
\textbf{hazatalálnak} \emph{haza} `home' (\emph{haz} arch. `house'
(modern \emph{ház}) + vb. \emph{talál} `to come across, to find'
$\rightarrow$ `to fine one's way home' \\
\textbf{hálóm} \emph{háló} `net' + 1st
pers. sg. poss. single\hyp{}poss. \emph{-m} \\
\textbf{hálót} \emph{háló} `net' + acc. \emph{-t} \\
\textbf{hámot} \emph{hám} `harness' + acc. \emph{-ot} \\
\textbf{háromszoros} \emph{három} `three' + \emph{-szoros} `-fold' \\
\textbf{hát} `back (body)', 'well...', `then, and, but' (questioning
back), `surely' \\
\textbf{házat} \emph{ház} `house' + acc. \emph{-at} \\
\textbf{háziúr} \emph{ház} `house' + adj.\hyp{}forming `of the' \emph{-i}
+ \emph{úr} `master, lord' \\
\textbf{házon} \emph{ház} `house' + super. \emph{-on}
`on (that manner)' \\
\textbf{helyemet} \emph{hely} `place, space' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-em} `my' + acc. \emph{-et} \\
\textbf{helyükön} \emph{hely} `place, space' + 3rd
pers. pl. single\hyp{}poss. poss. \emph{-ük} `their' +
super. \emph{-ön} `on' \\
\textbf{hever} vb. `to lie (down)' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{heves} adj. `violent, passionate, intense' \\
\textbf{hevül} vb. `to grow hot', 3rd pers. sg. ind. pres. indef. \\
\textbf{hiába} arch. adv. `in vain' \\
\textbf{hiányod} \emph{hiány} `absence' + 2nd pers.
sg. single\hyp{}poss. \emph{-d} \\
\textbf{hiányozni} \emph{hiány} `lack, absence, break' +
vb.-forming \emph{-zik} $\rightarrow$ `(smb) to be missing, missed',
inf. \emph{hiányozni} \\
\textbf{hibbant} perfect. vb. pref. \emph{el-} + vb. \emph{hibban}
  `to become imbalanced', arch. + past part. \emph{-t} \\
\textbf{hinnéd} vb. \emph{hisz} `to believe' + 2nd
pers. sg. inform. cond. pres. def. \\
\textbf{hinném} vb. \emph{hisz} `to believe' + 1st
pers. sg. cond. pres. def. \\
\textbf{hisz} conj. abbrev. \emph{hiszen} `as, for, because (smth obvious)' \\
\textbf{hiszen} conj. `as, because (smth obvious)' \\
\textbf{hitegetve} vb. \emph{hisz} `to believe' + caus. \emph{-et}
$\rightarrow$ vb. \emph{hitet} `to make smb believe smth' +
freq. vb.\hyp{}forming \emph{-get} + adv. part. \emph{-ve} result of action
or simultaneous actions \\
\textbf{hittem} vb. \emph{hisz} `to believe' + 1st pers. ind. past. \\
\textbf{hiú} adj. `vain' \\
\textbf{hídja} \emph{híd} `bridge' + 3rd
pers. sg. poss. single\hyp{}poss. \emph{-ja} \\
\textbf{hívj} vb. \emph{hív} `to call, name' + 2nd pers. sg. informal subj. indef. \\
\textbf{hívták} vb. \emph{hív} `to call, name' + 3rd
pers. pl. ind. past \\
\textbf{hogy} adv. interr. `how', conj. `that', `if'. \\
\textbf{hogy'} short. \emph{hogyan} `how' \\
\textbf{hogyha} coloq. conj. syn. \emph{ha} `if', `when, once, supposing' \\
\textbf{hold} `moon' \\
\textbf{holdat} \emph{hold} `moon' + acc. \emph{-at} \\
\textbf{holott} adv. arch. `where' \textbar\textbar{}
conj. lit. `though, but, for, since' \\
\textbf{holt} `dead' \\
\textbf{holtan} \emph{holt} `dead' + adv.\hyp{}forming \emph{-an} \\
\textbf{homályként} \emph{homály} `blur' + essive formal \emph{-ként}
`as' \\
\textbf{homlok} `forehead' \\
\textbf{homlokomra} \emph{homlok} `forehead' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-om} + subl. \emph{-ra}
`on[to]' \\
\textbf{homokon} \emph{homok} `sand' + super. \emph{-on} `on' \\
\textbf{hová} alt. \emph{hova} `where'  \\
\textbf{hozták} vb. \emph{hoz} `to carry, to bring', `to set smth, smb
into a state' + 3rd pers. pl. ind. past \emph{-ták} \\
\textbf{hozzá} pron. `to it, him, her' \\
\textbf{hozzád} \emph{hozzá} pron. `to it, him, her' + 2nd pers. sg. \\
\textbf{hull} vb. `to fall', alt. \emph{hullik} + 3rd
pers. sg. ind. pres. indef. \\
\textbf{hullámok} \emph{hullám} `wave' + pl. \emph{-ok} \\
\textbf{hullámzanak} vb. \emph{hullámzik} `to wave, billow, undulate' +
3rd pers. pl. ind. pres. indef. \\
\textbf{hullong} vb. `to fall', alt. \emph{hullik} + freq. \emph{-ong} \\
\textbf{hunyorgott} vb. \emph{hunyorg} `to squint' + 3rd
pers. sg. ind. past. indef. \emph{-rgott} \\
\textbf{huzat} vb. \emph{húz} `to pull' +
noun\hyp{}forming \emph{-at} $\rightarrow$ `a cover (of an obj.),
`draught (current of air)' \\
\textbf{húst} \emph{hús} `meat' + acc. \emph{-t} \\
\textbf{hűs} adj. `cool (water, weather)' \\
\textbf{hűség} adj. \emph{hű} `faithful' + \emph{-ség}
`state of' $\rightarrow$ `loyalty' \\
\textbf{hűvös} adj. `cool' \\
%%
\ \\
\noindent \textbf{\large I} \\
\textbf{idebent} adv. \emph{ide} `here' + \emph{bent} `within' \\
\textbf{ideg} `nerve, bowstring' \\
\textbf{ideges} \emph{ideg} `nerve' + adj.\hyp{}forming \emph{-es} \\
\textbf{idegeim} \emph{ideg} `nerve' + 1st
pers. sg. poss. mult.\hyp{}poss. \emph{-eim} \\
\textbf{idő} `time' \\
\textbf{ifju} \emph{ifjú} adj., noun `young' \\
\textbf{ifjúságod} adj. \emph{ifjú} `young' + noun\hyp{}forming \emph{-ság}
(state of being): `youth' + 2nd pers. informal
sg. poss. single\hyp{}poss. \emph{-od}: `your childhood' (informal) \\
\textbf{igazában} \emph{igaz} `true', arch. `honest, righteous' +
iness. \emph{-ban} `in' \\
\textbf{igazán} adv. `really, truly' \\
\textbf{illatosabb} adj. \emph{illatos} `fragant' +
comp. \emph{-abb} \\
\textbf{illenék} vb. \emph{illik} `to suit, match, fit' + 1st or 3rd pers. sg. cond. pres. indef. \\
\textbf{illik} vb. `to suit, match, fit' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{ily} arch. pron. `such, in this manner' \\
\textbf{ilyen} det. `such' \\
\textbf{im} lit., abbrv. \emph{íme} `here is, behold' \\
\textbf{imbolygott} vb. \emph{imbolyog} `to sway'+ 3rd
pers. sg. single\hyp{}poss. poss. \emph{-e} \\
\textbf{immár} lit. `now, already' \\
\textbf{indul} vb. `to set off, to head, to start' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{ingyen} fig. arch. `in vain'; usually `free of charge' \\
\textbf{iramlanak} \emph{ir} `movement' + rare
instr. vb.-forming \emph{-amlik} (inf.) $\rightarrow$ `to rush, hurry'
+ 3rd pers. pl. ind. pres. indef. \emph{-anak} \\
\textbf{is} doublet of \emph{és} `and',
adv. `also' \textbar\textbar{} `even, up to, as much as, as long
as' \textbar\textbar{} (after interr.) `again' \textbar\textbar{}
`sure, enough, indeed' \\
\textbf{ismer} vb. `to know'  + 2nd pers. sg. inform. (or 3rd
pers. sg.) ind. pres. indef.\\
\textbf{istene} \emph{isten} `god' + 3rd
pers. sg. sing. poss. \emph{-e} \\
\textbf{istenhez} \emph{isten} `god' + allat. \emph{-hez} `towards' \\
\textbf{itt} adv. `here, in this place' \\
%%
\ \\
\noindent \textbf{\large Í} \\
\textbf{így} `just so, in this way, as I, like this' \\
\textbf{íme} emph. `behold!, voilà! see!' \\
%%
\ \\
\noindent \textbf{\large J} \\
\textbf{jár} vb. `to move (location), to walk, to go, to dance',
inf. \emph{járni} + 3rd pers. sg. ind. pres. indef. \\
\textbf{jár-kel} vb. \emph{jár} `to move (location)' +
freq. \emph{-kel} \\
\textbf{járni} inf. vb. \emph{jár}  `to move (location), to walk, to
go' \\
\textbf{játékot} \emph{játék} `game + acc. \emph{-ot} \\
\textbf{játszom} vb. inf. \emph{játszik} `to play' + 1st
pers. sg. ind. pres. \\
\textbf{játszott} \emph{játék} `game' + vb.\hyp{}forming \emph{-szik}
$\rightarrow$ vb. \emph{játszik} `to play' +
3rd pers. sg. ind. past. indef. \emph{-ott} \\
\textbf{játszottam} vb. \emph{játszik} `to play' + 1st
pers. sg. ind. past. \emph{-ottam} \\
\textbf{jeges} \emph{jég} `ice' + adj.\hyp{}forming \emph{-es} \\
\textbf{jó} `good' \\
\textbf{jóhúsú} \emph{jó} `good' + \emph{hús} `meat' +
adj.\hyp{}forming \emph{-ú} `quality of' \\
\textbf{jós} `fortune teller, soothsayer' \\
\textbf{jön} vb. `to come' + 3rd pers. sg. ind. pres. indef. \\
\textbf{jönnél} vb. \emph{jön} `to come' + 2nd pers. sg.
cond. pres. indef. \\
\textbf{jött} vb. \emph{jön} `to come' + 3rd pers. sg. ind. past
indef. \\
\textbf{jut} vb. `to fall to someone's lot' \\
%%
\ \\
\noindent \textbf{\large K} \\
\textbf{kabát} `coat' \\
\textbf{kabátot} \emph{kabát} `coat' + acc. \emph{-ot} \\
\textbf{kacagunk} vb. \emph{kacag} `to laugh (a lot)' + 1st
pers. pl. ind. pres. indef. \emph{-unk} \\
\textbf{kalapja} \emph{kalap} `hat' + 3rd
pers. sg. poss. single\hyp{}poss. \emph{-ja} \\
\textbf{kalitkája} \emph{kalitka} `cage' + 3rd
pers. sg. sing. single\hyp{}poss. \emph{-ja} \\
\textbf{kanyarított} \emph{kanyar} `curve, bend, meander, meander's
neck, swerve' + vb.\hyp{}forming \emph{-ít}
$\rightarrow$ `to bend, cut in a circular motion (e.g. a piece of
bread)' + 3rd pers. sg. ind. pres. indef. \\
\textbf{kapás} Farm hand using a hoe, from \emph{kapa} `hoe' \\
\textbf{kapja} vb. \emph{kap} `to receive' + 3rd
pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{kapott} vb. \emph{kap} `to receive' + 3rd
pers. sg. ind. past indef. \emph{-ott} \\
\textbf{karaj} folk. `slice' (syn. \emph{karéj}) \\
\textbf{karja} \emph{kar} `arm' + 3rd
pers. sg. sing. single\hyp{}poss. \emph{-ja} \\
\textbf{karol} vb. `to put one's arm around
someone' \textbar\textbar{} lit. `to embrace' \\
\textbf{kedve} \emph{kedv} `spirit, humour', `disposition, desire' +
3rd pers. sg. single\hyp{}poss. \emph{-e} \\
\textbf{kedves} address `dear' \textbar\textbar{} adj. `kind, pleasant' \\
\textbf{kedvesem} \emph{kedves} `dear' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-em} `my' \\
\textbf{keleti} \emph{kelet} `east' + adj.\hyp{}forming \emph{-i} \\
\textbf{kell} aux. vb. `must, need to, have to + inf.' \\
\textbf{kemény} adj. `hard, tough' \\
\textbf{kenyerem} \emph{kenyér} `bread' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-em} `my' \\
\textbf{kenyeret} \emph{kenyér} `bread' + acc. \emph{-et}
$\rightarrow$ \emph{kenyeret} \\
\textbf{kenyeréből} \emph{kenyér} `bread' + 3rd
pers. single\hyp{}poss. poss. \emph{-é} + elative \emph{-ből} `from, out of' \\
\textbf{keresek} vb. \emph{keres} `to seek' + 1st
pers. sg. ind. pres. indef. \\
\textbf{kerestelek} vb. \emph{keres} `to seek' + 1st
pers. sg. ind. past 2nd obj. \\
\textbf{kerül} vb. `to cost' + 3rd pers. sg. ind. pres. \\
\textbf{kettő} `two' \\
\textbf{keveredsz-e} vb. \emph{kever} `to mix, stir' +
vb.\hyp{}forming \emph{-ed}
+ refl. \emph{-ik} + 2nd pers. inform. sg. ind. pres. indef. + polar
interr. \emph{-e} \\
\textbf{kevés} adj. `few, little (not as many/much as usual or as expected' \\
\textbf{kezed} \emph{kéz} `hand' + 2nd
pers. sg. single\hyp{}poss. poss. \emph{-ed} `your' \\
\textbf{kezem} \emph{kéz} `hand' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-em} `my' \\
\textbf{kék} `blue' \\
\textbf{kéken} \emph{kék} `blue' +
super. \emph{-en} `on' \\
\textbf{kél} vb. `to rise', 3rd pers. sg. ind. pres. def.,
  arch. alt. \emph{kel} \\
\textbf{kényedül} lit. vb. `to indulge', 3rd
pers. sg. ind. pres. indef. \\
\textbf{kényes} adj. `delicate' \\
\textbf{képed} \emph{kép} `picture' + 2nd
pers. sg. single\hyp{}poss. poss. \emph{-ed} \\
\textbf{képeket} \emph{kép} `picture' + pl. \emph{-ek} + acc. \emph{-et} \\
\textbf{képzelhetsz} \emph{kép} `image' + vb.-forming \emph{-zel}:
`to imagine' + pot. \emph{-het} + 2nd
pers. sg. ind. pres. \emph{-sz} \\
\textbf{képzetet} \emph{képzet} `idea, image, notion' +
acc. \emph{-t} \\
\textbf{kérded} arch. vb. \emph{kérd} `to ask, to question' + 2nd
pers. sg. ind. pres. def. \\
\textbf{kérdezed} vb. \emph{kérdez} `to ask' + 2nd
pers. sg. ind. pres. \\
\textbf{kérek} vb. \emph{kér} `to ask for' + 1st
pers. sg. ind. pres. indef. \emph{-ek} \\
\textbf{kérlelhetetlen} vb. \emph{kérlel} `to beg, to implore' +
pot. \emph{-het} + privative \emph{-etlen} $\rightarrow$
adj.\hyp{}forming \emph{-hetetlen} \\
\textbf{kérő} vb. \emph{kér} `to request, ask' +
pres. part. \textbar\textbar{} `suitor, requester' \\
\textbf{kés} `knife' \\
\textbf{kész} adj. `done, ready, finished' \\
\textbf{két} `two' (before a noun) \\
\textbf{kétségbeesés} `despair' (from \emph{két} `two' + \emph{-ség}
`state of' $\rightarrow$ `duality', implying division, doubt) and
vb. \emph{eszik} `to fall' \\
\textbf{kézen} \emph{kéz} `hand' + super. \emph{-en} `on' \\
\textbf{ki} pron. rel. arch. `he who',
alt. \emph{aki} \textbar\textbar{} adv. `out[wardly]' \\
\textbf{kiaggatom} vb. pref. \emph{ki-} `outwardly' + vb. \emph{aggat}
`to hang smth [on smb]' + 1st pers. sg. ind. pres. def. \emph{-om} \\
\textbf{kiben} \emph{ki} `who' +
iness. sg. \emph{-ben} `in' $\rightarrow$ `in who(m)' \\
\textbf{kieszeltem} vb. \emph{kieszel} `to devise' + 1st
pers. sg. ind. past. \emph{-tem} \\
\textbf{kikért} \emph{ki} + pl. \emph{-k} +
causal\hyp{}final \emph{-ért} $\rightarrow$ `for whom (pl.)' \\
\textbf{kinek} \emph{ki} pron. rel. arch. `he who' + dat. \emph{-nek} `to,
for' $\rightarrow$ `to whom' \\
\textbf{kinn} `outside' \\
\textbf{kinőttek} vb. \emph{kinő} `to outgrow' + 3rd
pers. pl. ind. past \\
\textbf{kint} adv. `outside' \\
\textbf{kiolthatnám} vb. pref. \emph{ki-} `outwardly' + \emph{olt} `to
extinguish, to put out' + potential \emph{-hat} + 1st
pers. sg. cond. pres. def. \emph{-nám} \\
\textbf{kipörögnek} vb. \emph{kipörög} `to spin' + 3rd
  pers. pl. ind. pres. indef. \emph{-nek} \\
\textbf{kirakják} vb. pref. \emph{ki-} `outwardly' + vb. \emph{rak}
`to place, to put' $\rightarrow$ `to take out' + 3rd
 pers. pl. ind. pres. def. See \emph{raktár} \\
\textbf{kis} adj. `small' \\
\textbf{kisgyerek} \emph{kis} `small' + \emph{gyerek} `child' \\
\textbf{kiszáradt} vb. \emph{kiszárad} `to dry (up)' +
past. part. \emph{-t} \\
\textbf{kiterített} vb. pref. \emph{ki-} `outwardly' +
vb. \emph{terít} `to spread out, to lay out, to set (a table for a
meal)' + 3rd pers. sg. ind. past indef. \textbar\textbar{} past part. \\
\textbf{kiülnek} vb. pref. \emph{ki-} `outwardly' + vb. \emph{ül}
`to sit' + 3rd pers. pl. ind. pres. indef. \\
\textbf{kiülnék} vb. pref. \emph{ki-} `outwardly' + vb. \emph{ül}
`to sit' + 1st pers. sg. cond. indef. \\
\textbf{kiváltam} vb. pref. \emph{ki-} `out[wardly]' + vb. inf. \emph{válik} `to
become' $\rightarrow$ `to leave, to stand out, to excel, to separate'
+ 1st pers. sg. ind. past. indef. \\
\textbf{kivert} vb. \emph{kiver} `to beat up' + 3rd
pers. sg. ind. past indef. \\
\textbf{kivilágított} vb. pref. \emph{ki-} `outwardly' + \emph{világ}
`light (arch.), world' $\rightarrow$ arch. `illumination' +
    caus. vb.-forming \emph{-ít} `to make smth ...-like' $\rightarrow$
`to illuminate' + past part. \emph{-ott} \\
\textbf{kín} lit. `suffering' \\
\textbf{kívánnak} vb. \emph{kíván} `to wish' + 3rd
pers. pl. ind. pres. indef. \emph{-nak} \\
\textbf{kívül} adv. `outdoor, outside' \\
\textbf{kocska} `a die' \\
\textbf{komoly} adj. `severe, austere' \\
\textbf{konnadt} syn. \emph{kókadt} `bent downwards', fig. `tired' \\
\textbf{konokon} adj. \emph{konok} `stubborn' + super. \emph{-on}
`on (that manner)' \\
\textbf{korallszirtje} \emph{korall} `coral' + \emph{szirt} `reef' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-je} \\
\textbf{koromban} \emph{korom} `soot' + iness. \emph{-ban}
`in' \\
\textbf{költemény} `poem' \\
\textbf{kölykét} \emph{kölyöl} `cub, young animal, kid (human)'
+ 3rd pers. sg. single\hyp{}poss. \emph{-e} + acc. \emph{-t} \\
\textbf{kölyök} dial. `young, cub' (animals) \textbar\textbar{}
colloq. `brat, kid' \\
\textbf{könny} `teardrop' \\
\textbf{könnyes} \emph{könny} `teardrop' +
adj.\hyp{}forming \emph{-es} $\rightarrow$ `tearful' \\
\textbf{könnyebb} adj. \emph{könnyű} `easy, light' + comp. \emph{-ebb}
$\rightarrow$ `easier' \\
\textbf{könnyen} \emph{könnyű} `easy, light' + super. \emph{-en} `on
(that manner)'
$\rightarrow$ `effortlessly, easily' \\
\textbf{könnyezik} \emph{könny} `teardrop' +
vb.\hyp{}forming \emph{-ezik} $\rightarrow$ `to cry, to water (eyes)' \\
\textbf{könnyű} \emph{könnyű} `simple, easy, light' \\
\textbf{könnyűség} \emph{könnyű} `simple, easy, light' +
  abst. noun\hyp{}forming \emph{-ség} \\
\textbf{könyöklök} \emph{könyök} `elbow' + vb.-forming \emph{-öl}
$\rightarrow$ `to rest on one's elbow' + 1st
pers. sg. ind. pers. \emph{-ök} \\
\textbf{könyvet} \emph{könyv} `book' + acc. \emph{-et}
(low\hyp{}vowel word) \\
\textbf{körém} \emph{kör} `circle' + lative \emph{-é} `towards' + 1st pers. sg. \emph{-m} \\
\textbf{körül} \emph{kör} `circle' + obs. ablat. \emph{-l} `from'
$\rightarrow$ post. `around, near (space)' \textbar\textbar{}
vb. pref. \emph{körül-} \\
\textbf{kötélen} vb. \emph{köt} `to bind' +
noun\hyp{}forming \emph{-él} $\rightarrow$ `rope' + super. \emph{-en}
`on (that manner)' \\
\textbf{kötéltáncos} vb. \emph{köt} `to bind' +
noun\hyp{}forming \emph{-él} $\rightarrow$ `rope' + \emph{tánc}
`dance' + adj.\hyp{}forming \emph{-os} $\rightarrow$ `acrobat' \\
\textbf{kövek} \emph{kő} `stone' + pl. \emph{-ek} \\
\textbf{közben} \emph{köz} `interval, community' + iness. \emph{-ben}
`in' \\
\textbf{közt} post. `between, amid' from \emph{között}: \emph{köz}
  `interval' + locat. \emph{-ött} \textbar\textbar{} \emph{köz}
  `space between' + acc. \emph{-t} \\
\textbf{között} post. `between, amid', \emph{köz} `interval' +
locat. \emph{-ött} \\
\textbf{közt} post. `between, amid' from \emph{között}: \emph{köz}
  `interval' + locat. \emph{-ött} \\
\textbf{köztük} \emph{közt} `between' + 3rd
pers. sg. single\hyp{}poss. \emph{-e} + 3rd pers. pl. $\rightarrow$
pron. `between them' \\
\textbf{közvagyon} \emph{köz} `public, community' \textbar\textbar{}
`interval' + \emph{vagyon} `propert, wealth' $\rightarrow$ `public
property' \\
\textbf{kő} `stone' \\
\textbf{kutya} `dog' \\
\textbf{különös} `strange' \\
\textbf{küszöbömig} \emph{küszöb} `threshold, doorstep' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-öm} `my' + term. `up to,
until' \emph{-ig} \\
\textbf{küszöbön} \emph{küszöb} `threshold, doorstep' +
super. \emph{-on} `on' \\
%%
\ \\
\noindent \textbf{\large L} \\
\textbf{lakom} vb. \emph{lak} `to dwell', inf. \emph{lakik}, 1st
pers. sg. ind. pres. \emph{-om} \\
\textbf{langy} arch. poetic \emph{langyos} `tepid' \\
\textbf{lapultam} vb. \emph{lapul} `to lie low' + 1st pers. sg. past
indef. \emph{-tam} \\
\textbf{lassan} \emph{lassú} `slow' + adv.\hyp{}forming \emph{-an}
$\rightarrow$ `slowly' \\
\textbf{lassudan} adv. `slowly'. See \emph{lassan} \\
\textbf{lábod} \emph{láb} `leg (human, animal), foot' + 2nd
pers. inform. single\hyp{}poss. poss. \emph{-od} \\
\textbf{lágy} adj. `soft, weak, gentle' \\
\textbf{lámpa} `lamp' \\
\textbf{lángra} \emph{láng} `flame, flare' + subl. \emph{-ra}
`onto' \\
\textbf{lány} `girl' \\
\textbf{lássam} vb. \emph{lát} `to see' + 1st
pers. sg. subj. pres. indef. \\
\textbf{látlak} vb. \emph{lát} `to see' + 1st pers. sg. ind. pres. 2nd obj
\textbf{látom} vb. \emph{lát} `to see' + 1st
pers. sg. ind. pres. def. \\
\textbf{látni} vb. \emph{lát} `to see' + inf. \emph{-ni} \\
\textbf{látszik} vb. \emph{lát} `to see' + vb.-forming \emph{-szik}
$\rightarrow$ `to be visible, to appear' \\
\textbf{láttam} vb. \emph{lát} `to see' + short base \emph{-t}, 1st
pers. sg. past def. \emph{-tam} \\
\textbf{lázad} vb. \emph{to rebel}, 3rd pers. sg. ind. pres. indef. \\
\textbf{lázat} \emph{láz} `fever' + acc. \emph{-at} \\
\textbf{le} comp. `down' \\
\textbf{leány} lit. \emph{lány} `girl' \\
\textbf{lebeg} vb. \emph{lebeg} `to float (in the air)' + 3rd
  pers. sg. ind. pres. indef. \\
\textbf{ledőlt} vb. \emph{ledől} `to fall down' + 3rd
pers. sg. ind. past indef. \emph{-t} \\
\textbf{legmagosabb} super. pref. \emph{leg-} `most' +
adj. \emph{magas} `tall' + comp. \emph{-abb} $\rightarrow$
`tallest' \\
\textbf{legyek} vb. \emph{van} `to be' \textbar\textbar{}
vb. \emph{lesz} `to become' + 1st
pers. sg. subj. indef. \emph{-jek} \textbar\textbar{} \emph{légy} `fly
(insect)' + pl. \emph{-ek} \\
\textbf{lehet} pot. vb. \emph{van, lesz} `may be, may become', 3rd
pers. sg. ind. \\
\textbf{leheletedbe} vb. \emph{lehel} `exhale' +
noun\hyp{}forming \emph{-et} + 2nd pers. sg. single
poss. \emph{-ed} `your' + illative \emph{-be}
`into (the inside of)' \\
\textbf{leheletével} vb. \emph{lehel} `exhale' +
noun\hyp{}forming \emph{-et} + poss. \emph{-é} + instr. \emph{-vel} `with' \\
\textbf{leheverünk} vb. pref. \emph{le-} completion, downward +
vb. \emph{hever} vb. `to lie (down)' + 1st
pers. pl. ind. pres. indef. \emph{-ünk} \\
\textbf{lehullt} vb. rare \emph{lehull} `it falls (impersonal)' + past
part. \emph{-t} \\
\textbf{lehunyhatnám} vb. \emph{lehuny} `to close (one's eyes)' + pot. 1st pers. sg. cond. def. \emph{-hatnám} \\
\textbf{lehunyod} vb. pref. \emph{le-} completion, downward +
vb. \emph{huny} `to close (one's eyes)' + 2nd
pers. sg. ind. pres. def. \\
\textbf{lejtése} vb. \emph{lejt} `to slope'+
noun\hyp{}forming \emph{-és} + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-e} \\
\textbf{lelked} \emph{lélek} `soul' + 2nd
  pers. sg. single\hyp{}poss. poss. \emph{-ed}
  $\rightarrow$ \emph{lelked} `your soul' \\
\textbf{lelkedet} \emph{lélek} `soul' + 2nd
  pers. sg. single\hyp{}poss. poss. \emph{-ed} $\rightarrow$ \emph{lelked} + acc. \emph{-et} \\
\textbf{lelkek} \emph{lélek} `soul' + pl. \emph{-ek} $\rightarrow$ \emph{lelkek} \\
poss. \emph{-ed} $\rightarrow$ \emph{lelked} + acc. \emph{-et} \\
\textbf{lelkem} \emph{lélek} `soul' + 1st pers. sg. single\hyp{}poss. \emph{-em} \\
\textbf{lelkiismeret} arch. dial. \emph{lelki} `his, her soul'
(alt. \emph{lelke}) + arch. dial. \emph{ismereti} `his, her, its
knowledge' (alt. \emph{ismerete}) $\rightarrow$ `(moral)
conscience' \\
\textbf{lelt} vb. \emph{lel} `to find' + 3rd pers. sg. ind. past indef. \\
\textbf{leltár} `inventory' \\
\textbf{leltáromat} \emph{leltár} `inventory' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-om} + acc. \emph{-at} \\
\textbf{lengedező} vb. \emph{leng} `to swing, sway' +
freq. vb.\hyp{}forming \emph{-edezik} (unrounded front\hyp{}vowel) +
pres. part. \emph{-ő} $\rightarrow$ `swinging, swaying' \\
\textbf{lengenek} arch. \emph{lengnek}, vb. \emph{leng} `to swing,
sway, oscillate' + 3rd pers. pl. ind. pres. indef. \\
\textbf{lenézel} vb. `to look down'+ 2nd
pers. sg. ind. pres. indef. \emph{-el} \\
\textbf{lenne} vb. \emph{van} `to be' \textbar\textbar{}
vb. \emph{lesz} `to become' + 3rd pers. sg. cond. pres. \\
\textbf{lennél} vb. \emph{van} `to be' \textbar\textbar{}
vb. \emph{lesz} `to become' + 2nd pers. sg. cond. pres. \\
\textbf{lepkék} \emph{lepke} `butterfly' + pl. \emph{-k}
$\rightarrow$ \emph{lepkék} \\
\textbf{lesikáltam} vb. pref. \emph{le-} completion, downward +
vb. \emph{sikál} `to scrub, scour (out)' + 1st pers. sg. ind. past. def.
\emph{-am} \\
\textbf{lestem} vb. \emph{les} `to spy' + 1st pers. sg. past
indef. \emph{-tem} \\
\textbf{lesz} vb. \emph{van} `to be', 3rd
pers. sg. ind. fut. indef. \\
\textbf{leszakítja} vb. \emph{leszakít} syn. \emph{leszed} `to pluck (a flower)' + 3rd
pers. sg. ind. pres. def. \emph{-ja}  \\
\textbf{leszedjük} vb. pref. \emph{le-} completion + vb. \emph{szed}
`to take from somewhere, to collect, to take medicine regularly'
$\rightarrow$ `to pick smth off smth, smb' + 1st
pers. pl. ind. pres. def. \textbar\textbar{} 1st
pers. pl. subj. def. \emph{-jük} \\
\textbf{lett} vb. \emph{lesz} `to become' + 3rd pers. sg. ind. past. indef. \\
\textbf{leülsz} vb. pref. \emph{le-} completion, downward + vb. \emph{ül}
`to sit' $\rightarrow$ `to sit down' \\
\textbf{levegőben} \emph{levegő} `air' + iness. \emph{-ben} `in' \\
\textbf{levelek} \emph{level} `leaf' + pl. \emph{-ek} \\
\textbf{leveles} \emph{level} `leaf' + adj.\hyp{}forming \emph{-es}
$\rightarrow$ `leafy' \\
\textbf{levesse} vb. \emph{levet} `to buck, throw off' + 3rd
pers. sg. subj. def. \\
\textbf{léha} `improvident, idler' \\
\textbf{lépked} vb. \emph{lép} `to walk' +
freq. vb.\hyp{}forming \emph{-ked} $\rightarrow$ `to step, to pace' +
3rd pers. sg. ind. pres. indef. \\
\textbf{libbenjetek} vb. \emph{libben} `to flutter' + 2nd
pers. pl. informal imper. \emph{-jetek} \\
\textbf{libegtetett} vb. \emph{libeg} `to float' \textbar\textbar{}
lit. `to sway' + caus. \emph{-tet} `to let' + past
part. \emph{-ett} \\
\textbf{lobbant} vb. \emph{lobban} `to flare up, to burst into', `to
fall (in love)' + 3rd pers. sb. ind. past indef. \textbar\textbar{} past
part. \\
\textbf{locska} adj. `blethering' \\
\textbf{lombok} \emph{lomb} `leaf' + pl. \emph{-ok} \\
\textbf{lombosodnak} \emph{lomb} `foliage, leaves' + \emph{-os}
adj.\hyp{}forming + vb.-forming \emph{-odik} `to become smth' (here, `foliage') + 3rd
pers. ind. pres. indef. $\rightarrow$ `they start to become leaves' \\
\textbf{lopcska} `chuck' (shoulder meat cut, probably pork) \\
\textbf{lopott} vb. \emph{lop} `to steal' + past part. \emph{-ott}
$\rightarrow$ `stolen' \\
\textbf{löknek} vb. \emph{lök} `to push' + 3rd pers. pl. ind. pres. indef. \\
%%
\ \\
\noindent \textbf{\large M} \\
\textbf{ma} `today' \\
\textbf{macska} `cat' \\
\textbf{madarat} \emph{madár} `bird' + acc. \emph{-t} \\
\textbf{madár} `bird' \\
\textbf{magadban} \emph{maga} `oneself' + 2nd pers. sg. single
  poss. \emph{-d} + iness. \emph{-ban} `in' \\
\textbf{magadnak} \emph{maga} `oneself' + 2nd pers. sg. single
  poss. \emph{-d} + dat. \emph{-nak} \\
\textbf{magadra} \emph{maga} `oneself, itself' + subl. \emph{-ra}
`onto' \\
\textbf{magamat} \emph{magam} `myself' + acc. \emph{-at} \\
\textbf{magamba} \emph{maga} `oneself' + 1st pers. sg. \emph{-m} +
illative \emph{-ba} `into (the inside of)' \\
\textbf{magamban} \emph{maga} `oneself' + 1st pers. sg. \emph{-m} +
1st pers. sg. pers. \emph{-m} $\rightarrow$ `myself' +
iness. \emph{-ban} `in' \\
\textbf{magányos} adv., adj. \emph{magán} `separate[ly]' +
adj.\hyp{}forming \emph{-os} $\rightarrow$ `lonely (psych.)' \\
\textbf{magának} pron. refl. 3rd pers. `himself' \\
\textbf{magokat} \emph{mag} `seed' + pl. \emph{-ok} +
acc. \emph{-at} \\
\textbf{maguknak} \emph{mag(a)} `oneself' + 3rd pers. pl. \emph{-uk} +
dat. \emph{-nak} \\
\textbf{magyarázat} vb. \emph{magyaráz} `to explain' +
noun\hyp{}forming \emph{-at} \\
\textbf{majd} adv. `sometime, later' \\
\textbf{majdnem} adv. \emph{majd} `sometime, later', lit. `and then,
afterwards' + \emph{nem} `not'
$\rightarrow$ `almost' \\
\textbf{marad} `to stay, remain (also state of
being)' + 3rd pers. sg. ind. pres. indef. \\
\textbf{maradnál} vb. \emph{marad} `to stay, remain (also state of
being)' + 2nd pers. sg. cond. pres. indef. \\
\textbf{Marosra} prop. noun. \emph{Maros}, a Hungarian and Romanian
river + subl. \emph{-ra} `onto' \\
\textbf{martjára} lit. \emph{mart} `embankment' + 3rd
pers. sg. poss. single\hyp{}poss. \emph{-ja} + subl. \emph{-ra}
`on[to]' \\
\textbf{már} adv. `already, yet', `anymore, no longer (in the
negative)' \\
\textbf{más} adj. `other, different' \textbar\textbar{}
pron. `smb/smth else' \\
\textbf{másfél} adj. \emph{más} `other, different' + \emph{fél} `half'
$\rightarrow$ `one and a half' \\
\textbf{másikát} adj. \emph{más} `other' +
noun\hyp{}forming \emph{-ik} $\rightarrow$ `another' + 3rd
  pers. sg. single\hyp{}poss. \emph{-a} + acc. \emph{-t}
  $\rightarrow$ \emph{-át} \\
\textbf{mást} adj. \emph{más} `other, different' + acc. \emph{-t} \\
\textbf{mázsa} `quintal' \\
\textbf{meg} arch. emph. \textbar\textbar{} conj. `and, plus' \\
\textbf{megálltam} perfect. vb. pref. \emph{meg-} + vb. \emph{áll} `to
  stand' + 1sr pers. sg. ind. past \\
\textbf{megáradt} perfect. vb. pref. \emph{meg-} + vb. \emph{árad} `to
flood' + past part. \emph{-t} \\
\textbf{megcsókolni} perfect. vb. pref. \emph{meg-} +
vb. \emph{csókol} `to kiss' + inf. \emph{-ni} \\
\textbf{megelfárad} emph. vb. pref. \emph{meg-} + vb. \emph{fárad} `to
tire, to be tired' + 3rd
pers. sg. ind. pres. indef. \textbar\textbar{} past part. \\
\textbf{megfog} emph. vb. pref. \emph{meg-} + vb. \emph{fog} `to hold'
+ 1st pers. sg. ind. pres. indef. \\
\textbf{megfoltozom} perfect. vb. pref. \emph{meg-} +
vb. \emph{foltoz} `to mend (by sewing), to patch' + 1st
pers. sg. ind. pres. def. \\
\textbf{meghajlik} perfect. vb. pref. \emph{meg-} + vb. \emph{hajlik}
`to bend, to curve' + 3rd pers. sg. ind. pres. indef. \\
\textbf{meghalni} perfect. vb. pref. \emph{meg-} + \emph{halni}
vb. inf. \emph{hal} `to die' \\
\textbf{meghasadt} perfect. vb. pref. \emph{meg-} + vb. \emph{hasad}
`to crack, split apart' + past part. \emph{-t} \\
\textbf{meghallgattál} perfect. vb. pref. \emph{meg-}
+ vb. \emph{hallgat} vb. `to remain silent' + 2nd pers. sg. ind. past
indef. \\
\textbf{megint} \emph{meg} `again' (short action) +
adv.\hyp{}forming \emph{-int} \\
\textbf{megjelent} vb. inf. \emph{megjelenik} `to appear' + 2nd
pers. sg. inform. (or 3rd pers. sg.) ind. past. indef. \\
\textbf{meglátnám} perfect. vb. pref. \emph{meg-} + vb. \emph{lát}
`to see' + 1st pers. sg. cond. pres. def. \emph{-nám} \\
\textbf{megláttalak} perfect. vb. pref. \emph{meg-} + vb. \emph{lát}
`to see' + 1st pers. sg. ind. past 2nd obj \\
\textbf{megleled} perfect. vb. pref. \emph{meg-} + vb. \emph{lel} `to
find' + 2nd pers. sg. inform. def. \emph{-ed} \\
\textbf{meglett} obs. `adult' \\
\textbf{megnézzük} vb \emph{megnéz} `to see, take a look' + 1st
pers. pl. ind. pres. indef. \\
\textbf{megnő} vb. `to grow' + 3rd pers. sg. ind. pres. indef. \\
\textbf{megöregszel} perfect. vb. pref. \emph{meg-} +
\emph{öreg} `old' + vb.\hyp{}forming \emph{-szik}
$\rightarrow$ `to grow, become old' + 2nd
pers. sg. inform. pres. indef. \\
\textbf{megtört} perfect. vb. pref. \emph{meg-} + vb. \emph{tör} `to
    break, crush' + past. part. \emph{-t} \\
\textbf{megy} vb. `to go, travel, pass by (of time)' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{meleg} adj. `warm, hot, heartfelt, cordial' \\
\textbf{melled} \emph{mell} `breast' + 2nd pers. sg. single
poss. \emph{-ed} \\
\textbf{mellett} \emph{mell} `breast' + locat. \emph{-ett}
$\rightarrow$ `next to' \\
\textbf{mellé} post. `next to' (movement implied, compare
with locative \emph{mellett}) \\
\textbf{melléd} post. \emph{mellé} `next to' + 2nd pers. sg. single
  poss. \emph{-d} $\rightarrow$ pron. `next yo you' \\
\textbf{mely} lit. `which, that' \\
\textbf{melybe} lit. arch. \emph{mely} `which' + illative \emph{-be}
`into (the inside of)' \\
\textbf{melyben} lit. \emph{mely} `which, that' + iness. \emph{-ben}
`in' \\
\textbf{menekülnék} vb. \emph{menekül} `to flee, to run away, to
escape' + 1st pers. sg. cond. indef. \\
\textbf{merek} vb. \emph{} `to dare' + 1st pers. sg. ind. pres. indef \\
\textbf{mert} conj. `because' \\
\textbf{mező} `field (coutryside)' \\
\textbf{még} adv. `still, moreover, later, sometime' \\
\textbf{mégis} adv. \emph{még} `still, moreover, later, sometime'
+ \emph{is} `also' $\rightarrow$ `however, still, yet, but,
nevertheless, after all' \\
\textbf{méla} adj. `dreamy, pensive' \\
\textbf{mély} adj. `deep' \\
\textbf{mélyein} adj. \emph{mély} `deep' + poss. pl. \emph{-(e)i} +
iness. \emph{-n} `in' \\
\textbf{mérte} vb. \emph{mér} `to measure (time, length
etc.)' \textbar\textbar{} `to strike' + 3rd pers. sg. ind. past def. \\
\textbf{mi} `what' \\
\textbf{miért} \emph{mi} `what' + causal\hyp{}final \emph{ért} `for'
$\rightarrow$ `why' \\
\textbf{mik} \emph{mi} `what' + pl. \emph{-k} \\
\textbf{miként} \emph{mi} `what' + \emph{-kent} $\rightarrow$
essive\hyp{}formal sg. `as a, like a' (task, role,
manner) \textbar\textbar{} syn. \emph{hogyan} form. adv. `how?, in what way?' \textbar\textbar{} syn. \emph{ahogyan} form. adv. `as, like' \\
\textbf{mikor} adv. `when' \\
\textbf{míg} lit. adv. `as long as, while (of
time)' \textbar\textbar{} `until' \\
\textbf{mily} poetic `how ...(!)' \\
\textbf{milyen} pron. `what', `how + adj. or adv.' \\
\textbf{mind} pron. sg. `all, each of' \\
\textbf{mindegy} \emph{mind} + \emph{egy} $\rightarrow$ `it matters not
+ embedded question', adj. `irrelevant' \\
\textbf{minden} pron. det. `every' \\
\textbf{mindenség} pron. det. \emph{minden} `every' \emph{-ség}
`state of' $\rightarrow$ `universe, world' \\
\textbf{mindenre} pron. det. \emph{minden} `every' + subl. \emph{-re}
`onto' \\
\textbf{mindenfélét} pron. det. \emph{minden} `every' + \emph{-féle}
`kind, type' + acc. \emph{-t} $\rightarrow$ `all kinds of things' \\
\textbf{mindent} pron. det. \emph{minden} `every' + acc. \emph{-t}
$\rightarrow$ `everything' \\
\textbf{mindenik} folk. `each, every one'; alt. \emph{mindegyik}
  $leftarrow$ \emph{mind} `all, every' + \emph{egyik} `one [of]'. \\
\textbf{mindig} \emph{mind} `always' (repetition) + term. `until'
  \emph{-ig} \\
\textbf{minek} \emph{mi} `what' + dat. \emph{-nek} $\rightarrow$ `for what' \\
\textbf{mint} conj. `like' \textbar\textbar{} comp. different degrees
`than' \\
\textbf{mintha} \emph{mint} `like, as' + \emph{ha} `if' \\
\textbf{mit} interj. \emph{mi} `what' + acc. \emph{-t} \\
\textbf{mivel} \emph{mi} `what' + instr. \emph{-vel}
`with' \textbar\textbar{} conj. `as (because)' \\
\textbf{mondanám} vb. \emph{mond} `to say, tell' + 1st
pers. sg. cond. pres. def. \\
\textbf{mondd} vb. \emph{mond} `to say,
tell, to forecast' + 2nd pers. sg. inform. subj. def. \\
\textbf{mondják} vb. \emph{mond} `to say, tell' + 3rd
pers. pl. ind. pres. def. \emph{-ják} \\
\textbf{mondottál} arch. \emph{mondtál}, vb. \emph{mond} `to say,
tell' + 2nd pers. sg. ind. past indef. \\
\textbf{most} adv. `now' \\
\textbf{mosolygása} \emph{mosolygás} `smile' + 3rd pers. sg. single
poss. \emph{-a} \\
\textbf{mosolyog} vb. `to smile' + 3rd pers. sg. ind. pres. indef. \\
\textbf{motyogod} vb. \emph{motyog} `to mumble, mutter' + 2nd
pers. sg. ind. pres. def. \emph{-od} \\
\textbf{mozdulnak} vb. \emph{mozdul} `to move, budge' + 3rd
pers. pl. ind. pres. indef. \\
\textbf{mögött} \emph{mög} `space behind smth' +  locat. \emph{-ött}
$\rightarrow$ `behind' \\
\textbf{mult} \emph{múlt} rare, past part. `past',
inf. \emph{múlik} \\
\textbf{multból} \emph{múlt} rare, past part. `past' +
elative \emph{-ból} `from, out of' \\
\textbf{munkások} \emph{munka} `work' + noun\hyp{} and
adj.\hyp{}forming \emph{-s} $\rightarrow$ \emph{munkás} `labourer,
worker' + pl. \emph{-ok} \\
\textbf{munkátlanul} \emph{munka} `work' + privative
adj.\hyp{}forming \emph{-tlan} `-less' $\rightarrow$ `idle,
unemployed' + adv.\hyp{}forming \emph{-ul} \\
\textbf{muszáj} noun (not a verb!) `must, have to, be obliged to'
(often followed by an infinitive) \\
\textbf{muzsikás} dialect \emph{muzsik} `music' +  agent
noun\hyp{}forming \emph{-ás} $\rightarrow$
poet. adj. `musical' \textbar\textbar{} `musician' \\
%%
\ \\
\noindent \textbf{\large N} \\
\textbf{nagy} adj. `great' \\
\textbf{nagyon} adj. \emph{nagy} `great' + adv.\hyp{}forming \emph{-on}
    $\rightarrow$ `very (to a high degree)' \\
\textbf{nap} \emph{nap} `sun, day' \\
\textbf{nappal} \emph{nap} `day' + instr. \emph{-val} `with'
$\rightarrow$ \emph{nappal} \\
\textbf{nappalok} \emph{nap} `day' + instr. \emph{-val}
  `with' $\rightarrow$ \emph{nappal} + pl. \emph{-ok} \\
\textbf{napszámosokhoz} \emph{napszámos} `day labourer' +
pl. \emph{-ok} + allat. `\emph{-hoz} `towards' \\
\textbf{napszülte} \emph{nap} `sun/day' + vb. \emph{szül} `to give
birth, bear' + 3rd pers. sg. ind. past def. \emph{-te} \\
\textbf{napvilágra} \emph{nap} `sun/day' + \emph{világ}
  `world' + subl. \emph{-ra} `onto' \\
\textbf{nálad} pron. \emph{nála} `at his place/house, on him' + 2nd
pers. sg. $\rightarrow$ `at your place' \\
\textbf{ne} \emph{nem} becomes \emph{ne} in imperative, subjunctive \\
\textbf{nehéz} adj. `difficult' \textbar\textbar{} `heavy' \\
\textbf{neked} 2nd pers. sg. of \emph{neki} `to/for him'
$\rightarrow$ `to/for you' \\
\textbf{neveddel} \emph{név} `name' + 2nd
pers. poss. single\hyp{}poss. \emph{-ed} + instr. \emph{-del}
`with' \\
\textbf{nevetni} vb. \emph{nevet} `to laugh', inf. \emph{nevetni} \\
\textbf{néha} adv. `sometimes' \\
\textbf{néhány} pron. `some (of us, you, them)' \\
\textbf{nékünk} lit. \emph{néki} (alt. \emph{neki}) `to/for him' +
1st pers. pl. $\rightarrow$ `to us, for us' \\
\textbf{nélkül} post. `without' \\
\textbf{nélkűletek} pron. \emph{nélküle} `without it' + 2nd
pers. pl. $\rightarrow$ `without you (pl.)' \\
\textbf{néma} `mute' \\
\textbf{némán} \emph{néma} `mute' +
deadj. adv.\hyp{}forming \emph{-an} $\rightarrow$ `mutely, silently' \\
\textbf{némaság} \emph{néma} `mute' + noun\hyp{}forming \emph{-ság}
(state of being) $\rightarrow$ `mutism' \\
\textbf{nézd} vb. \emph{nez} `to look at' + 2nd
pers. sg. subj. indef. \\
\textbf{néznek} vb. \emph{néznek} `to look at smth' + 3rd
  pers. pl. ind. pres. indef. \emph{-nek} \\
\textbf{néztél} vb. \emph{nez} `to look at' + 2nd
pers. sg. ind. past. indef. \emph{-tél} \\
\textbf{nézzük} vb. \emph{nez} `to look at' + 1st
pers. pl. ind. pres. \emph{-jük} \\
\textbf{no} interj. `well, now' \\
\textbf{növesztett} vb. \emph{növeszt} `to make smth grow' + 3rd
pers. sg. ind. past indef. \\
\textbf{nő} `woman' \\
%%
\ \\
\noindent \textbf{\large Ny} \\
\textbf{nyafog} vb. `to whine' + 3rd pers. sg. ind. pres. indef. \\
\textbf{nyakamba} \emph{nyak} `neck' + 1st
pers. sg. poss. single\hyp{}poss. \emph{-am} `my' +
illative \emph{-ba} `into (the inside of)' \\
\textbf{nyár} `summer' \\
\textbf{nyári} \emph{nyár} `summer' + adj.\hyp{}forming \emph{-i} \\
\textbf{nyáron} \emph{nyár} `summer' + super. \emph{-on} `on' \\
\textbf{nyers} adj. `raw' \\
\textbf{nyírfa} `birch' \\
\textbf{nyirkos} adj. `humid, damp' \\
\textbf{nyitott} vb. \emph{nyit} `to open' + 3rd pers. sg. ind. past
indef \textbar\textbar{} past part. \emph{-ott} \\
\textbf{nyomja} vb. \emph{nyom} `to weigh' + 3rd
pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{nyugalmat} \emph{nyugalom} `serenity, peacefulness' +
acc. \emph{-at} \\
\textbf{nyugalom} `serenity, peacefulness' \\
\textbf{nyugalomban} `serenity, peacefulness' + iness. \emph{-ban}
`in' \\
\textbf{nyugodt} adj. `tranquil, calm'\\
\textbf{nyugton} arch. adv. `calmly, in peace' \\
\textbf{nyujtózik} vb. \emph{nyujt} `to stretch out, extend, prolong' + freq.\hyp{}reflex. \emph{-ózik} \\
\textbf{nyüzsögő} vb. \emph{nyüzsög} `to bustle' +
pres. part \emph{-ő} \\
%%
\ \\
\noindent \textbf{\large O} \\
\textbf{odáat} \emph{odá} `there (to that place)' + acc. \emph{-at} \\
\textbf{odahallik} vb. pref. (adv.) \emph{oda-} `there' +
arch. vb. \emph{hallik} syn. \emph{hallatszik} `to be able to be
heard, to sound, to be audible' \\
\textbf{odasujtsz} vb. pref. (adv.) \emph{oda-} `there' +
vb. \emph{sújt} `to strike (smb/smth) with smth' + 2nd
pers. sg. imper. \\
\textbf{okos} \emph{ok} `reason, motive' + ornative \emph{-os} `smth
with' $\rightarrow$ `smart, clever, bright' \\
\textbf{oly} arch. pron. \emph{oly} `so' \\
\textbf{olyan} arch. pron. \emph{oly} `such, so' + deadj. adv.\hyp{}forming
\emph{-an} \\
\textbf{olyanigen} adj. \emph{olyan} `such, so' + \emph{igen} `yes'
$\rightarrow$ emphatic affirmation, `yes indeed, exactly so' \\
\textbf{olvassa} vb. \emph{olvas} `to read' + 2nd
pers. sg. inform. (or 3rd pers. sg.) subj. indef. \\
\textbf{omlás} vb. \emph{omlik} `to crumble, collapse, fall down' +
agent noun\hyp{}forming \emph{-ás} \\
\textbf{omlott} vb. \emph{omlik} `to crumble, collapse, fall down' +
3rd pers. sg. ind. past indef. \emph{-ott} \\
\textbf{ondoltam} vb. \emph{gondol} `to think, plan, guess, worry' +
1st pers. sg. ind. past \emph{-tam} \\
\textbf{országos} \emph{ország} `country' + + adj.\hyp{}forming \emph{-os}
$\rightarrow$ `national, public' \\
\textbf{orvostól} \emph{orvos} `physician' + ablat. \emph{-tól}
`without, from' \\
\textbf{ostoba} adj. `stupid, foolish' \\
\textbf{ott} demonst. `there, over there' \\
\textbf{otthontalan} \emph{otthon} `home' + privative
adj.\hyp{}forming \emph{-talan} $\rightarrow$ `homeless' \\
%%
\ \\
\noindent \textbf{\large Ó} \\
\textbf{óh} lit. arch. interj. \emph{oh} \\
\textbf{ólomkatonák} \emph{ólom} `lead' + \emph{katona} `soldier' +
pl. \emph{-k} \\
\textbf{óriás} adj. `giant, huge, vast' \\
%%
\ \\
\noindent \textbf{\large Ö} \\
\textbf{öledbe} \emph{öl} `(someone's) lap' + 2nd pers. sg. single
poss. \emph{-ed} \\
\textbf{ölelő} vb. \emph{ölel} `to
embrace, encircle' + pres. part. \emph{-ő} \\
\textbf{öleltem} perfect. vb. pref. \emph{meg-} + vb. \emph{ölel} `to
embrace' + 1st pers. ind. past \\
\textbf{önmagamba} \emph{ön-} `self-' + \emph{magam} `myself' +
illative \emph{-ba} `into (the inside of)' \\
\textbf{önnön} arch. emph. poss. adj. by duplication of
pers. pron. \emph{ön} (3rd pers. sg.) $\rightarrow$ `one's own' \\
\textbf{öntudat} `self-awareness' \\
\textbf{öreg} adj. `old' \\
\textbf{örök} adj. `perpetual, eternal' \\
\textbf{örökre} `perpetual, eternal' + subl. \emph{-re} `onto' \\
\textbf{örömöt} \emph{öröm} `joy, pleasure' + acc. \emph{-öt} \\
\textbf{örül} vb. `to rejoice' \\
\textbf{örülj} vb. \emph{örül} `to rejoice' + 2nd
pers. sg. subj. indef. \emph{-j} \\
\textbf{örülnék} vb. `to rejoice' + 1st
pers. sg. cond. pres. indef. \emph{-nék} \\
\textbf{összefogja} \emph{össze} `together' + vb. \emph{fog}
  `to hold' + 3rd pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{összekent} \emph{össze} `together' + \emph{ken} `smear'
$\rightarrow$ \emph{összeken} + 3rd pers. sg. past indef. \emph{-t} \\
\textbf{összetört} \emph{össze} `together' + vb. \emph{tör} `to break'
+ past. part. \emph{-t} \\
\textbf{összetörték} \emph{össze} `together' + vb. \emph{tör} `to
break' + 3rd pers. pl. ind. past def. \\
%%
\ \\
\noindent \textbf{\large Ő} \\
\textbf{ők} pron. `they' \\
\textbf{őket} pron. \emph{ők} `they' + acc. \emph{-et} $\rightarrow$
pron. `them' \\
\textbf{őrizlek} vb. \emph{őriz} `to guard, keep safe' + 1st pers. sg. ind. pres. 2nd obj. \\
\textbf{őrizz} vb. \emph{őriz} `to guard, keep safe' + 2nd
pers. sg. inform. subj. indef. \\
\textbf{őrt} \emph{őr} `guard' + acc. \emph{-t} \\
\textbf{őrzi} vb. \emph{őriz} `to guard' + 3rd
pers. sg. ind. pres. def. \emph{-i} \\
\textbf{őszi} \emph{ősz} `autumn' + adj.\hyp{}forming \emph{-i}
$\rightarrow$ `automnal, of autumn' \\
\textbf{őz} roe deer \\
%%
\ \\
\noindent \textbf{\large P} \\
\textbf{papja} \emph{pap} `priest' + 3rd
pers. sg. sing. single\hyp{}poss. \emph{-ja} \\
\textbf{papot} \emph{pap} `priest' + acc. \emph{-ot} \\
\textbf{paprikák} \emph{paprika} `pepper (powder and vegetable)' +
pl \emph{-k} \\
\textbf{paraszt} sg. `peasant'\\
\textbf{partjára} \emph{part} `bank, shore, coast' + 3rd
pers. sg. poss. single\hyp{}poss. \emph{-ja} + subl. \emph{-ra}
`on[to]' \\
\textbf{partszegélyein} \emph{part} `shore, bank' + \emph{szegély}
`border, margin' + poss. pl. \emph{-(e)i} + iness. \emph{-n} `on' \\
\textbf{pára} `moisture, mist' \\
\textbf{pedig} conj. expressing contrast `whereas, but, although' or
addition `and (by contrast)' \\
\textbf{perceim} \emph{perc} + 1st
pers. sg. poss. mult.\hyp{}poss. \emph{-eim} \\
\textbf{pihen} vb. `to rest' + 3rd pers. sg. ind. pres. indef. \\
\textbf{pihensz} vb. `to rest' + 2nd
pers. sg. inform. ind. pres. indef. \\
\textbf{pihéi} \emph{pihe} `down, fluff' + adj.\hyp{}forming \emph{-i} \\
\textbf{pillanat} `instant, moment, second' \\
\textbf{pillék} \emph{pille} arch. rare `snowflake', `butterfly' \\
\textbf{piros} `red' \\
\textbf{pislog} vb. `to blink' \\
\textbf{poros} \emph{por} `dust' + adj.\hyp{}forming \emph{-os}
$\rightarrow$ `dusty' \\
\textbf{porra} \emph{por} `dust' + subl. \emph{-ra} `onto' \\
\textbf{porszem} \emph{por} `dust' + \emph{szem} `eye, speck' \\
\textbf{pólusa} \emph{pólus} `pole' + 3rd pers. sg. single
poss. \emph{-a} (`his pole, pole of smth') \\
\textbf{puha} adj. `soft, squishy' \\
\textbf{pusztaság} \emph{puszta} `bare, plain (geography)' +
noun\hyp{}forming \emph{-ság} (state of being) $\rightarrow$
`wilderness, desert' \\
\textbf{pusztulásban} \emph{pusztulás} `destruction' +
iness. \emph{-ban} `in' \\
%%
\ \\
\noindent \textbf{\large R} \\
\textbf{rab} `captive' \\
\textbf{rabul} \emph{rab} `captive' + essive
  \emph{-ul} `as a, with the intention of' \\
\textbf{ragyog} vb. `to glitter, to shine, to glisten, to sparkle' \\
\textbf{raját} \emph{raj} `swarm' + acc. \emph{-át} \\
\textbf{rajta} pron. `on it' \\
\textbf{rak} vb. `to put, set' + 3rd pers. sg. ind. pres. indef.\\
\textbf{raksz} vb. \emph{rak} `to set up, to build', 2rd
pers. sg. ind. pres. indef. \emph{-sz} \\
\textbf{raktár} `warehouse, storage room' \\
\textbf{ráadás} `addition, extra' \\
\textbf{ráadásul} \emph{ráadás} `addition, extra' + essive
  \emph{-ul} `as a, with the intention of' \\
\textbf{ráér} vb. `to have (the) time' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{rácsok} \emph{rács} `grid, grille' + pl. \emph{-ok} \\
\textbf{rám} pron. \emph{rá} `upon, on, unto him' + 1st
pers. sg. `on me' \\
\textbf{ránk} pron. \emph{rá} `upon, on, unto him' + 1st
pers. pl. poss. single\hyp{}poss. \emph{-nk} \\
\textbf{rántottam} vb. \emph{ránt} `to pull (suddenly)', `to fry in
breadcrumbs' + 1st pers. sg. ind. past. def. \emph{-ottam} \\
\textbf{rászálltak} pointed action \emph{rá-} `on' + vb. \emph{száll}
`to fly, land, to come down' + 3rd pers. sg. past indef. \emph{-tak} \\
\textbf{ráugrott} pointed action \emph{ra-} `at' + vb. \emph{ugrik}
`to leap/jump' + 3rd pers. sg. ind. past. indef. \emph{-ott}
$\rightarrow$ \emph{ugrott} \\
\textbf{reám} lit. arch. pron. 1st pers. \emph{réa} `upon, on, unto,
of
him' \\
\textbf{reggeli} \emph{reggel} `morning' + adj.\hyp{}forming \emph{-i} \\
\textbf{remeg} vb. `to shake, to quiver' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{remélem} vb. \emph{remél} `to hope' + 1st
pers. sg. ind. pres. def  \\
\textbf{rend} `order' \\
\textbf{rendben} \emph{rend} `order' + inness. \emph{-ben} `in' \\
\textbf{reszketni} vb. inf. `to quiver, to tremble, to flutter' \\
\textbf{rezgő} poetic, arch. vb. \emph{rezg} `to tremble, quiver'
(alt. \emph{rezeg}) + pres. part. \emph{-ő} \\
\textbf{reám} pron. lit. \emph{reá} `upon/on him' , syn. \emph{rá} +
1st pers. sg. \\
\textbf{rikoltoz} vb. \emph{rik} `to screech' + freq. \emph{-oltoz} +
3rd pers. sg. pres. indef. \\
\textbf{ring} vb. `to swing, to (gently) sway' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{ringatóan} vb. \emph{ringat} `to (gently) sway
(boat), to rock (cradle)' + part. pres. \emph{-ó} +
adv.\hyp{}forming \emph{-an} \emph{-an} \\
\textbf{ringatóztak} vb. inf. \emph{ringatódzik} `to sway, to rock' +
3rd pers. pl. ind. past indef. \\
\textbf{ringatózom} vb. \emph{ringat} `to (gently) sway
(boat), to rock (cradle)' + freq.\hyp{}reflex. \emph{-ózik}
$\rightarrow$ inf. \emph{ringatózik} +
1st pers. sg. ind. pres. indef. \emph{-ózom} \\
\textbf{rokonok} \emph{rokon} `relative' + pl. \emph{-ok} \\
\textbf{romlás} `decay, deterioration, fade' \\
\textbf{rossz} adj. `bad' \\
\textbf{rólad} pron. \emph{róla} `about, off her' + 2nd pers. sg. \\
\textbf{rózsatő} `rose bush' \\
\textbf{röffent} \emph{röffen} `to grunt once (of a pig)' \\
\textbf{röppen} vb. `to make a short fly (bird)' \\
%%
\ \\
\noindent \textbf{\large S} \\
\textbf{s} poetic abbrv. \emph{es} `and' (does not count as a
syllable) \\
\textbf{sabadságot} \emph{szabad} `free' + \emph{-ság} (state of
being): `freedom' + acc. \emph{-at} \\
\textbf{saját} pron. `own' \\
\textbf{sárga} `yellow' \\
\textbf{se} alt. \emph{sem}, \emph{sem ..., sem ...}  `neither
  ... nor ...' \\
\textbf{seb} `wound' \\
\textbf{sebed} \emph{seb} `wound' + 2nd pers. sg. single
poss. \emph{-ed} \\
\textbf{senkinek} pron. \emph{senki} `no one, nobody' \\
\textbf{semmi} `nothing' \\
\textbf{semmiben} \emph{semmi} `nothing' + iness. \emph{-ben} `in' \\
\textbf{semmije} \emph{semmi} `nothing' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-je} \\
\textbf{semmiképp} \emph{semmi} `nothing' +
adv.\hyp{}forming \emph{-képp} `in a certain way (-ly), as',
alt. \emph{semmiképpen} $\rightarrow$ `by no means, not at all, in no
way'\\
\textbf{sikáltam} vb. \emph{sikál} `to scrub, scour (out)' + 1st
pers. sg. ind. past. def. \emph{-am} \\
\textbf{sikongva} vb. \emph{sikong} `to scream (in joy), to shriek
(inanimate)' + result of accomplished action \emph{-va} \\
\textbf{sincs} contraction of \emph{sem} (`not either')
and \emph{nincs} (`there is not') $\rightarrow$ defective vb. `there
is no ... either, he is not ... either, not have ... either' (3rd
pers. sg. ind. pres. only) \\
\textbf{siránkozz} vb. \emph{sír} vb. \emph{sír} `to cry, weep, to whine' + imperative \\
\textbf{sírni} vb. \emph{sír} `to cry, weep' + inf. \emph{-ni} \\
\textbf{sodort} vb. \emph{sodor} `to drift' + 3rd pers. sg. ind. past
indef. \\
\textbf{soha} adv. `never' \\
\textbf{sok} adj. `much, many, a lot' \\
\textbf{sokat} \emph{sok} `many' + acc. \emph{-at} \\
\textbf{sokba} \emph{sok} `much, many, a lot' + illative \emph{-ba}
`into' \\
\textbf{sors} `fate' \\
\textbf{sorhoz} \emph{sors} `fate' + allat. `towards' \emph{-hoz} \\
\textbf{sorsom} \emph{sors} `fate' + 1st
pers. sg. single\hyp{}poss. poss. \emph{-om} \\
\textbf{sovány} `slim, thin', fig. `poor' \\
\textbf{sóvárogtam} vb. \emph{sóvárog} `to crave, yearn, long' + 1st
pers. sg. ind. past indef. \\
\textbf{sötétben} adj., subst. \emph{sötét} `dark[ness]' +
iness. sg. \emph{-ben} `in' \\
\textbf{sugár} `ray, beam (of light), jet, stream (of liquid), radius
(geometry)' \\
\textbf{suhint} vb. \emph{suhint} `to swing, to whisk' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{súlyos} \emph{súly} `weight' + adj.\hyp{}forming \emph{-os}
$\rightarrow$ adj. `severe, austere' \\
\textbf{sült} vb. \emph{sül} `to roast' + past part. \emph{-t} \\
\textbf{süt} vb. `to roast, shine' + 3rd pers. sg. ind. pres. indef. \\
%%
\ \\
\noindent \textbf{\large Sz} \\
\textbf{szabad} adj. `free' \\
\textbf{szabadulsz} adj. \emph{szabad} `free' + vb.\hyp{}forming \emph{-ul} +
2nd pers. sg. ind. pres. indef. \emph{-sz} \\
\textbf{szakadt} vb. \emph{szakad} `to pour, to tear (cloth), to crack
up (laughter)' + 3rd pers. sg. ind. past indef. \\
\textbf{szaladsz} vb. \emph{szalad} `to run' + 2nd
pers. sg. ind. pres. indef. \emph{-sz} \\
\textbf{szalló} vb. \emph{száll} `to fly' + pres. part. \emph{-ó} \\
\textbf{szavam} \emph{szó} `voice', `word' + 1st
pers. sg. single\hyp{}poss. \emph{-am} \\
\textbf{szavára} \emph{szó} `voice', `word' + 3rd pers. sg. single\hyp{}poss. \emph{-a} $\rightarrow$ \emph{szava} + subl. \emph{-ra} `by, on[to]'
$\rightarrow$ \emph{szavára} \\
\textbf{szád} \emph{száj} `mouth' + 2nd
pers. sg. single\hyp{}poss. poss. \\
\textbf{szálaiból} \emph{szál} `thread' + 3rd
pers. sg. poss. \emph{-a} + poss. pl. \emph{-i} + elative \emph{-ból}
`from, out of' \\
\textbf{száll} vb. `to fly, to land, to come down', 3rd
pers. sg. ind. pres. \\
\textbf{szálljon} vb. \emph{száll} `to fly, to land, to come down'
+ 3rd pers. sg. subj. indef. \\
\textbf{szállhatnék} vb. \emph{száll} `to fly, to land, to come down'
+ potential \emph{-hat} + 1st pers. sg. cond. pres. indef. \emph{-nék} \\
\textbf{szállnak} vb. \emph{száll} `to fly, land, come down' + 3rd
pers. pl. ind. pres. indef. \\
\textbf{számhoz} \emph{száj} `mouth' + 1st pers. sg. single\hyp{}poss. \emph{-m} $\rightarrow$ \emph{szám} +
allat. `towards' \emph{-hoz} $\rightarrow$ \emph{számhoz} \\
\textbf{számító} vb. \emph{számít} `to compute, to scheme' +
pres. part. \emph{-ó} \\
\textbf{számban} \emph{száj} `mouth' + 1st
pers. sg. single\hyp{}poss. \emph{-m} + iness. \emph{-ban} `in' \\
\textbf{szárnyas} \emph{szárny} `wing' + adj.\hyp{}forming \emph{-as}
$\rightarrow$ `winged' \\
\textbf{százannyit} \emph{száz} `hundred' + \emph{annyit} `that much,
as much' $\rightarrow$ `hundredfold' \\
\textbf{százszor} \emph{száz} `hundred' + \emph{szor} `times` + $\rightarrow$ adv. `hundredfold' \\
\textbf{szebben} adj. \emph{szép} `beautiful' + comp. \emph{-bb} +
super. \emph{-en} `on' $\rightarrow$ adv. `more
beautifully' \textbar\textbar{} \emph{szép} `beautiful' +
comp. \emph{-bb} + super. \emph{-ben} `in' $\rightarrow$ adj. `more
beautiful' \\
\textbf{szegénységemhez} \emph{szegény} `poor' +
abst. noun\hyp{}forming \emph{-ség} + 1st pers. sg. single\hyp{}poss. \emph{-em} +
allat. \emph{-hez} `to' \\
\textbf{szemle} From \emph{szemlél}; `examination, survey,
inspection, review, parade, display' \\
\textbf{szemlélni} \emph{szem} `eye' + vb.\hyp{}forming
freq. \emph{-lél} $\rightarrow$ `to contemplate' \\
\textbf{szelid} adj. `tame (of an animal), gentle' \\
\textbf{szem} `eye, speck' \\
\textbf{szemed} \emph{szem} `eye' + 2nd
pers. sg. single\hyp{}poss. poss. \emph{-ed} (but plural implied
here!) \\
\textbf{szemeim} \emph{szem} `eye' + 1st
pers. sg. poss. mult.\hyp{}poss. \emph{-eim} \\
\textbf{szemem} \emph{szem} `eye' + 1st
pers. poss. single\hyp{}poss. \emph{-em} \\
\textbf{szeméből} \emph{szem} `eye' + 3rd
pers. sg. single\hyp{}poss. \emph{-e} + elative \emph{-ből} `from, out
of' \\
\textbf{szenem} \emph{szen} `coal' + 1st pers. sg. single\hyp{}poss. \emph{-em} \\
\textbf{szenvedés} vb. \emph{szenved} `to suffer' +
noun\hyp{}forming \emph{-és} \\
\textbf{szerelem} `love' \\
\textbf{szerelmemtől} vb. \emph{szeret} `to love' +
noun\hyp{}forming \emph{-elem} $\rightarrow$ \emph{szerelem} `love' +
1st pers. sg. single\hyp{}poss. \emph{-em} + ablat. \emph{-től}
`without, from' \\
\textbf{szerelmes} \emph{szerelem} `love' +
adj.\hyp{}forming \emph{-es} $\rightarrow$ `in love' \\
\textbf{szeress-e} vb. \emph{szeret} `to love' + 2nd pers. sg. subj. +
polar interr. \emph{-e} \\
\textbf{szeret} vb. `to love, like' + 2nd
pers. sg. inform. (or 3rd pers. sg.) ind. pres. indef. \\
\textbf{szereti} vb. \emph{szeret} `to love' + 3rd
pers. sg. ind. pres. indef. \emph{-i} \\
\textbf{szeretlek} vb. `to love, like' 1st pers. sg. ind. pres. 2nd
obj. \\
\textbf{szeretném} vb. \emph{szeret} `to love' + 1st
pers. sg. cond. pres. def. \\
\textbf{szeretnél} vb. \emph{szeret} `to love' + 2nd
pers. sg. inform. ind. pres. indef. \\
\textbf{szeretni} inf. vb. \emph{szeret} `to love' \\
\textbf{szeretnivalómmal} vb. inf. \emph{szeretni} + \emph{-való} `fit
to be, deserving of being' + 1st
pers. sg. poss. single\hyp{}poss. \emph{-m} `my' + instr. \emph{-mal} `with' \\
\textbf{szerettem} vb. \emph{szeret} `to love' + 1st
pers. sg. ind. past \\
\textbf{szeretett} vb. \emph{szeret} `to love' + 3rd
pers. sg. ind. past indef. \textbar\textbar{} past part. \emph{-t} \\
\textbf{szeretsz} vb. \emph{szeret} `to love' + 2nd
pers. sg. ind. pres. indef. \\
\textbf{egyszerűség} \emph{egy} `one' +
adj.\hyp{}forming \emph{-szerű} `-like' + abst. noun\hyp{}forming
  \emph{-ség} $\rightarrow$ `simplicity, plainness' \\
\textbf{széjjel} adv. `scattered' \textbar\textbar{} conj. `acros' \\
\textbf{széken} \emph{szék} `chair' + super. \emph{-en} `on' \\
\textbf{szél} `wind' \\
\textbf{széles} adj. `wide' \\
\textbf{szénre} \emph{szén} `coal' + subl. \emph{-re} `onto' \\
\textbf{szépen} adj. \emph{szép} `beautiful' \\
\textbf{széthull} vb. `to fall apart', 3rd
pers. sg. ind. pres. indef. \\
\textbf{szétosztja} vb. pref. \emph{szét-} `apart, around'
+ \emph{oszt} `dispense, divide, deal' + 3rd
pers. sg. ind. pres. def. \emph{-ja} \\
$\rightarrow$ `to distrivute, give out' \\
\textbf{szigorú} adj. `severe, strict' \\
\textbf{szikrázik} vb. `to sparkle, to coruscate' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{szikrázón} vb. \emph{szikrázik} `to sparkle, to coruscate' +
pres. part. \\
\textbf{szilánkhoz} \emph{szilánk} `chip, shard, splinter' +
allat. \emph{-hoz} `towards' \\
\textbf{szilánkja} \emph{szilánk} `chip, shard, splinter' + 3rd
    pers. sg. single\hyp{}poss. \emph{-ja} \\
\textbf{szintén} \emph{szinte} `almost' +
adv.\hyp{}forming \emph{-n} $\rightarrow$ `likewise, also, too, as
well' \\
\textbf{szisszenj} vb. \emph{szissen} `to hiss' + 2nd
pers. sg. imper. \\
\textbf{szívek} `hear' + pl. \emph{-ek} \\
\textbf{szívben} \emph{szív} `hear' + inness. \emph{-ben} `in' \\
\textbf{szívem} \emph{szív} `heart' + 1st pers. sg. single
poss. \emph{-em} `my' \\
\textbf{szivemhez} \emph{szívemhez}: \emph{szív} `heart' + 1st
pers. sg. single\hyp{}poss. \emph{-em} `my' + allat. \emph{-hez}
`towards' \\
\textbf{szivükben} \emph{szívükben}: \emph{szív} `heart' + 3rd
pers. pl. single\hyp{}poss. poss. \emph{-ük} `their' + iness. \emph{-ben} `in' \\
\textbf{szíved} \emph{szív} `heart' + 2nd pers. sg. single
poss. \emph{-ed} \\
\textbf{szívesen} \emph{szív} `heart' + adj.\hyp{}forming \emph{-es} +
adv.\hyp{}forming \emph{-en} $\rightarrow$ `heartily, gladly' \\
\textbf{szívében} \emph{szív} `heart' + 3rd
pers. sg. single\hyp{}poss. \emph{-e} `his' +
iness. `in' \emph{-ben} \\
\textbf{szívére} \emph{szív} `heart' + 3rd
pers. sg. single\hyp{}poss. \emph{-e} `his' + subl. \emph{-re} `onto' \\
\textbf{szívvel} \emph{szív} `heart' + instr. \emph{-vel}
  `with' \\
\textbf{szomszédunkba} \emph{szomszéd} `neighbour' + 1st
pers. pl. single\hyp{}poss. \emph{-unk} + illative \emph{-ba} `into' \\
\textbf{szorítja} vb. \emph{szorít} `to press, constrict',
inf. \emph{szorítani} + 3rd pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{szobában} \emph{szoba} `room' + iness. \emph{-ban} `in' \\
\textbf{szótlan} \emph{szó} `word, voice' + privative
adj.\hyp{}forming \emph{-tlan} `-less' \\
\textbf{szösz} `harl, fluff' \\
\textbf{szövedéke} \emph{szövedék} `fabric, cloth, web' + 3rd
pers. sg. single\hyp{}poss. \emph{-e} \\
\textbf{szövőszéke} \emph{szövő} `weaver' + \emph{szék} `chair':
`loom' + 3rd pers. sg. single\hyp{}poss. \emph{-e} \\
\textbf{szőke} adj. \emph{sző} `blonde' + dim. \emph{-ke} \\
\textbf{szőtt} vb. \emph{sző} `to weave', past part. \\
\textbf{szunyókálva} (vb. pref. \emph{el-} `away' +)
vb. \emph{szunyókál} `to take a nap, to doze off' +
adv.\hyp{}part. \emph{-va} \\
\textbf{szuronyát} \emph{szurony} `bayonet' + acc. \emph{-át} \\
\textbf{születtem} vb. inf. \emph{születik} `to be born' + 1st
pers. sg. ind. past. indef. \emph{-tem} \\
\textbf{szürke} adj. `grey' \\
%%
\ \\
\noindent \textbf{\large T} \\
\textbf{tagjaim} \emph{tag} `member, limb' + 1st
pers. sg. poss. \emph{-ja} + poss. pl. \emph{-i} + 1st
  pers. sg. pers. \emph{-m} \\
\textbf{tagjaimban} \emph{tag} `member, limb' + 1st
pers. sg. poss. \emph{-ja} + poss. pl. \emph{-i} + 1st
  pers. sg. pers. \emph{-m} + iness. \emph{-ban} `in'
  $\rightarrow$ \emph{tagjaimban} \\
\textbf{takarj} vb. \emph{takar} `to cover (protect)' + snd pers. sg. imper. \\
\textbf{takarót} vb. \emph{takar} `to cover (protect)' +
pres. part. \emph{-ó} $\rightarrow$ `cover, blanket' +
acc. \emph{-t} \\
\textbf{talán} adv. `perhaps, maybe' \\
\textbf{talált} vb. \emph{talált} `to find' + past part. \\
\textbf{tangót} \emph{tangó} `tango' + acc. \emph{-t} \\
\textbf{tanul} vb. `to study, to learn' + 3rd pers. sg. ind. pres. indef. \\
\textbf{tart} vb. `to hold (in place), to support (an
obj.), to preserve, to maintain, to nurse, to keep (a rule), ' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{tartalak} vb. \emph{tart} `to hold (in place), to support (an
obj.), to preserve, to maintain, to nurse, to keep (a rule), ' + 1st
pers. sg. ind. pres. 2nd pers.\hyp{}obj. \emph{-alak} \\
\textbf{tartasz} vb. \emph{tart} `to hold (in place), to support (an
obj.), to preserve, to maintain, to nurse, to keep (a rule), ' + 2nd
pers. sg. inform. ind. pres. indef. \\
\textbf{tavasz} `spring' \\
\textbf{tájra} \emph{táj} `landscape, region' + subl. \emph{-ra}
`on[to]' \\
\textbf{tán} adv. `maybe, perhaps' \\
\textbf{tánca} \emph{tánc} `dance' + 3rd
pers. sg./pl. poss. single\hyp{}poss. \emph{-a} \\
\textbf{tárgyat} \emph{tárgy} `object, thing, subject' +
acc. \emph{-at} \\
\textbf{távozzon} vb. \emph{távozik} `to leave (formal), to go away' + 3rd
pers. (or 2nd pers. formal) sg. subj. indef. \\
\textbf{te} inform. sg. `you' \\
\textbf{tedd} vb. \emph{tesz} `to place, to put, to do, to make' + 2nd
pers. sg. inform. subj. def. \\
\textbf{tehettem} vb. \emph{tesz} `to do' + poten. \emph{-het} + 1st
pers. sg. ind. past \\
\textbf{teher} `burden, freight, load'\\
\textbf{teherpályaudvaron} \emph{teher} `freight' + \emph{pálya}
  `track' + \emph{udvar} `courtyard' + super. \emph{-on} `on' \\
\textbf{tehervonatok} \emph{teher} `freight' + vb. \emph{von} `to
pull, draw' + noun\hyp{}forming \emph{-at} $\rightarrow$ `freight train' +
pl. \emph{-ok} \\
\textbf{települ} vb. `to settle (somewhere)' + 3rd
pers. sg. pres. ind. indef. \\
\textbf{teljen} vb. inf. \emph{tellik} `to fill, get filled' + 3rd
pers. subj. indef. \\
\textbf{tenéked} pron. pers. \emph{te} `you (sg.)' + lit. \emph{néked}
`to/for you' (syn. \emph{neked}). Syn. \emph{neked} with emphasis. \\
\textbf{tenger} `sea' \\
\textbf{tengerfenékről} \emph{tenger} `sea' + \emph{fenék} `bottom' +
delat. \emph{-ől} `from, about, off' \\
\textbf{tenyerekkel} \emph{tenyér} `hand' + pl. \emph{-ek} +
instr. \emph{-kel} `with' \\
\textbf{tenyérrel} \emph{tenyér} `hand' + instr. \emph{-rel} `with' \\
\textbf{terád} pron. sg. \emph{te} `you' (here emph.) +
pron. \emph{rá} + 2nd pers. sg. `on, of you' \\
\textbf{terheim} \emph{teher} `freight, burden' + 1st
pers. sg. poss. mult. poss. \emph{-eim}
$\rightarrow$ \emph{terheim} \\
\textbf{természet} vb. \emph{terem} `to produce, yield' +
noun\hyp{}forming \emph{-észet} $\rightarrow$ `nature' \\
\textbf{termő} adj. `productive (plants, fruits), yielding, fertile,
fruitful' \\
\textbf{testvér} `sibling, sister, brother' \\
\textbf{tetten} \emph{tett} `action, deed, act' +
super. \emph{-en} `on' \\
\textbf{tettél} vb. \emph{tesz} `to place, to put, to do, to make' + 2nd
pers. sg. inform. ind. past indef. \\
\textbf{téged} pron. \emph{te} `you (sg.)' + acc. \\
\textbf{tégedet} folksy syn. \emph{téged} \\
\textbf{tépem} vb. \emph{tép} `to tear, rend' + 1st
pers. sg. ind. pres. def. \emph{-em} \\
\textbf{télen} \emph{tél} `winter' + + super. \emph{-en} `on' \\
\textbf{térdig} \emph{tér} `knee' + term. \emph{-ig} `up to'
$\rightarrow$ `knee\hyp{}deep' \\
\textbf{tétovázva} vb. \emph{tétova} `to hesitate'
(inf. \emph{tétovazik}) $\rightarrow$ adv. part. `doing smth
hesitantly' \\
\textbf{ti} pers. pron. 2nd pers. pl. `you (pl.)' \\
\textbf{tipegsz} vb. \emph{tipeg} `to toddle' + 2nd
pers. sg. inform. ind. pres. \\
\textbf{tiszta} adj. `bright' \\
\textbf{tisztábban} adj. \emph{tiszta} `clean, clear' +
comp. \emph{-bb} + deadj. adv.\hyp{}forming \emph{-an} $\rightarrow$
`more clearly, cleanly' \\
\textbf{titeket} pers. pron. \emph{ti} 2nd pers. pl. `you (pl.)' +
acc. \emph{-et} \\
\textbf{toboz} `pinecone' \\
\textbf{tolatnak} vb. \emph{tol} `to reverse, back up' +
caus. \emph{-at} `to make smth, smb do smth' + 3rd
pers. pl. ind. pres. indef. \\
\textbf{toll} `feather, pen, writer' \\
\textbf{topogásod} vb. \emph{topog} `to stomp, pace, stamp, fidget' +
agent noun\hyp{}forming \emph{-ás} $\rightarrow$ `stomping, pacing' + 2nd
pers. informal sg. poss. single\hyp{}poss. \emph{-od} \\
\textbf{totyogott} vb. \emph{totyog} `to toddle' (syn. \emph{tipeg}) +
2nd pers. form. ind. pres. indef. \\
\textbf{tovább} arch. \emph{tova} `away, far, yonder' +
comp. \emph{-bb} $\rightarrow$ `further farther' \\
\textbf{tóban} \emph{tó} `lake' + iness. \emph{-ban} `in' \\
\textbf{tócsa} `puddle' \\
\textbf{tócsába} \emph{tócsa} `puddle' + illative \emph{-ba} `into' \\
\textbf{többe} \emph{több} `more, several' + illative \emph{-be}
`into (the inside of)' \\
\textbf{többé} `anymore, again' (negative construction) \\
\textbf{törökök} \emph{török} `Turk'+ pl. \emph{-ök} \\
\textbf{töröm} vb. \emph{tör} `to break apart, to chop' + 1st
pers. sg. ind. pres. def. \\
\textbf{törvényt} \emph{törveny} `law, principle' + acc. \emph{-t} \\
\textbf{tövéhez} \emph{tö} `base of the stem of a plant, stump' + 3rd
pers. sg. single\hyp{}poss. \emph{-e} `of' $\rightarrow$ \emph{töve} +
allat. \emph{-hez} `towards' $\rightarrow$ \emph{tövéhez} \\
\textbf{tövit} \emph{tö} `base of the stem of a plant, stump' + dial. \emph{-vit} \\
\textbf{tőlem} pron. \emph{tőle} `from him, of him' + 1st pers. sg. $\rightarrow$
  `from me' \\
\textbf{tőletek} pron. \emph{tőle} `from him, of him' + 2nd
pers. pl. $\rightarrow$ `from you (pl.)' \\
\textbf{tőkét} \emph{tő} \emph{tö} `base of the stem of a plant, stump'
+ dim. \emph{-ke} + acc. \emph{-t} $\rightarrow$\
`stump' \textbar\textbar{} `capital (economy), funds, stock` \\
\textbf{tövis} `thorn' \\
\textbf{tréfát} \emph{tréfá} `joke' + acc. \emph{-at} \\
\textbf{tudja} vb. \emph{tud} `to know, to be able' + 3rd
pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{tudjak} vb. \emph{tud} `to know, to be able' + 1st
pers. sg. subj. indef. \emph{-jak} \\
\textbf{tudok} vb. \emph{tud} `to know, to be able' \textbar\textbar{} aux. `can +
inf.' + 1st pers. sg. ind. pres. indef. \emph{-ok} \\
\textbf{tudtam} vb. \emph{tud} `to know, to be able' + 1st
pers. sg. ind. past \\
\textbf{túl} adv. `beyond' \\
\textbf{túlsokat} adv. \emph{túl} `beyond' + adj.\hyp{}forming \emph{-só}
$\rightarrow$ `opposite, far, other side' + pl. \emph{-k} + acc. \emph{-at} \\
\textbf{tünődik} \emph{tűnődik} vb. `to ponder, wonder' \\
\textbf{tűntetek} vb. inf. \emph{tűnik} `to disappear' + 2nd
pers. pl. inform. ind. past indef. \\
\textbf{tüzek} \emph{tűz} `fire, passion' + pl. \emph{-ek}
$\rightarrow$ \emph{tüzek} \\
%%
\ \\
\noindent \textbf{\large U} \\
\textbf{udvar} `yard' \\
\textbf{ugrál} vb. \emph{ugrik} `to jump' + freq. \emph{-ál}
$\rightarrow$ `to bounce, leap (repeatedly)' \\
\textbf{urak} \emph{úr} `master, lord' + pl. \emph{-ak} \\
\textbf{ujságot} \emph{új} `new' + noun\hyp{}forming \emph{-ság}
(state of being)
$rightarrow$ `newspaper , news, novelty (rare)' + acc. \emph{-ot} \\
\textbf{űr} `space (astronomy)', `emptiness' \\
\textbf{űrok} `space (astronomy)', `emptiness' + pl. \emph{-ok} \\
%%
\ \\
\noindent \textbf{\large Ú} \\
\textbf{úgy} adv. `so, thusly (in this/that manner)'; \emph{úgy
... mint ...} `like'; \emph{úgy ... ha ...} `then ... if ...' \\
\textbf{úsztak} vb. inf. \emph{úszik} `to swim, to float' + 3rd
pers. pl. ind. past indef. \\
\textbf{útjuk} \emph{út} `road' + 3rd
pers. pl. single\hyp{}poss. poss. \emph{-juk} \\
\textbf{úr} `Lord' (the Christian god), `master' \\
\textbf{úri} \emph{úr} `master, Lord' + poss. pl. \emph{-i} \\
%%
\ \\
\noindent \textbf{\large Ü} \\
\textbf{ügyel} vb. `to watch over smb or smth' \\
\textbf{ügyeskedhet} vb. \emph{ügyesked} `to be clever' +
  pot. 3rd pers. sg. ind. pres. \emph{-het} \\
\textbf{ülhetek} vb. \emph{ül}
`to sit' + pot. 1st pers. sg. ind. pres. indef. \emph{-hetek} \\
\textbf{üres} `empty' \\
\textbf{üresség} \emph{üres} `empty' +
abst. noun\hyp{}forming \emph{-ség} $\rightarrow$ `emptiness' \\
\textbf{üszke} \emph{üszok} `ember, cinder' + 3rd
pers. sg. poss. single\hyp{}poss. \\
\textbf{ütött} vb. \emph{üt} `to strike' + 3rd pers. sg. ind. past
indef. \emph{-ött} \\
\textbf{ütnek} vb. \emph{üt} `to hit' + 3rd
pers. pl. ind. pres. indef. \emph{-nek} \\
%%
\ \\
\noindent \textbf{\large V} \\
\textbf{vagónokon} \emph{vagón} `wagon' + pl. \emph{-ok} +
super. \emph{-on} `on' \\
\textbf{vagy} vb. \emph{van} `to be' + 2nd
pers. sg. ind. pres. indef. \\
\textbf{vagyok} vb. \emph{van} `to be' + 1st
pers. sg. ind. pres. indef. \\
\textbf{valahol} pron. pref. \emph{vala} `some-' + interr. adv. \emph{hol} `where?' $\rightarrow$ `somewhere' \\
\textbf{valaki} pron. pref. \emph{vala} `some-' +
pron. rel. arch. \emph{ki} `who (the person(s) that)' $\rightarrow$
`somebody, someone' \\
\textbf{valamelyik} pron. pref. \emph{vala} `some-' +
    interj. \emph{myelik} `which, what' $\rightarrow$ `one, any of' \\
\textbf{valamiért} pron. pref. \emph{vala} `some-' + \emph{mi} `what' $\rightarrow$ `something' +
causal\hyp{}final \emph{ért} `for' $\rightarrow$ `for some reason' \\
\textbf{vallásban} \emph{vallás} `religion' + iness. \emph{-ban}
`in' \\
\textbf{vallhatom} vb. \emph{vall} `to confess, admit, declare, show,
speak of' + 1st pers. sg. pot. ind. pres. def. \\
\textbf{vannak} vb. \emph{van} `to be'+ 3rd
pers. pl. ind. pres. indef. \\
\textbf{vas} `iron' \\
\textbf{vasat} \emph{vas} `iron' + acc. \emph{-at} \\
\textbf{vaságyamra} \emph{vas} `iron' + \emph{ágy} `bed' + 1st
pers. sg. single\hyp{}poss. \emph{-am} + subl. \emph{-ra} `on[to]' \\
\textbf{vasútnál} \emph{vas} `iron' + \emph{ut} `way, road'+
adess. \emph{-nál} `at, by' \\
\textbf{vasszilánk} \emph{vas} `iron' + \emph{szilánk} `shard' \\
\textbf{vág} vb. `to cut' + 3rd pers. sg. ind. pres. indef. \\
\textbf{vágom} vb. \emph{vág} `to cut' + 1st pers. sg. ind. pres. def. \\
\textbf{vágy} `desire, longing' \\
\textbf{válaszolhat} \emph{válasz} `reply, answer' +
vb.-forming \emph{-ol} + poten. \emph{-hat} + 3rd
pers. sg. ind. pres. indef. \\
\textbf{vállamon} \emph{váll} `shoulder' + 1st
pers. sg. poss. single\hyp{}poss. \emph{-am} `my' + super. \emph{-on} `on' \\
\textbf{vált} vb. inf. \emph{válik} `to separate, grow apart' +
caus. \emph{-t} \\
\textbf{vár} vb. `to wait' + 3rd pers. sg. ind. pres. indef. \\
\textbf{várod} vb. \emph{vár} `to wait' + 2nd
pers. sg. ind. pres. def. \\
\textbf{vásik} rare vb. `to wear out' \\
\textbf{veled} pron. \emph{vele} `with him' + 2nd
pers. sg. $\rightarrow$ `with you' \\
\textbf{velem} pron. \emph{vele} `with him' + 1st
pers. sg. $\rightarrow$ `with me' \\
\textbf{veretni} inf. vb. `to hammer', alt. \emph{veretés} \\
\textbf{versemet} \emph{vers} `verse, poem' + 1st pers. sg. single\hyp{}poss. \emph{-em} + acc. \emph{-et} \\
\textbf{verset} \emph{vers} `verse, poem' + acc. \emph{-et} \\
\textbf{veszek} vb. \emph{vesz} `to take, grab, buy, get, consider, put on'
  + 1st pers. sg. ind. pres. indef. \emph{-ek} \\
\textbf{veszett} perfect. vb. pref. \emph{el-} + vb. \emph{vész}
or \emph{veszik} `to perish' + 3rd pers. sg. ind. past indef. $\rightarrow$ \emph{elveszett} \\
\textbf{veszteg} adv. `still, immobile' \\
\textbf{vetettek} vb. \emph{vet} `to sow', inf. \emph{vetni} + 3rd
pers. pl. ind. past. indef. \emph{-tek} \\
\textbf{védjetek} vb. \emph{véd} `to defend, protect, guard' + 2nd pers. pl. informal imper. \emph{-jetek} \\
\textbf{végén} \emph{vég} `end, finish' + 3rd
pers. sg. single\hyp{}poss. poss. \emph{-e} + iness. \emph{-n} `in' \\
\textbf{végzem} \emph{vég} `end, finish' + vb.\hyp{}forming \emph{-ez}
$\rightarrow$ vb. \emph{végez} `to finish, end, perform, do' + 1st
pers. ind. pres. def \emph{-em} \\
\textbf{véle} poetic, arch. form of \emph{vele} `with him'
\textbf{véletlen} `chance' \\
\textbf{vélt} arch. vb. \emph{vél} `to think, believe' + past
part. \emph{-t} \\
\textbf{vén} adj. `old' \\
\textbf{vidám} adj. `cheerful' \\
\textbf{vigalommal} lit. adj. \emph{víg} `joyful, happy' +
abstr. noun\hyp{}forming \emph{-alom} + 1st
pers. sg. poss. single\hyp{}poss. \emph{-om} + instr. \emph{-mal} `with'  \\
\textbf{vigyázva} vb. \emph{vigyáz} `to protect, pay attention' +
  adv. part. \emph{-va} result of action or simultaneous actions \\
\textbf{világ} `light (arch.), world' \\
\textbf{világít} \emph{világ} `light (arch.), world' +
caus. vb.-forming \emph{-ít} `to make smth ...-like' + 1st
pers. sg. ind. pres. indef. `to make shine' \\
\textbf{világokat} \emph{világ} `world' + pl. \emph{-ok} +
acc. \emph{-at} \\
\textbf{villog} vb. `to flash, to blink' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{vissza} adv. `back[wards]' \\
\textbf{villog} vb. `to flash, to blink'  + part. pres. \emph{-ó} \\
\textbf{visszaadja} vb. \emph{visszaad} `to give back' + 3rd
  pers. sg. ind. pres. def. \emph{-ja} \\
\textbf{visszajössz} vb. pref. \emph{vissza-} `back-' + vb. \emph{jön}
  `to come'+ 2nd pers. sg. ind. pres. indef. \emph{-sz} \\
\textbf{visszfénye} \emph{vissz-} `back' from adv. \emph{vissza}
+ \emph{fény} `light' 3rd
pers. sg. single\hyp{}poss. \emph{-e} \\
\textbf{viszlek} vb. \emph{visz} `to carry, to take' + 1st
pers. sg. ind. pres. 2nd obj. \emph{-lek} \\
\textbf{virág} `flower' \\
\textbf{virágokat} \emph{virág} `flower' + \emph{-ok} +
acc. \emph{-at} \\
\textbf{virul} vb. `to flower, to flourish' + 3rd
pers. sg. ind. pres. indef. \\
\textbf{vizek} \emph{víz} `water' + pl. \emph{-ek} \\
\textbf{vizen} \emph{víz} `water' + super. \emph{-en} `on'
$\rightarrow$ \emph{vízen} \\
\textbf{volna} vb. \emph{van} `to be' + 3rd
pers. sg. cond. pres. `would be, would have been' \\
\textbf{volt} vb. \emph{van} `to be'+ 3rd pers. sg. ind. past
indef. \\
\textbf{voltam} vb. \emph{van} `to be'+ 1st
pers. sg. ind. past. indef. \\
\textbf{vonat} vb. \emph{von} `to pull, draw' +
noun\hyp{}forming \emph{-at} $\rightarrow$ `train' \\
\textbf{vonszolnak} vb. \emph{vonszol} `to drag' + 3rd
pers. pl. ind. pres. indef. \\
\textbf{vontató} vb. \emph{von} `to pull, draw' + vb. modifier `let,
make smb do <action>' \emph{-tat} $\rightarrow$ `to tug, to tow' +
pres. part. \emph{-ó} \\
%%
\ \\
\noindent \textbf{\large Z} \\
\textbf{zene} `music' \\
\textbf{zokogok} vb. \emph{zokog} `to sob, speak while sobbing' + 1st
pers. sg. ind. pres. indef. \\
\textbf{zokogva} vb. \emph{zokog} `to sob, speak while sobbing' +
adv.\hyp{}part. \emph{-va} $\rightarrow$ `sobbing' \\
\textbf{zöld} `green' \\
\textbf{zöngicsél} vb. `to croon, to sing softly' \\
\textbf{zörgetek} vb. \emph{zörget} `to knock (on a door)' + 1st
pers. sg. ind. pres. indef. \\
\textbf{zörgő} vb. \emph{zörög} `to rattle' + pres. part. \emph{-ő} \\
\textbf{zúzmara} `hoarforst' \\
\textbf{zúdítja} vb. \emph{zúdít} `to pour (liquid, or abstract) on
smb/smth, to throw, to dump, to turn against' + 3rd
pers. sg. ind. pres. def. \\
\textbf{zúg} vb. `to rumble, to hum' + 3rd
pers. sg. ind. pres. indef. \\
%%
\ \\
\noindent \textbf{\large Zs} \\
\textbf{zsidó} noun. `Jew' \textbar\textbar{} adj. `Jewish' \\
\textbf{zsenge} `first born, young, immature' \\

\end{document}
