\begin{verse}[\versewidth]
  Olyan\myfoot{arch. pron. \emph{oly} `such/so' + deadj. adv.-forming
    \emph{-an} $\rightarrow$ `such a/an'}
  bolond\myfoot{adj. `silly/foolish'}
  vagy\myfoot{vb. \emph{van} `to be' + 2nd pers. sg. ind. pres. indef.} \\
  szaladsz\myfoot{vb. \emph{szalad} `to run' + 2nd
    pers. sg. ind. pres. indef. \emph{-sz}} \\
  akár\myfoot{lit. `just like'} a reggeli\myfoot{\emph{reggel}
    `morning' + adj.-forming \emph{-i}} szél.\myfoot{`wind'} \\
  Még\myfoot{adv. `moreover/later/sometime'}
  elüt\myfoot{vb. pref. \emph{el-} `away' + vb. \emph{üt} `to hit' +
    3rd pers. sg. ind. pres. indef.}
  valamelyik\myfoot{pron. pref. \emph{vala} `some-' +
    interj. \emph{myelik} `which/what' $\rightarrow$ `one/any of'}
  autó.\myfoot{`car'} \\
  Pedig\myfoot{conj. expressing contrast `whereas/but/although' or
    addition `and (by contrast)'}
  lesikáltam\myfoot{vb. pref. \emph{le-} completion/downward +
    vb. \emph{sikál} `to scrub/scour (out)' + 1st
    pers. sg. ind. past.def. \emph{-am}}
  kis\myfoot{adj. `small'} asztalomat\myfoot{\emph{asztal} `table' +
    1st pers. sg. single poss. \emph{-(o)m} + acc. \emph{-(a)t}} \\
  és most\myfoot{adv. `now'} \\
  tisztábban\myfoot{adj. \emph{tiszta} `clean/clear' + comp. \emph{-bb} +
    deadj. adv.-forming \emph{-an} $\rightarrow$ `more
    clearly/cleanly'} világít\myfoot{\emph{világ}
    `light (arch.)/world' +
    caus. vb.-forming \emph{-ít} `to make smth ...-like' + 1st
    pers. sg. ind. pres. indef. `to make shine'}
  kenyerem\myfoot{\emph{kenyér} `bread' + 1st pers. sg. single
    poss. \emph{-em}}
  enyhe\myfoot{\emph{enyh} `light/mild/tender/faint' + 3rd
    pers. sg. single poss. \emph{-e} `of'} fénye.\myfoot{\emph{fény}
    `light' + 3rd pers. sg. single poss. \emph{-e}} \\
  No\myfoot{interj. `well/now'} gyere\myfoot{vb. \emph{jön} `to come' + 2nd
    pers. sg. ind. pres. indef.} vissza,\myfoot{adv. `back[wards]'}
  ha\myfoot{conj. `if', `when/once'}
  akarod\myfoot{vb. \emph{akar} `to want' + 2nd
    pers. sg. ind. pres. def. \emph{-od}} \\
  veszek\myfoot{vb. \emph{vesz} `to take/grab/buy/get/consider/put on'
  + 1st pers. sg. ind. pres. indef. \emph{-ek}}
  takarót\myfoot{vb. \emph{takar} `to cover (protect)' +
    pres. part. \emph{-ó} $\rightarrow$
    `cover/blanket' + acc. \emph{-t}} vaságyamra.\myfoot{\emph{vas}
    `iron' + \emph{ágy} `bed' + 1st pers. sg. single poss. \emph{-am} +
  subl. \emph{-ra} `onto/on'} \\
  Egyszerű,\myfoot{\emph{egy} `one' + \emph{-szerű} `-like'
    $\rightarrow$ `plain/simple/unadorned/humble (< 1945)'}
  szürke\myfoot{adj. `grey'} takarót. \\
  Illik\myfoot{vb. `to suit/match/fit' + 3rd
    pers. sg. ind. pres. indef.} az \\
  szegénységemhez,\myfoot{\emph{szegény} `poor' + abst. noun-forming
  \emph{-ség} + 1st
  pers. sg. single poss. \emph{-em} + allat. \emph{-hez} `to'}
  aki\myfoot{conj. `who'} szeret\myfoot{vb. `to love'+ 3rd
    pers. sg. ind. pres. indef.} téged\myfoot{pron. \emph{te} `you'
    sg. + acc.} \\
  és az Úr\myfoot{`Lord' (the Christian god)} is
  szereti\myfoot{vb. \emph{szeret} `to love' + 3rd
    pers. sg. ind. pres. indef. \emph{-i}}
  nagyon\myfoot{adj. \emph{nagy} `great' + adv.-forming \emph{-on}
    $\rightarrow$ `very (to a high degree)'} \\
  és engem\myfoot{pron. \emph{én} `I' + acc. $\rightarrow$ `me'} is
  szeret\myfoot{vb. `to love'+ 3rd pers. sg. ind. pres. indef.} az Úr \\
  nem jön\myfoot{vb. `to come' + 3rd pers. sg. ind. pres. indef.}
  soha\myfoot{adv. `never'} nagy\myfoot{adj. `great'}
  fényességgel\myfoot{\emph{fény} `light' + adj.-forming \emph{-es}
    `with/having' $\rightarrow$ `bright' + abst. noun-forming
  \emph{-ség} $\rightarrow$ `brightness' +
  instr. \emph{-vel} `with' (\emph{-gel} after consonant
  \emph{g}) $\rightarrow$ `with brightness'} \\
  Nem akarja,\myfoot{vb. `to want' + 3rd
    pers. sg. ind. pres. indef. \emph{-ja}} hogy
  elromoljanak\myfoot{perfect. vb. pref. \emph{el-} +
    vb. \emph{romlik} `to break down/deteriorate' + 3rd
    pers. pl. subj. indef. \emph{-janak}} \\
  szemeim,\myfoot{\emph{szem} `eye' + 1st
    pers. sg. mult. poss. \emph{-eim}} akik\myfoot{pron. `who' +
    pl. \emph{-k}} \\
  nagyon\myfoot{adj. \emph{nagy} `great' + adv.-forming \emph{-on}
    $\rightarrow$ `very (to a high degree)'}
  kívánnak\myfoot{vb. \emph{kíván} `to wish' + 3rd
    pers. pl. ind. pres. indef. \emph{-nak}}
  látni\myfoot{vb. \emph{lát} `to see' + inf. \emph{-ni}}
  téged.\myfoot{pron. \emph{te} `you' sg. + acc.} \\
  És nagyon\myfoot{adj. \emph{nagy} `great' + adv.-forming \emph{-on}
    $\rightarrow$ `very (to a high degree)'} szépen\myfoot{adj. \emph{szép}
  `beautiful'} néznek\myfoot{vb. \emph{néznek} `to look at smth' + 3rd
  pers. pl. ind. pres. indef. \emph{-nek}}
  majd\myfoot{adv. `sometime/later'} terád\myfoot{pron. sg. \emph{te}
    `you' (here emph.) + pron. \emph{rá} + 2nd pers. sg. `on/of you'} \\
  ha\myfoot{conj. `if', `when/once'}
  visszajössz\myfoot{vb. pref. \emph{vissza-} `back-' + vb. \emph{jön}
  `to come'+ 2nd pers. sg. ind. pres. indef. \emph{-sz}} \\
  vigyázva\myfoot{vb. \emph{vigyáz} `to protect/pay attention' +
  adv. part. \emph{-va} result of action or simultaneous actions}
  foglak\myfoot{vb. aux. \emph{fog} `to will' + 1st
    pers. sg. ind. pres. 2nd obj. \emph{-lak} `you'}
  megcsókolni,\myfoot{perfect. vb. pref. \emph{meg-} +
    vb. \emph{csókol} `to kiss' + inf. \emph{-ni}} \\
  nem tépem\myfoot{vb. \emph{tép} `to tear/rend' + 1st
    pers. sg. ind. pres. def. \emph{-em}} le\myfoot{comp. `down'}
  rólad\myfoot{pron. \emph{róla} `about/off her' + 2nd pers. sg.}
  a kabátot\myfoot{\emph{kabát} `coat' + acc. \emph{-(o)t}} \\
  és elmondom\myfoot{vb. \emph{elmond} `to tell' + 1st
    pers. sg. ind. pres. def. \emph{-om}}
  mind\myfoot{pron. sg. `all/each of'} a sok\myfoot{adj. `many'}
  tréfát,\myfoot{\emph{tréfá} `joke' + acc. \emph{-(a)t}} \\
  mert\myfoot{conj. `because'} sokat\myfoot{\emph{sok} `many' +
    acc. \emph{-at}} kieszeltem\myfoot{vb. \emph{kieszel} `to devise'
    + 1st pers. sg. past. \emph{-tem}} azóta,\myfoot{art. \emph{az}
    `that' + \emph{óta} `since' $\rightarrow$ `since then'} \\
  hogy te\myfoot{pron. sg. `you'} is\myfoot{doublet of \emph{és}
    `and', adv. `also'} örülj,\myfoot{vb. \emph{örül} `to rejoice' +
    2nd pers. sg. subj. indef. \emph{-j}} \\
  majd\myfoot{adv. `sometime/later'}
  elpirulsz,\myfoot{perfect. vb. pref. \emph{el-} + vb. \emph{pirul}
    `to blush' + 2nd pers. sg. ind. pres. indef. \emph{-sz}} \\
  lenézel\myfoot{vb. `to look down'+ 2nd
    pers. sg. ind. pres. indef. \emph{-el}} a
  földre\myfoot{\emph{föld} `earth/ground' + subl. \emph{-re} `onto'}
  és kacagunk\myfoot{vb. \emph{kacag} `to laugh (a lot)' + 1st
    pers. pl. ind. pres. indef. \emph{-unk}} \\
  hangosan,\myfoot{adv. `loudly'} hogy
  behallatszik\myfoot{vb. pref. \emph{be-} `inwardly' +
    vb. \emph{hall} `to hear' + caus. \emph{-at}
    + vb.-forming \emph{-szik} $\rightarrow$ `to cause to sound in'}
  szomszédunkba\myfoot{\emph{szomszéd} `neighbour' + 1st
    pers. pl. single poss. \emph{-unk} + illative \emph{-ba} `into'} \\
  a szótlan,\myfoot{\emph{szó} `word/voice' + privative adj.-forming
    \emph{-tlan} `-less'} komoly\myfoot{adj. `severe/austere'}
  napszámosokhoz\myfoot{\emph{napszámos} `day labourer' +
    pl. \emph{-(o)k} + allat. `\emph{-hoz} `towards'} is
  behallik\myfoot{syn. \emph{behallatszik}} \\
  és fáradt,\myfoot{vb. \emph{fárad} `to tire' +
    past. part. \emph{-t}} összetört\myfoot{\emph{össze} `together' +
    vb. \emph{tör} `to break' + past. part. \emph{-t}} \\
  álmukban\myfoot{\emph{álom} `dream'+ 3rd pers. pl. single
    poss. \emph{-uk} + innes. \emph{-ban} `in'}
  majd\myfoot{adv. `sometime/later'}
  elmosolyodnak\myfoot{vb. inf. \emph{elmosolyodni}
  `to start to smile' + 3rd pers. ind. pres. indef.}
  ők\myfoot{pron. `they'} is.\myfoot{doublet of \emph{és} `and',
    adv. `also'} \\!
\end{verse}

%\poemtitle*{}

\settowidth{\versewidth}{s nem lesz emlék, melyben magadra hagyna.}

\begin{verse}%[\versewidth]
  Majd\myfoot{adv. `sometime/later'}
  megöregszel\myfoot{perfect. vb. pref. \emph{meg-} +
    vb. inf. \emph{öregszik} `to grow/become old' + 2nd
    pers. sg. inform. pres. indef.} és bánni\myfoot{vb. \emph{megbán}
    `to regret': perfect. vb. pref. \emph{meg-} + vb. \emph{bán} `to
    mind (negatively)/be bothered'} fogod,\myfoot{aux. vb. \emph{fog} `to
    will' + 2nd pers. sg. ind. pres. def. (e.g. \emph{Meg fog bánni.})} \\
  hogy\myfoot{adv. `how'} bántasz\myfoot{vb. \emph{bán} `to
    be bothered' + caus. \emph{-t}: `to hurt' + 2nd
    pers. sg. ind. pres. indef.} --- azt,\myfoot{\emph{az} `that' +
    acc. \emph{-t}} amire\myfoot{\emph{ami} `that/which' +
    subl. \emph{-re} `onto'} büszke\myfoot{`proud'}
  vagy\myfoot{vb. \emph{van} `to be' + 2nd
    pers. sg. ind. pres. indef.} ma.\myfoot{`today'} \\
  A lelkiismeret\myfoot{arch. dial. \emph{lelki} `his/her soul'
    (alt. \emph{lelke}) + arch. dial. \emph{ismereti} `his/her/its
    knowledge' (alt. \emph{ismerete}) $\rightarrow$ `(moral)
    conscience'} majd bekopog\myfoot{vb. pref. \emph{be-} `inwardly' +
    vb. \emph{kopog} `to knock'} \\
  s\myfoot{poetic abbrv. \emph{es} `and'
  (does not count as a syllable)} nem lesz\myfoot{vb. \emph{van} `to
    be', 3rd pers. sg. ind. fut. indef.}
  emlék,\myfoot{`memory'} melyben\myfoot{lit. \emph{mely}
    `which/that' + innes. \emph{-ben} `in'} magadra\myfoot{\emph{maga}
    `oneself/itself' + subl. \emph{-ra} `into'}
  hagyna.\myfoot{vb. \emph{hagy} `to allow/leave' + 3rd
    pers. sg. cond. pres. indef. \emph{-na}} \\
  \ \\
  Lesz\myfoot{vb. `to be', 3rd pers. sg. ind. fut. indef.} vén\myfoot{adj. `old'} ebed\myfoot{noun. \emph{eb} `dog' + 2nd
    pers. sg. single-poss. poss. $\rightarrow$ `your dog'} s az
  melléd\myfoot{post. \emph{mellé} `next to' + 2nd pers. sg. single
  poss. \emph{-d} $\rightarrow$ pron. `next yo you'}
  települ.\myfoot{vb. `to settle somewhere' + 3rd
    pers. sg. pres. ind. indef.} \\
  Nappal\myfoot{\emph{nap} `day' + instr. \emph{-val}
  `with' $\rightarrow$ \emph{nappal}} pihensz\myfoot{vb. `to rest' +
    2nd pers. sg. inform. ind. pres. indef.}
  majd,\myfoot{adv. `sometime/later'}  széken\myfoot{\emph{szék}
    `chair' +  super. \emph{-en} `on'}
  szunyókálva,\myfoot{(vb. pref. \emph{el-} `away' +)
    vb. \emph{szunyókál} `to take a nap/to doze off' +
    adv.-part. \emph{-va}} \\
  mert\myfoot{conj. `because'} éjjel\myfoot{poetic \emph{éj} `night' +
    instr. \emph{-vel} `with' $\rightarrow$ \emph{éjjel}}
  félni\myfoot{inf. vb. \emph{fél} `to fear/be afraid of'}
  fogsz\myfoot{aux. vb. `to will' + 2nd
    pers. sg. inform. ind. pres. def.}
  majd\myfoot{adv. `sometime/later'} egyedül.\myfoot{\emph{egy} `one'
    + noun-forming \emph{-ed} + adv.-forming \emph{-ül} $\rightarrow$
    `alone'} \\
  Árnyak\myfoot{\emph{árny} `shadow/ghost' + non. pl. \emph{-ak}}
  ütnek\myfoot{vb. \emph{üt} `to hit' + 3rd
  pers. pl. ind. pres. indef. \emph{-nek}} a
  rezgő\myfoot{poetic, arch. vb. \emph{rezg} `to tremble/quiver'
    (alt. \emph{rezeg}) + pres. part. \emph{-ő}}
  anyókára.\myfoot{\emph{anyó} `elderly woman' + dim. \emph{-ka}
    (term of endearment; rare. term of address to the wife by her
    husband in old age) + subl. \emph{-ra} `onto'} \\
  \ \\
  Az öreg\myfoot{`old'} kutya\myfoot{`dog'}
  néha\myfoot{adv. `sometimes'} majd\myfoot{adv. `sometime/later'}
  nyafog,\myfoot{vb. `to whine' + 3rd pers. sg. ind. pres. indef.} \\
  de\myfoot{conj. `but'} a szobában\myfoot{\emph{szoba} `room' +
    innes. \emph{-ban} `in'} csend\myfoot{`silence/hush'}
  lesz,\myfoot{vb. \emph{van} `to be', 3rd pers. sg. ind. fut. indef.}
  csupa\myfoot{`all/mere/pure'} rend\myfoot{`order'} lesz; \\
  hanem\myfoot{\emph{ha} conj. `if', `when/once' + \emph{nem} `no':
    conj. `but (following a negative sentence, making it positive)'}
  valaki\myfoot{pron. pref. \emph{vala} `some-' + pron. rel. arch. \emph{ki}
    `who (the person(s) that)' $\rightarrow$ `somebody/someone'}
  hiányozni\myfoot{\emph{hiány} `lack/absence/break' + vb.-forming
    \emph{-zik} $\rightarrow$ `(smb) to be missing/missed',
    inf. \emph{hiányozni}} fog\myfoot{aux. vb. \emph{fog} `to will'} \\
  a multból\myfoot{\textbf{\emph{múlt}}? rare, past part. `past' +
    elative \emph{-ból} `from/out of'} ahhoz\myfoot{\emph{az}
    `that' + allat. \emph{-hoz} `towards'} a
  magányos\myfoot{adv./adj. \emph{magán} `separate[ly]' + adj.-forming
    \emph{-os} $\rightarrow$ `lonely (psych.)'}
  csendhez.\myfoot{`silence' + allat. \emph{-hez} `to'} \\
  \ \\
  Majd tipegsz\myfoot{vb. \emph{tipeg} `to toddle' + 2nd
    pers. sg. inform. ind. pres} s\myfoot{poetic abbrv. \emph{es} `and'
  (does not count as a syllable)} ha\myfoot{conj. `if', `when/once'} eleget\myfoot{\emph{elég} `enough' + acc. \emph{-et}
    $\rightarrow$ adv.} totyogott\myfoot{vb. \emph{totyog} `to toddle'
    (syn. \emph{tipeg}) + 2nd pers. form. ind. pres. indef.} \\
  rossz\myfoot{adj. `bad'} lábod,\myfoot{\emph{láb} `leg
    (human/animal)/foot/pillar/support/stand' + 2nd
    pers. inform. poss. single-poss. \emph{-od}}
  leülsz.\myfoot{vb. pref. \emph{le-} completion/downward + \emph{ül}
    `to sit' $\rightarrow$ `to sit down'} Fönn\myfoot{adv. `above',
    alt. `fent/fenn'}
  aranykeretben\myfoot{\emph{arany} `gold' + \emph{keret} `frame' +
    innes. \emph{-ben} `in'} \\
  áll\myfoot{vb. `to stand'} ifju\myfoot{\textbf{\emph{ifjú}}?
    adj./noun `young'} képed.\myfoot{\emph{kép} `picture' + 2nd
    pers. sg. single-poss. poss. \emph{-ed}} Hozzá\myfoot{pron. `to
    it/him/her'} motyogod:\myfoot{vb. \emph{motyog} `to mumble/mutter'
  + 2nd pers. sg. ind. pres. def. \emph{-od}} \\
  "Nem öleltem\myfoot{vb. \emph{ölel} `to embrace' + 1st
    pers. ind. past} meg,\myfoot{perfect. vb. pref. \emph{meg-} (with
    \emph{ölel})} hiszen\myfoot{conj. `as/because (smth obvious)'} nem
  szerettem."\myfoot{vb. \emph{szeret} `to love' + 1st
    pers. sg. ind. past} \\
  \ \\
  "Mit\myfoot{interj. `what'} is\myfoot{doublet of \emph{és}
    `and', adv. `also'} tehettem\myfoot{vb. \emph{tesz} `to do' +
    poten. \emph{-het} + 1st pers. sg. ind. past}
  volna?"\myfoot{vb. \emph{van} `to be' + 3rd
    pers. sg. cond. pres. `would be'} ---
  kérdezed,\myfoot{vb. \emph{kérdez} `to ask' + 2nd
    pers. sg. ind. pres.} \\
  de\myfoot{adv. `how!'/conj. `but/surely'} fogatlan\myfoot{\emph{fog}
  `tooth' + priv. adj.-forming \emph{-less}} szád\myfoot{\emph{száj}
    `mouth' + 2nd pers. sg. single-poss. poss.}
  már\myfoot{adv. `already/yet/any more'} nem
  válaszolhat;\myfoot{\emph{válasz} `reply/answer' + vb.-forming
    \emph{-ol} + poten. \emph{-hat} + 3rd pers. sg. ind. pres. indef.} \\
  s\myfoot{poetic abbrv. \emph{es} `and'
  (does not count as a syllable)} ki\myfoot{adv. `out[wardly]'} a nap\myfoot{\emph{nap} `sun/day'}
  előtt\myfoot{\emph{elő} `the front part of smth' + locat. \emph{-tt}
  `in front of/ahead of/before/prior to'}
  lehunyod\myfoot{vb. pref. \emph{le-} completion/downward +
    vb. \emph{huny} `to close (one's eyes)' + 2nd
    pers. sg. ind. pres. def.} szemed,\myfoot{\emph{szem} `eye' + 2nd
    pers. sg. single-poss. poss. \emph{-ed} (but plural implied here!)} \\
  alig\myfoot{adv. `barely'}
  várod,\myfoot{vb. \emph{vár} `to wait' + 2nd
    pers. sg. ind. pres. def.} hogy\myfoot{conj. `that'}   feljöjjön,\myfoot{vb. pref. \emph{fel-} `upwards' (see also
    \emph{föl-}) + vb. \emph{jön} `to come' + 3rd
    pers. sg. subj. pres. indef.} a holdat.\myfoot{`moon' +
    acc. \emph{-(a)t}} \\
  \ \\
  Mert\myfoot{conj. `because'} ha\myfoot{conj. `if', `when/once'} elalszol\myfoot{vb. pref. \emph{el-} `away' +
    vb. \emph{alszik} `to sleep' $\rightarrow$ `to fall asleep' + 2nd
    pers. sg. ind. pres. indef.},
  ugrál\myfoot{vb. \emph{ugrik} `to jump' + freq. \emph{-ál}
    $\rightarrow$ `to bounce/leap (repeatedly)'} majd\myfoot{adv. `sometime/later'} az ágy,\myfoot{`bed'} \\
  mint\myfoot{conj. `like'} a csikó,\myfoot{`foal'} hogy a
  hámot\myfoot{\emph{hám} `harness' + acc. \emph{-(o)t}}
  levesse.\myfoot{vb. \emph{levet} `to buck/throw off' + 3rd
    pers. sg. subj. pres. def.} \\
  S\myfoot{poetic abbrv. \emph{es} `and'} a
  félelem\myfoot{vb. \emph{fél} `to be afraid' + noun-forming
    \emph{-elem} $\rightarrow$ fear'}
  tünődik,\myfoot{\textbf{\emph{tűnődik}}? vb. `to ponder/wonder'} nem
  a vágy,\myfoot{`desire/longing'} \\
  a fejedben:\myfoot{\emph{fej} `head' + 2nd
    pers. sg. single-poss. poss. \emph{-ed} + innes. \emph{-ben} `in'}
  Szeress-e,\myfoot{vb. \emph{szeret} `to love' + 2nd
    pers. sg. subj. + polar interr. \emph{-e} }
  ne\myfoot{\emph{nem} becomes \emph{ne} in imperative/subjunctive}
  szeress-e. \\
  \ \\
  Magadban\myfoot{\emph{maga} `oneself' + 2nd pers. sg. single
  poss. \emph{-d} + iness. \emph{-ban} `in'}
  döntöd\myfoot{perfect. vb. pref. \emph{el-} + vb. \emph{dönt} `to
    decide' + 2nd pers. sg. ind. pres. def.}
  el.\myfoot{perfect. vb. pref. \emph{el-}} Én\myfoot{pron. `I'}
  fájlalom,\myfoot{vb. \emph{fáj} `to hurt' + vb.-forming \emph{-lal}
    $\rightarrow$ literary, arch. `to regret' + 1st
    pers. sg. ind. pres. def.} \\
  hogy nem felelhetek,\myfoot{\emph{fél} `human being' + vb.-forming
    \emph{-el} $\rightarrow$ vb. \emph{felel} `to answer/reply' +
    poten. 1st pers. sg. ind. pres. indef.} ha\myfoot{conj. `if',
    `when/once'} kérded:\myfoot{arch. vb. \emph{kérd} `to
    ask/question' + 2nd pers. sg. ind. pres. def.}
  él-e.\myfoot{vb. \emph{él} `to live' + 3rd
    pers. sg. ind. pres. indef. +
polar interr. \emph{-e}} \\
  Mert\myfoot{conj. `because'}
  elfárad\myfoot{perfect. vb. pref. \emph{el-} + vb. \emph{fárad} `to
    tire/to be tired' + 3rd pers. sg. ind. pres. indef.}
  bennem\myfoot{pron. \emph{benne} `inside smth/smb' + 1st
    pers. sg. pers. \emph{-m}} a fájdalom,\myfoot{vb. \emph{fáj} `to
    ache' + noun-forming \emph{-dalom} $\rightarrow$
    `pain/grief/sorrow'} \\
  elalszik,\myfoot{perfect. vb. pref. \emph{el-} +
    vb. \emph{alszik} `to sleep' $\rightarrow$ `to fall asleep' + 3rd
    pers. sg. ind. pres. indef.} mint\myfoot{conj. `like'} a
  gyermek\myfoot{arch. `child', alt. \emph{gyerek}} s én is
  véle.\myfoot{poetic, arch. form of \emph{vele} `with it/him/her'} \\
\end{verse}

\oldstylenums{1936} November

\bigskip

\noindent \textbf{Original Hungarian}

\selectlanguage{hungarian}


\begin{verse}%[\versewidth]
  Vasútnál\myfoot{\emph{vas} `iron' + \emph{ut} `way/road'+
    adess. \emph{-nál} `at/by'} lakom.\myfoot{vb. \emph{lak} `to
    dwell', inf. \emph{lakik}, 1st pers. sg. ind. pres. \emph{-(o)m}}
  Erre\myfoot{pron. \emph{ez} `this/it' + subl. sg. \emph{-re}
    `onto/to'}
  sok\myfoot{`much/many'} \\
  vonat\myfoot{vb. \emph{von} `to pull/draw' + noun-forming
    \emph{-at} $\rightarrow$ `train'} jön\myfoot{vb. `to come' + 3rd
    pers. sg. ind. pres. indef.}-megy\myfoot{vb. `to go/travel/pass by
    (of time)' + 3rd pers. sg. ind. pres. indef.} és
  el\myfoot{adv./conj. `afar/far
    away'}-elnézem,\myfoot{vb. pref. \emph{el-} `continuity over a long
    time' + vb. \emph{nez} `to look at'} \\
  hogy'\myfoot{short. \textbf{\emph{hogyan?}} `how'}
  szállnak\myfoot{vb. \emph{száll} `to fly/land/come down' + 3rd
  pers. pl. ind. pres. indef.} fényes\myfoot{\emph{fény}
    `light/glitter' + adj.-forming \emph{-es} $\rightarrow$ `bright'}
  ablakok\myfoot{\emph{ablak} `window' + pl. \emph{-(o)k}} \\
  a lengedező\myfoot{vb. \emph{leng} `to swing/sway/wave' +
    freq. vb.-forming \emph{-edezik} (unrounded front\hyp{}vowel) +
    pres. part. \emph{-ő} $\rightarrow$ `swinging/waving'}
  szösz\myfoot{`harl/fluff'}-sötétben.\myfoot{adj./subst. \emph{sötét}
    `dark[ness]' + iness. sg. \emph{-ben} `in'} \\
  Így\myfoot{`just so'} iramlanak\myfoot{\emph{ir} `movement' +
  rare instr. vb.-forming \emph{-amlik} (inf.) $\rightarrow$ `to
  rush/hurry' + 3rd pers. pl. ind. pres. indef. \emph{-(a)nak}}
  örök\myfoot{`perpetual/eternal'} éjben\myfoot{poetic \emph{éj} `night' +
  innes. `in' \emph{-ben}} \\
  kivilágított\myfoot{vb. pref. \emph{ki} `outwardly' + \emph{világ}
    `light (arch.)/world' $\rightarrow$ arch. `illumination' +
    caus. vb.-forming \emph{-ít} `to make smth ...-like' $\rightarrow$
    `to illuminate' + past part. \emph{-ott}}
  nappalok\myfoot{\emph{nap} `day' + instr. \emph{-val}
  `with' $\rightarrow$ \emph{nappal} + pl. \emph{-(o)k}} \\
  s én\myfoot{pron. `I'} állok\myfoot{vb. \emph{áll} `to stand' + 1st
    pers. sg. ind. pres. indef. \emph{-ok}}
  minden\myfoot{det. `every'} fülke\myfoot{\emph{fül}
    `semi\hyp{}circular obj./shape' + dim. \emph{-ke} $\rightarrow$
    `compartment'}-fényben,\myfoot{\emph{fény}
    `light/glitter' + innes. \emph{-ben} `in'} \\
  én könyöklök\myfoot{\emph{könyök} `elbow' + vb.-forming \emph{-öl}
    $\rightarrow$ `to rest on one's elbow' + 1st
    pers. sg. ind. pers. \emph{-ök}} és
    hallgatok.\myfoot{vb. \emph{hallgat} `to remain silent/to listen'
      + 1st pers. sg. ind. pres. indef. \emph{-ok}} \\
\end{verse}

\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: o(8)/é-e(9)/a-o(8)/é-e(9) \\
5-8: é-e(9)/a-o(8)/é-e(9)/o(8)


\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: ki-nek(8)/ap-ja(9)/e-e(8)/a[p]-ja(9)\\
5-8: a[d]-ja(9)/i-e(8)/ap-ja(9)/ki-nek(8)

\newpage

\selectlanguage{english}

\noindent \textbf{Cursory remark}

\medskip

While this seems a rather descriptive poem, the first and last two
verses are subjective. Also the last verse echoes the first, in the
fashion of a chiasm: verse~1 is about seeing happiness, and the last
features light and playfulness: the Hungarian uses \emph{fény} for
`light', but it could figuratively also mean `joy', hence my
translation: `the sunlight toyed'.

\bigskip

\selectlanguage{english}

\noindent \textbf{Original Hungarian}

\selectlanguage{hungarian}


\begin{verse}%[\versewidth]
  Láttam\myfoot{vb. \emph{lát} `to see' +
  short base \emph{-t}, 1st pers. sg. past def. \emph{-tam}} a
  boldogságot\myfoot{adj. \emph{boldog} `happy' + noun-forming
  \emph{-ság} (state of being) + acc. \emph{-(o)t}}
  én,\myfoot{pron. `I'} \\
  lágy\myfoot{adj. `soft'} volt,\myfoot{vb. \emph{van} `to
  be'+ 3rd pers. sg. ind. past indef.} szőke\myfoot{adj. \emph{sző}
  `blonde' + dim. \emph{-ke}} és másfél\myfoot{\emph{más}
  `other/different' + \emph{fél} `half' $\rightarrow$ `one and a
  half'} mázsa.\myfoot{`quintal'} \\
  Az udvar\myfoot{`yard'} szigorú\myfoot{adj. `severe/strict'}
  gyöpén\myfoot{\emph{gyep} `lawn/grass' + super. \emph{-(e)n} `on'} \\
  imbolygott\myfoot{vb. \emph{imbolyog} `to sway'+ 3rd
  pers. sg. single poss. \emph{-e}} göndör\myfoot{adj. `curly'}
  mosolygása.\myfoot{\emph{mosolygás} `smile' + 3rd
  pers. sg. single poss. \emph{-a}} \\
  Ledőlt\myfoot{vb. \emph{ledoől} `to fall down' + 3rd
  pers. sg. ind. past indef. \emph{-t}} a puha,\myfoot{adj. `soft/squishy'}
  langy\myfoot{arch. poetic \emph{langyos} `tepid'}
  tócsába,\myfoot{\emph{tócsa} `puddle' + illative \emph{-ba}
  `into'} \\
  hunyorgott,\myfoot{vb. \emph{hunyorg} `to squint' + 3rd
  pers. sg. ind. past. indef. \emph{-rgott}}
  röffent\myfoot{\emph{röffen} `to grunt once (of a pig)'}
  még\myfoot{adv. `still/moreover'} felém\myfoot{post. \emph{felé}
  `to/towards' + 1st pers. sg. poss \emph{-m} $\rightarrow$
  pron. `towards me'} --- \\
  ma\myfoot{adv. `today'} is\myfoot{doublet of \emph{és} `and',
  adv. `also'} látom,\myfoot{vb. \emph{lát} `to see'+ 1st
  pers. sg. ind. pres. def.} mily\myfoot{poetic `how ...(!)'}
  tétovázva\myfoot{vb. \emph{tétova} `to hesitate'
  (inf. \emph{tétovazik}) $\rightarrow$ adv. part. `doing smth
  hesitantly'} \\
  babrált\myfoot{vb. \emph{babrál} `to fiddle'}
  pihéi\myfoot{\emph{pihe} `down, fluff' + adj.-forming \emph{-i}}
  közt\myfoot{post. `between/amid' from \emph{között}: \emph{köz}
  `interval' + locat. \emph{-(ö)tt}} a fény.\myfoot{`light/glitter',
  fig. `happiness/joy'} \\
\end{verse}

\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: o-é(8)/á-a(9)/ö-é(8)/á-a(9) \\
5-8: á-a(9)/e-é(8)/á-a(9)/a-é(8)

\newpage


\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: a-at(8)/et-ni(9)/a-a[dt](8)/e[d]-ni(9) \\
5-8: e[t]-ni(9)/a-a[tt](8)/et-ni(9)/u-at(8)

\newpage

\selectlanguage{english}


\noindent \textbf{Cursory remark}

\medskip

The integral man (the Hungarian language is actually gender neutral)
knows that his parents cannot protect him anymore from the truth of
mortality. This truth is not treasured by gods, as they are immortal,
nor Christian priests, who preach eternal life.

\bigskip

\selectlanguage{english}

\noindent \textbf{Original Hungarian}

\selectlanguage{hungarian}

\begin{verse}%[\versewidth]
  Az meglett\myfoot{obs. `adult'} ember\myfoot{`person'},
  akinek\myfoot{pron. \emph{aki} `who' + attr. poss. \emph{-nek}} \\
  szívében\myfoot{\emph{szív} `heart' + poss. `his' \emph{-e} +
  innes. `in' \emph{-ben}} nincs se anyja,\myfoot{\emph{anya}
  `mother' + 3rd pers. sg. single poss. \emph{-(j)a}}
  apja,\myfoot{\emph{apa} `father' + 3rd pers. sg. single
  poss. \emph{-(j)a}} \\
  ki\myfoot{pron. rel. arch. `he who', alt. \emph{aki}}
  tudja,\myfoot{vb. \emph{tud} `to know' + 3rd
  pers. sg. ind. pres. def. \emph{-ja}}
  hogy\myfoot{conj. `so that'} az életet\myfoot{\emph{élet}
  `life' + acc. \emph{-(e)t}} \\
  halálra\myfoot{vb. \emph{hal} `to die' + arch. noun-forming
  \emph{-ál} $\rightarrow$ `death' + subl. `on/for' \emph{-ra}
  $\rightarrow$ `to death'}
  ráadásul\myfoot{\emph{ráadás} `addition/extra' + essive
  `as/with the intention of' \emph{-ul}}
  kapja\myfoot{vb. \emph{kap} `to receive' + 3rd
  pers. sg. ind. pres. def. \emph{-ja}} \\
  s mint\myfoot{`like'} talált\myfoot{vb. \emph{talált} `to
  find' + past part.} tárgyat\myfoot{\emph{tárgy}
  `object/thing/subject' + acc. \emph{-(a)t}}
  visszaadja\myfoot{vb. \emph{visszaad} `to give back' + 3rd
  pers. sg. ind. pres. def. \emph{-ja}} \\
  bármikor\myfoot{\emph{bár} `any' + adv. \emph{mikor} `when'
  $\rightarrow$ `any time'} --- ezért\myfoot{conj. `therefore'}
  őrzi\myfoot{vb. \emph{őriz} `to guard' + 3rd
  pers. sg. ind. pres. def. \emph{-i}} meg,\myfoot{arch. emph.} \\
  ki\myfoot{`he who'} nem istene\myfoot{\emph{isten} `god' +
  3rd pers. sg. sing. poss. \emph{-e}} és nem
  papja\myfoot{\emph{pap} `priest' + 3rd
  pers. sg. sing. poss. \emph{-(j)a}} \\
  se\myfoot{alt. \emph{sem}, \emph{sem ..., sem ...}  `neither
  ... nor ...'} magának,\myfoot{pron. refl. 3rd pers. `himself'}
  sem senkinek.\myfoot{pron. \emph{senki} `no one/nobody'} \\
\end{verse}


\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: ü-ött(8)/á-o(9)/ö-ö(8)/á-o(9) \\
5-8: á-o(9)/ö-ö(8)/á-o(9)/ö-ött(8)

\newpage


\noindent \textbf{Cursory remark}

\medskip

Verses 1-2 refer to the sounds of iron being forged and quenched in
water. The blacksmith is fixing a crack, which reminds J of a painful
past event, possibly a break\hyp{}up. Verses 3-6 form a chiasm:
verse~5 echoes verse~4: to avoid his love fading away, he will keep
loving; verse~6 answers verse~3: instead of hammering the crack to
make it disappear, J bends under it. Then there is no need to forge a
sword blade to become strong. J's self\hyp{}consciousness is golden
like the glow of the kiln's interior.

\bigskip

\noindent \textbf{Original Hungarian}


\begin{verse}%[\versewidth]
  Hallottam\myfoot{vb. \emph{hall} `to hear' + 1st
  pers. sg. ind. past def. \emph{-ottam}} sírni\myfoot{vb.
  \emph{sír} `to cry/weep' + inf. \emph{-ni}} a
  vasat,\myfoot{\emph{vas} `iron' + acc. \emph{-(a)t}} \\
  hallottam az esőt\myfoot{vb. \emph{es}, inf. \emph{esik} +
  pres. part. \emph{-ő} + acc. \emph{-t}} nevetni.\myfoot{vb.
  \emph{nevet} `to laugh', inf. \emph{nevetni}} \\
  Láttam,\myfoot{vb. \emph{lát} `to see' +
  short base \emph{-t}, 1st pers. sg. past def. \emph{-tam}}
  hogy\myfoot{adv. `how'} a mult\myfoot{\textbf{\emph{múlt}}?
  rare, past part. `past', inf. \emph{múlik}}
  meghasadt\myfoot{perfect. vb. pref. \emph{meg-} + vb.
  \emph{hasad} `to crack/split apart' + past
  part. \emph{-t}} \\
  s csak\myfoot{adv. `only'} képzetet\myfoot{\emph{képzet}
  `idea/image/notion' + acc. \emph{-t}}
  lehet\myfoot{pot. vb. \emph{van/lesz}:
  `may be/may become', 3rd pers. sg. ind.}
  feledni;\myfoot{lit. vb. \emph{feled} `to forget', alt. \emph{felejt}} \\
  s hogy\myfoot{adv. `how'} nem tudok\myfoot{vb.
  \emph{tud} `to know'/aux. `can + inf.' + 1st
  pers. sg. ind. pres. indef. \emph{-ok}}
  mást,\myfoot{adj. \emph{más} `other/different' + acc. \emph{-t}}
  mint\myfoot{comp. different degrees `than'}
  szeretni,\myfoot{inf. vb. \emph{szeret} `to love'} \\
  görnyedve\myfoot{vb. \emph{görnyed} `to bend (under a
  burden)' + adv.-forming \emph{-ve}} terheim\myfoot{\emph{teher}
  `freight/burden' + 1st
  pers. sg. poss. mult. poss. \emph{-eim} $\rightarrow$ \emph{terheim}}
  alatt\myfoot{post. \emph{al-} `lower part'+ locat. \emph{-(a)tt}}
  --- \\
  minek\myfoot{\emph{mi} + dat. \emph{-nek}: `for what'}
  is\myfoot{doublet of \emph{és} `and', adv. `also'}
  kell\myfoot{aux. vb. `must/need to/have to + inf.'}
  fegyvert\myfoot{\emph{fegyever} `weapon' + acc. \emph{-t}}
    veretni\myfoot{inf. vb. `to hammer', alt. \emph{veretés}} \\
  belőled,\myfoot{pron. \emph{belőle} `out of him/her/it' + 2nd
  pers. sg. \emph{-d}} arany\myfoot{`gold/golden'}
  öntudat!\myfoot{`self-awareness'} \\
\end{verse}


\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: a-ól(8)/é-e(9)/a-ól(9)/é-e(9) \\
5-8: é-e(9)/a-ól(8)/é-e(9)/a-ol(8)

\newpage

\selectlanguage{english}

\bigskip

\noindent \textbf{Cursory remark}

\medskip

This is likely a memory of J's imprisonment in \oldstylenums{1931}.

\bigskip

\noindent \textbf{Original Hungarian}

\selectlanguage{hungarian}

\begin{verse}%[\versewidth]
  Fülelt\myfoot{\emph{fül} `ear' + vb.-forming \emph{-el}: `to
  listen (intently)' + 3rd pers. ind. past indef. \emph{-t}} a
  csend\myfoot{`silence'. alt. \emph{csönd}}
  --- egyet\myfoot{\emph{egy} + acc. \emph{-t}}
  ütött\myfoot{vb. \emph{üt} `to strike' + 3rd
  pers. sg. ind. past indef. \emph{-(ö)tt}} \\
  Fölkereshetnéd\myfoot{\emph{föl-}
  `upwards' + \emph{keres} `to seek': `to visit' +
  pot. `may/might' \emph{-het} + 2nd
  pers. sg. ind. pres. cond. \emph{-ned}, alt. \emph{felkeres}}
  ifjúságod;\myfoot{adj. \emph{ifjú} `young' + noun-forming
  \emph{-ság} (state of being): `youth' + 2nd pers. informal sg.
  poss. single poss. \emph{-od}: `your childhood' (informal)} \\
  nyirkos\myfoot{adj. `humid/damp'}
  cementfalak\myfoot{\emph{cement} `cement'+ \emph{fal} `wall' +
  pl. \emph{-ak}} között\myfoot{post. `between/amid', \emph{köz}
  `interval' + locat. \emph{-(ö)tt}} \\
  képzelhetsz\myfoot{\emph{kép} `image' + vb.-forming
  \emph{-(z)el}: `to imagine' + pot. \emph{-het} + 2nd
  pers. sg. ind. pres. \emph{-sz}} egy kis
  sabadságot\myfoot{\emph{szabad} `free' + \emph{-ság} (state of
  being): `freedom' + acc. \emph{-(a)t}} --- \\
  gondoltam.\myfoot{vb. \emph{gondol} `to
  think/plan/guess/worry' + 1st pers. sg. ind. past
  \emph{-tam}} S hát\myfoot{`back body','well...',
  `then/and/but' (questioning back), `surely'}
  amint\myfoot{\emph{az} + \emph{mint} `like': `as soon as',
  literary `as'} fölállok\myfoot{\emph{föl-}
  `upwards' +  vb. \emph{áll} `to stand' + 1st
  pers. sg. ind. pres. indef. \emph{-ok}, alt. \emph{felállok}} \\
  a csillagok,\myfoot{\emph{csillag} `star' + pl. \emph{-ak}} a
  Göncölök\myfoot{\emph{Göncöl} + pl. \emph{-ök}: `the Great Bear'
  (constellation)} \\
  úgy\myfoot{adv. `so (in this/that manner)'; \emph{úgy ... mint ...}
    `like'} fénylenek\myfoot{\emph{fény} `light' +
  freq. vb.-forming \emph{-lik}: `to glitter' + 3rd
  pers. pl. ind. pres. indef. \emph{-ek}}
  fönt,\myfoot{adv. `above/up/awake/upstairs', alt. \emph{fent}}
  mint a rácsok\myfoot{\emph{rács} `grid/grille' + pl. \emph{-ok}} \\
  a hallgatag\myfoot{vb. \emph{hallgat} `to remain silent/to
  listen' + adj.-forming \emph{-ag}} cella\myfoot{`cell'}
  fölött.\myfoot{\emph{föl-} `up/upwards' +  locat. \emph{-(ö)tt}:
  `above', alt. \emph{elett}} \\
\end{verse}


\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: e-ül(8)/á-at(9)/e-ül(8)/á-at(9) \\
5-8: á-ad(9)/e-ül(8)/á-at(9)/e-ül(8)

\newpage

\selectlanguage{english}

\noindent \textbf{Cursory remark}

\medskip

Whereas verses 1-2 echo the ancient mechanical understanding of the
firmament as concentric spheres, and verses 2-4 hearken back at the
mythological Fates (Parc\ae), verses 5-8 evoke the image of an
aerostat, whose aeronaut is J, where J's dreams are the steam that
tries to lift him up, and the law is the envelope, whose tears are J's
constant misfortune. (The metaphor is a little off, as balloons use
hot air, not steam.)

\bigskip

\noindent \textbf{Original Hungarian}

\selectlanguage{hungarian}


\begin{verse}%[\versewidth]
  Én fölnéztem\myfoot{vb. pref. \emph{föl-} `upwards' + vb.
  \emph{néz} `to look' + 1st pers. sg. past. def. \emph{-tem}} az
  est\myfoot{arch. `(in the) evening'} alól\myfoot{\emph{al}
  `lower part' + abl. \emph{-l} `from'} \\
  az eget\myfoot{\emph{ég} `sky' + acc. \emph{-(e)t}}
  fogaskerekére\myfoot{\emph{fogas} `tooth' +
  \emph{kerék} `wheel': `cogwheel' + 3rd pers. sg. single
  poss. \emph{-e} + subl. \emph{-re} `towards'} --- \\
  csilló\myfoot{vb. \emph{csillog} `to glitter' +
  adj.-forming \emph{-o} $\rightarrow$
  \emph{csilló}; usually \emph{csillogó}}
  véletlen\myfoot{`chance'} szálaiból\myfoot{\emph{szál}
  `thread' + 3rd pers. sg. poss. \emph{-a} + poss. pl. \emph{-i} +
  elative \emph{-ból} `from/out of'} \\
  törvényt\myfoot{\emph{törveny} `law/principle' + acc. \emph{-t}}
  szőtt\myfoot{vb. \emph{sző} `to weave', past part.} a
  mult\myfoot{\textbf{\emph{múlt}}? rare, past part. `past',
  inf. \emph{múlik}} szövőszéke\myfoot{\emph{szövő} `weaver' +
  \emph{szék} `chair': `loom' + 3rd pers. sg. single poss. \emph{-e}} \\
  és megint\myfoot{\emph{meg} `again' (short action) +
  adv.-forming \emph{-int}} fölnéztem az égre\myfoot{\emph{ég}
  `sky' + subl. sg. \emph{-re} `onto'} \\
  álmaim\myfoot{\emph{álom} `dream' +
  1st pers. sg. poss. \emph{-a} + poss. pl. \emph{-i} + 1st
  pers. sg. pers. \emph{-m}} gőzei\myfoot{\emph{gőz} `steam,
  vapour' + 3rd pers. sg. poss. \emph{-e} + pl. poss. \emph{-i}} alól \\
  s láttam,\myfoot{vb. \emph{lát} `to see' +
  short base \emph{-t}, 1st pers. sg. past def. \emph{-tam}} a törvény
  szövedéke\myfoot{\emph{szövedék} `fabric/cloth/web' + 3rd
  pers. sg. single poss. \emph{-e}} \\
  mindig\myfoot{\emph{mind} `always' (repetition) + term. `until'
  \emph{-ig}} fölfeslik\myfoot{vb. pref. \emph{föl-} `upwards'+
  arch., poetic, vb.
  \emph{feslik} `to come unstitched', 3rd pers. sg. ind. pres. indef.}
  valahol.\myfoot{`somewhere'} \\
\end{verse}


\selectlanguage{english}

\noindent Chiastic structure: \\
1-4: a-o(8)/é-e(9)/e-o(8)/é-e(9) \\
5-8: é-e(9)/o-o(8)/é-e(9)/o-o(8)


\bigskip

\noindent \textbf{Cursory remark}

\medskip

We are not captives of the world, but of our own heart, which invited
the world (the landlord) in for us to contend with. The inner struggle
is felt as anguish and yields a feverish soul.

\bigskip

\noindent \textbf{Original Hungarian}


\begin{verse}%[\versewidth]
  Im\myfoot{literary, abbrv. \emph{íme} `here is/behold'}
  itt a szenvedés\myfoot{vb. \emph{szenved} `to suffer'
  + noun-forming \emph{-és}} belül,\myfoot{`inside'} \\
  ám\myfoot{poetic, interj. contrad. `but'} ott
  kívül\myfoot{adv. `outdoor/outside'} a
  magyarázat.\myfoot{vb. \emph{magyaráz} `to explain' +
  noun-forming \emph{-at}} \\
  Sebed\myfoot{\emph{seb} `wound' + 2nd
  pers. sg. single poss. \emph{-ed}} a világ --- ég,\myfoot{vb.
  `to be on fire', 3rd pers. sg. ind. pres. indef.}
  hevül\myfoot{vb. `to grow hot', 3rd
  pers. sg. ind. pres. indef.} \\
  s te\myfoot{inform. sg. `you'} lelkedet\myfoot{\emph{lélek} `soul' + 2nd
  pers. sg. single poss. \emph{-ed} $\rightarrow$ \emph{lelked} +
  acc. \emph{-(e)t}} érzed,\myfoot{vb. \emph{érez} `to feel', 2nd
  pers. sg. ind. pres. def.} a lázat.\myfoot{\emph{láz} `fever' +
  acc. \emph{-(a)t}} \\
  Rab\myfoot{`captive'} vagy, amíg\myfoot{adv. `as long as';
  `until' if followed by negated verb} a
  szíved\myfoot{\emph{szíved} `heart' + 2nd pers. sg. single
  poss. \emph{-ed}} lázad\myfoot{vb. \emph{to rebel}, 3rd
  pers. sg. ind. pres. indef.} --- \\
  úgy\myfoot{\emph{úgy ... ha ...} `then ... if ...'}
  szabadulsz,\myfoot{\emph{szabad} `free' + vb.-forming \emph{-ul}
  + 2nd pers. sg. ind. pres. indef. \emph{-sz}} ha\myfoot{`if'}
  kényedül\myfoot{literary, vb. `to indulge', 3rd
  pers. sg. ind. pres. indef.} \\
  nem raksz\myfoot{vb. \emph{rak} `to set up/to build', 2rd
  pers. sg. ind. pres. indef. \emph{-sz}}
  magadnak\myfoot{\emph{maga} `oneself' + 2nd pers. sg. single
  poss. \emph{-d} + dat. \emph{-nak}}
  olyan\myfoot{pron. `such/that kind of thing'}
  házat,\myfoot{\emph{ház} `house' + acc. \emph{-(a)t}} \\
  melybe\myfoot{literary, arch. \emph{mely}
   `which' + illative \emph{-be} `into (the inside of)'}
  háziúr\myfoot{\emph{ház} `house' + adj.-forming `of the'
  \emph{-i} + \emph{úr} `master'} települ.\myfoot{vb. `to
  settle (somewhere)', 3rd pers. sg. ind. pres. indef.} \\
\end{verse}
