\documentclass[twocolumn,11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[charter]{mathdesign}
\usepackage{verse}

\setlength\columnsep{0em}

\title{Perigee of Mercury}
\author{}
\date{}

\begin{document}

\maketitle

\thispagestyle{empty}

%\poemtitle{}

%\settowidth{\versewidth}{reviving the fire under the crucible}

\bigskip

\begin{verse}%[\versewidth]
  My eyes wander in the firmament, \\
  that same sky that separates lovers, \\
  and I ponder with gravity a detour, \\
  like a winged thief without return.

  I lower my gaze towards the earth. \\
  Will you forgive me \\
  after this season in hell?

  With the tip of my tongue \\
  I used to carve furrows in your hair, \\
  like a ploughman sowing wheat \\
  in the shiver of dawn.

  When the quicksilver descended \\
  you swallowed tears of rage \\
  for not being free to follow me. \\
  You wanted to write \\
  but your ink spilt on the page \\
  and became evening on the frosted fields.

  But the time has come and I am back.

  The alchemy between us is not dead, \\
  only asleep and frozen. \\
  I will revive the fire under the crucible \\
  until the terracota cracks.

  \newpage

  Already \\
  mounts arise, springs gush, \\
  continents smash into each other!

  Already \\
  you sigh like bellows \\
  on the furnace renewed, \\
  on the origin of the world.

  Finally \\
  our lips join and I taste \\
  the vermilion flesh\\
  of a pomegranate \\
  under the moonlight.

  You lay hands on my head \\
  and crown me witness \\
  of the mystical union \\
  of water and fire.

  I lay hands on a mute book \\
  about the celestial geography \\
  of your beauty spots, \\
  like braille I decipher \\
  a secret zodiac \\
  on your body of soft amber \\
  where I die and am reborn.
\end{verse}

\end{document}
