\documentclass[twocolumn,12pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[charter]{mathdesign}
\usepackage{verse}

\begin{document}

\thispagestyle{empty}

\poemtitle{}

\settowidth{\versewidth}{in the calm stream of your words}

\bigskip

\begin{verse}[\versewidth]
  Sleds rushing down a hill, \\
  each bump threading a note \\
  to the braids of their life lines, \\
  to the necklaces of their laughter.

  Like them, \\
  I draw on immaculate pages \\
  invisible parallels \\
  that conjure up your silhouette.

  Hide and seek in the mist, \\
  your white hands over my eyes, \\
  your panting for breath \\
  sends a long shiver down, \\
  blending with that of winter.

  I caress the blank pages \\
  of a book open to the invisible, \\
  invoking your pale face \\
  under my fingers trembling \\
  from soft and familiar hills, \\
  naked under the ermine coat, \\
  my breath taken away \\
  at each descent, \\
  so impatient at each ascent!

  Exhausted, lying on the snow \\
  we watch passing clouds \\
  and I quench my thirst \\
  in the calm stream of your words \\
  lest they drop into oblivion \\
  like a frozen cataract.

  \newpage

  Beloved phantom, \\
  was it a word too pure, \\
  a heart too warm \\
  that made you evaporate?

  You only left, \\
  in a margin \\
  of the white grimoire, \\
  a golden hair.

  A prod to the tiny spring \\
  has my heart palpitating \\
  like a watch.

  Bent over those pages \\
  in the corner of a table \\
  we used to admire... what?

  The smile erases the memory, \\
  like an overexposed film, \\
  and only remain these verses \\
  in braille \\
  where I feel your steps \\
  towards the white paradise.
\end{verse}

\end{document}
